---
output: github_document
---

# DataRaft

[![R-CMD-check](https://github.com/JanWein/tidyweave/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/JanWein/tidyweave/actions/workflows/R-CMD-check.yaml)
[![Coverage](https://github.com/JanWein/tidyweave/actions/workflows/coverage.yaml/badge.svg)](https://github.com/JanWein/tidyweave/actions/workflows/coverage.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)

Deliver checked data with reusable specifications, recipes and workflows.
A changed column type or a negative amount blocks a delivery before its target
is written. The result retains the checks and offending rows for diagnosis.

DataRaft is under development. Before 1.0 the current API is the supported API.

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r failure}
library(dataraft)

orders <- dr_product("orders") |>
  dr_add_contract(c(id = "integer", amount = "numeric")) |>
  dr_add_quality(~ amount >= 0)

flow <- dr_workflow() |>
  dr_add_product(orders) |>
  dr_add_recipe(dr_recipe() |> dr_step_mutate(amount = round(amount, 2)))

bad_delivery <- data.frame(id = 1:3, amount = c(25, -75, 50))
checked <- dr_trial(flow, data = bad_delivery)
checked$status
dr_quality_rows(checked)
```

Correct the delivery and reuse the same workflow:

```{r success}
next_delivery <- data.frame(id = 1:3, amount = c(25, 75, 50))
result <- dr_trial(flow, data = next_delivery)
dr_collect(result)
```

A **product** defines identity, contract and quality requirements. A **recipe**
defines preparation in step order. A **workflow** binds these definitions to
sources and a target. Definitions do no I/O. `dr_trial()` checks without writing;
`dr_run()` executes the configured target. `dr_collect()` retrieves the output.

## Packages

| Package | Responsibility |
|---|---|
| `dataraft` | Metapackage and shared introduction |
| `dataraft.core` | Products, contracts, recipes, workflows and quality |
| `dataraft.lake` | Lake storage, releases, coordinated publication and recovery |
| `dataraft.adapters` | Database, API, Parquet and pins adapters; targets integration |
| `dataraft.dbt` | dbt execution and artifacts |
| `dataraft.catalog` | Catalog applications and metadata publication |
| `dataraft.metrics` | Metrics and frozen report evidence |

Use `library(dataraft.core)` for in-memory work without the extensions. Each
extension is independently installable with its declared dependencies. The
metapackage re-exports the public family API; its own R code contains no engine.

During development, install the family from a checkout:

```r
source("scripts/install-family.R")
```

## When to use it

Use DataRaft when repeated deliveries need a reusable preparation definition,
quality gates and an inspectable execution result. Use pointblank for standalone
data validation, targets for pipeline scheduling, pins for board-based object
storage and dbt for SQL model development. DataRaft adapters connect these tools
to a checked delivery workflow.

See the [introduction](vignettes/get-started.Rmd),
[integration guide](vignettes/integrations.Rmd) and
[guarantees and limits](vignettes/guarantees.Rmd).
