---
title: "Migrate to the tw_ API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrate to the tw_ API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
```

All tidyweave exports now start with `tw_`. The development API intentionally
has no unprefixed compatibility aliases. This avoids masking parsnip, recipes,
workflows and common R functions when packages are attached together.

## Update calls

For tidyweave-owned functions, add `tw_` to the previous name. For example:

| Previous call | Current call |
|---|---|
| `product()` | `tw_product()` |
| `recipe()` / `step_mutate()` | `tw_recipe()` / `tw_step_mutate()` |
| `workflow()` / `add_recipe()` | `tw_workflow()` / `tw_add_recipe()` |
| `set_engine()` | `tw_set_engine()` |
| `trial()` / `run()` / `publish()` | `tw_trial()` / `tw_run()` / `tw_publish()` |
| `collect()` / `explain()` | `tw_collect()` / `tw_explain()` |
| `source_database()` / `target_database()` | `tw_source_database()` / `tw_target_database()` |

Apply this only to tidyweave functions. Keep `recipes::recipe()`,
`parsnip::set_engine()` and `workflows::workflow()` unchanged. Keep argument names,
column names and result fields unchanged: `data =`, `product =`, `recipe =`,
`result$status` and `result$quality` still have their existing meanings.

`dplyr::collect()`, `dplyr::explain()` and dplyr transformation methods remain
supported. tidyweave no longer re-exports `filter()`: call `dplyr::filter()`
explicitly or attach dplyr. `tw_collect()` and `tw_explain()` delegate to the dplyr
generics, preserving backend dispatch.

## Separate an existing definition

Direct product pipelines still work under the new names. For modular reuse,
move preparation into a recipe and bind data at execution:

```{r}
library(tidyweave)
spec <- tw_product("orders") |> tw_add_quality(~ amount >= 0)
preparation <- tw_recipe() |> tw_step_mutate(amount = round(amount, 2))
flow <- tw_workflow() |> tw_add_product(spec) |> tw_add_recipe(preparation)
tw_collect(tw_trial(flow, data = data.frame(amount = c(10.123, 20))))
```

Do not pass a product with embedded transformations to `tw_add_product()`.
Rebuild its preparation in the recipe. Existing release identifiers and stored
table data do not need renaming. Recreate saved executable definitions from
their source code so their function references use the current API.

## Update extension methods

The S3 generics themselves have been renamed, not just their exported aliases.
Rename methods such as `read_source.my_source` to `tw_read_source.my_source` and
regenerate your extension package's NAMESPACE. Explicit registrations must use
the new generic string, for example `"tw_read_source"`. Class names, method
arguments and return contracts remain unchanged. See
[Write an adapter](extending-tidyweave.html) for the complete runnable protocol.
