---
title: "Build a monthly cancellation report"
description: "Define reusable contracts and products for customers, policies and brokers."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Build a monthly cancellation report}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**Goal:** define your own contracts and products, diagnose an invalid delivery,
and correct a cancellation report without changing an issued report.
You need tidyweave, dplyr and dm. Saving reports also needs DuckDB.
All data are synthetic. Run the visible code in order in a fresh session.

## Decide what the number means

For August 2026, the cancellation rate is the number of policies from the
opening portfolio cancelled during August divided by that opening portfolio.
It is a policy count, not a premium-weighted rate or a customer churn rate.
Dates are effective dates: start is inclusive and cancellation is exclusive.
A policy belongs to the opening portfolio if it starts before August 1 and
has no cancellation before August 1. Cancellation on August 1 counts; cancellation
on September 1 does not. Policies starting during August are excluded even if
cancelled during August. An empty opening portfolio makes the rate undefined, never zero.
The formula produces NA and `tw_measure()` refuses to issue that result.

Each delivery is a **complete snapshot**, including historical policies.
One customer may have many policies. Every policy has exactly one customer and
one broker. Their IDs are required and unique in their respective tables.
Missing references, duplicate keys and cancellation before inception block use.
Broker attribution uses the supplied snapshot, not an inferred historical mapping.
A correction replaces the snapshot; it is not a patch of individual rows.

```{r deliveries}
library(tidyweave)
library(dplyr)
customers_data <- tibble(
  customer_id = c("C1", "C2", "C3"),
  segment = c("Private", "Private", "Business")
)
brokers_data <- tibble(
  broker_id = c("B1", "B2"), channel = c("Broker", "Direct")
)
policies_data <- tibble(
  policy_id = paste0("P", 1:8),
  customer_id = c("C1", "C1", "C2", "C2", "C3", "C3", "C1", "C2"),
  broker_id = c("B1", "B1", "B2", "B2", "B1", "B2", "B1", "B2"),
  started_on = as.Date(c(rep("2026-01-01", 4), "2026-08-10",
    "2026-01-01", "2026-01-01", "2026-08-01")),
  cancelled_on = as.Date(c(NA, "2026-08-01", "2026-08-31", "2026-09-01",
    "2026-08-20", "2025-12-31", NA, NA))
)
```

## Define reusable contracts and products

These are ordinary named R objects. The same contracts can be reused for the
next delivery. Optional cancellation dates are excluded from `required`.

```{r definitions}
customer_contract <- tw_contract(
  columns = c(customer_id = "character", segment = "character"),
  key = "customer_id", grain = "One customer"
)
broker_contract <- tw_contract(
  columns = c(broker_id = "character", channel = "character"),
  key = "broker_id", grain = "One broker"
)
policy_contract <- tw_contract(
  columns = c(policy_id = "character", customer_id = "character",
    broker_id = "character", started_on = "Date", cancelled_on = "Date"),
  required = c("policy_id", "customer_id", "broker_id", "started_on"),
  key = "policy_id", grain = "One policy with at most one cancellation",
  rules = list(date_order = ~ is.na(cancelled_on) | cancelled_on >= started_on)
)
customers <- tw_product("customers", customers_data, contract = customer_contract)
brokers <- tw_product("brokers", brokers_data, contract = broker_contract)
policies <- tw_product("policies", policies_data, contract = policy_contract)
```

## Find and fix the bad delivery

```{r repair}
attempt <- tw_trial(policies)
tw_quality_rows(attempt)
stopifnot(identical(tw_quality_rows(attempt)$policy_id, "P6"))
fixed_policies <- policies_data
fixed_policies$cancelled_on[fixed_policies$policy_id == "P6"] <- as.Date("2026-07-31")
policies <- tw_replace_sources(policies, policies = fixed_policies)
stopifnot(nrow(tw_collect(tw_trial(policies))) == 8L)
```

P6 was cancelled before it started. The corrected source says July 31, so it
is outside August's opening portfolio. Checks do not guess or repair dates.

## Describe the model and choose a reporting grain

A dm model keeps related tables separate. It does not imply a single joined
table. Here the reporting view deliberately has one row per policy.

```{r relationships}
portfolio_model <- dm::dm(
  customers = tw_collect(tw_trial(customers)),
  policies = tw_collect(tw_trial(policies)),
  brokers = tw_collect(tw_trial(brokers))
) |>
  dm::dm_add_pk(customers, customer_id) |>
  dm::dm_add_pk(policies, policy_id) |>
  dm::dm_add_pk(brokers, broker_id) |>
  dm::dm_add_fk(policies, customer_id, customers) |>
  dm::dm_add_fk(policies, broker_id, brokers)
stopifnot(all(dm::dm_examine_constraints(portfolio_model)$is_key))

portfolio <- tw_product("portfolio", portfolio_model) |>
  tw_replace_sources(customers = customers, policies = policies, brokers = brokers)
checked_model <- tw_trial(portfolio)
tw_collect(checked_model)
```

The named table products retain their preparation and contracts. Alternatively,
pass raw dm tables with `contracts = list(customers = customer_contract, ...)`
to `tw_product()`. The model product rechecks all table contracts and dm relationships on every
execution. `tw_collect()` returns the related tables as a dm. Publication saves all
of them together. A new delivery uses its table name in `sources`.

## Define the monthly calculation once

Keep date logic in an ordinary reusable R function. Keep the metric separate
from the storage connection and the product preparation.

```{r calculation}
reporting_month <- function(data, from, until) {
  data |>
    mutate(
      opening = started_on < from & (is.na(cancelled_on) | cancelled_on >= from),
      cancelled = opening & !is.na(cancelled_on) &
        cancelled_on >= from & cancelled_on < until
    )
}
reporting_product <- function(model_result) {
  tw_product("august_portfolio", model_result, table = "policies") |>
    tw_add_lookup(model_result, table = "customers", by = "customer_id") |>
    tw_add_lookup(model_result, table = "brokers", by = "broker_id") |>
    tw_add_transform(\(data) reporting_month(data, as.Date("2026-08-01"),
      as.Date("2026-09-01")))
}
august <- reporting_product(checked_model)
define_cancellation_metrics <- function(approved = FALSE, code_version = NULL) {
  tw_metric_set(
  "august_portfolio",
  opening = sum(opening, na.rm = TRUE),
  cancellations = sum(cancelled, na.rm = TRUE),
  cancellation_rate = if (sum(opening, na.rm = TRUE) == 0) NA_real_ else sum(cancelled, na.rm = TRUE) / sum(opening, na.rm = TRUE),
  dimensions = "channel",
  units = c(opening = "policies", cancellations = "policies", cancellation_rate = "ratio"),
  na_policy = "expression", approved = approved, code_version = code_version
  )
}
cancellation_metrics <- define_cancellation_metrics()
preview <- tw_trial(august)
values <- tw_measure(preview, metrics = cancellation_metrics, by = character())
tw_collect(values)
stopifnot(nrow(tw_collect(preview)) == 8L,
  sum(tw_collect(preview)$opening) == 5,
  sum(tw_collect(preview)$cancelled) == 2)
```

Expected: 5 opening policies, 2 cancellations and a rate of **0.4 (40%)**.
Use `by = "channel"` for channel rates. Recompute the overall numerator and
denominator; averaging channel percentages gives the wrong answer.
No period aggregation rule should add these monthly rates together.
`na_policy = "expression"` delegates input missing-value handling to the formula;
it does not permit missing metric results. A channel with no opening policies
therefore blocks the rate calculation. Inspect the `opening` and `cancellations`
metrics separately and describe the rate as unavailable in your report. Do not
replace it with zero. Input contracts still reject missing required fields.

## Save the report and accept a late correction

The following steps need optional DuckDB. Approval here is an explicit declaration
for this synthetic example, not automatic approval of a production metric.

```{r storage-availability, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(eval = requireNamespace("duckdb", quietly = TRUE))
```

```{r save}
root <- tempfile("cancellation-reference-")
first_model <- tw_publish(portfolio, to = root)
first <- tw_publish(reporting_product(first_model), to = root)
reviewed_metrics <- define_cancellation_metrics(
  approved = TRUE, code_version = "opening-cohort-v1"
)
first_values <- tw_measure(first, metrics = reviewed_metrics, by = character())
tw_report_release(first_values, "august-original", code_version = "report-v1")

corrected_policies <- fixed_policies
corrected_policies$cancelled_on[corrected_policies$policy_id == "P7"] <- as.Date("2026-08-15")
second_model <- tw_publish(portfolio, to = root, previous = first_model,
  sources = list(policies = corrected_policies))
second <- tw_publish(reporting_product(second_model), to = root, previous = first)
second_values <- tw_measure(second, metrics = reviewed_metrics, by = character())
tw_report_release(second_values, "august-corrected", code_version = "report-v1")
original <- tw_report_read(root, "august-original", values_only = TRUE)
corrected <- tw_report_read(root, "august-corrected", values_only = TRUE)
original
corrected
stopifnot(original$value[original$.metric == "cancellation_rate"] == 0.4,
  corrected$value[corrected$.metric == "cancellation_rate"] == 0.6)
stopifnot(sum(tw_collect(first)$cancelled) == 2,
  sum(tw_collect(second)$cancelled) == 3,
  sum(tw_collect(second)$opening) == 5)
```

`previous` says which published state you corrected. A newer release causes a
conflict, so another person's correction cannot be silently replaced. The model
and the derived reporting table each publish atomically; the two calls are not
one transaction. If reporting preparation fails, repeat it from `second_model`.
The original model and report remain available.

The correction changes the rate from **40% to 60%**. The original report remains
40%. Use `tw_collect(second_values)` with Quarto or your usual reporting tool;
`tw_report_release()` saves values and evidence, not a rendered document.

```{r cleanup, include=FALSE}
unlink(root, recursive = TRUE)
```

[Back to Learn](learn.html) · [Create a local DuckLake](create-ducklake.html)
