---
title: "Check a delivery"
description: "Find an unacceptable row and rerun the same definition with corrected data."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Check a delivery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
```

**Goal:** find and fix an unacceptable delivery. **Needs:** tidyweave only.
This guide repeats its setup so you can run it independently.

## State what acceptable means

```{r definition}
library(tidyweave)
delivery <- data.frame(id = 1:3, amount = c(25, 75, 50))
spec <- tw_product("orders") |>
  tw_add_contract(tw_contract(
    columns = c(id = "integer", amount = "numeric"), key = "id")) |>
  tw_add_quality(~ amount >= 0)
orders <- tw_workflow() |> tw_add_product(spec)
```

A **contract** describes the table: these columns, these types, one row per `id`.
By default it also rejects missing required values, extra columns and empty
tables. A **quality rule** checks the values: every amount must be non-negative.
The `~` introduces a predicate evaluated against the table's columns.

These are business decisions. tidyweave cannot infer whether negative amounts
are valid refunds in your real data.

## Inspect a failed attempt

```{r failed}
bad_delivery <- data.frame(id = 1:3, amount = c(-25, 75, 50))
failed <- orders |> tw_trial(data = bad_delivery)
tw_status(failed)
tw_quality_report(failed)
tw_quality_rows(failed)
```

`tw_status()` summarizes the attempt. `tw_quality_report()` lists the checks and their
outcomes. `tw_quality_rows()` selects the single failed rule and returns its row:
order **1**, amount **-25**. It can expose source values, so inspect it locally.

`tw_trial()` returns a failed result for inspection by default. Use
`tw_trial(..., stop_on_failure = TRUE)` when you want a strict trial that stops.
`tw_run()` and `tw_publish()` still stop on failure by default. Returning a result
does not make unacceptable data pass. If several rules fail, `tw_quality_rows()`
lists the choices; select one by its rule name. Do not try to collect a
successful output from a blocked result.

## Correct the source and try again

```{r fixed}
corrected <- bad_delivery
corrected$amount[corrected$id == 1L] <- 25
accepted <- orders |> tw_trial(data = corrected)
tw_collect(accepted)
```

The corrected total is **150**. The same contract and quality rule still apply.
The check did not silently remove a row or change its value.

Checks run on the prepared product output. If you need separate checks on
received data before preparation, that is the later [receipt workflow](layered-data-stack.html).

```{r verify, include=FALSE}
stopifnot(failed$status == "blocked", sum(tw_collect(accepted)$amount) == 150,
  tw_quality_rows(failed)$id == 1L)
```

## Continue

[Previous: Build a product](first-product.html) · [Next: Publish and correct data](publish-data.html) · [Get started](get-started.html)
