---
title: "Compose products and combine tables"
description: "Use checked lookups, reusable products and lazy sources."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compose products and combine tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Use this after [Build a product](first-product.html) when one table is no
longer enough. The first six sections need no optional packages; storage and
lazy database examples need DuckDB. Each section builds on the preceding one.

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
```

A product names the table you want and describes how to produce it. The smallest
useful product has one source. Everything else in this guide is an addition,
not a prerequisite. No service or database is needed for the first six steps.

## 1. Make your first result

```{r minimal}
library(tidyweave)
input <- data.frame(id = 1:3, amount = c(25, 75, 50))
orders <- tw_product("orders", input)
result <- tw_trial(orders)
tw_collect(result)
stopifnot(sum(tw_collect(result)$amount) == 150)
```

`orders` is the definition. `result` is one execution with data, status and
metadata. `tw_collect(result)` returns a tibble. Nothing has been written to disk.

Sources accept an R table, a file path, a zero-argument function or an adapter.
For example, `tw_add_source("orders.csv")` chooses a CSV reader. Excel uses optional
readxl. Defining or printing a product does not fetch its data.

## 2. Add preparation and a business check

```{r transform-and-quality}
orders <- orders |>
  dplyr::mutate(amount = round(amount, 2)) |>
  tw_add_quality(~ amount >= 0, name = "nonnegative")
tw_quality(tw_trial(orders))
```

Supported methods are `mutate()`, `filter()`, `select()`, `rename()`,
`relocate()`, `arrange()`, `group_by()`, `ungroup()`, `summarise()`, `distinct()`
and `count()`. They capture expressions without evaluating the source.
At execution, they invoke real dplyr on the ordinary or lazy table. External
bindings keep ordinary R semantics: changing a captured value can change a later
run. Inject `!!value` to freeze a small value, and supply an explicit code version
when enabling cached publication. Use qualified
calls such as `dplyr::filter()` when another attached package has the same name.
Unsupported verbs and R algorithms can be passed through `tw_add_transform()`.
Transformations run in addition order. A check happens after preparation and
never silently drops bad rows. Formulas and functions can return a logical
value per row; `NA` counts as failure. Use `engine = "pointblank"` for the same simple predicate with optional
Pointblank validation. Advanced agents still use `tw_pointblank_checks()`. A function
may also return one overall decision or `tw_quality_counts(n_failed, n_total)` for an aggregate check.

For an R-only transform following a lazy source, collect explicitly:
`tw_add_transform(function(data) my_r_algorithm(tw_collect(data)))`.

## 3. State the table you expect, if you need to

```{r contract}
orders <- orders |> tw_add_contract(c(id = "integer", amount = "numeric"))
```

This compact contract requires the declared columns and non-missing values.
`numeric` accepts integer or decimal values; `integer` is stricter. Explicit
contracts reject extra columns and empty tables by default. For a business key,
use a reusable contract:

```{r reusable-contract}
order_contract <- tw_contract(
  columns = list(id = integer(), amount = double()), key = "id"
)
orders <- orders |> tw_add_contract(order_contract)
```

The zero-length values describe types, not sample rows. Use
`as.Date(character())` for dates. Descriptions, owners, freshness deadlines and
manual versions are optional. Without a contract, tidyweave checks inferred
structure and nonempty data; it does not guess your business key.
`tw_validate(input, order_contract)` returns a quality table immediately.

## 4. Inspect the definition and a failed attempt

```{r inspect}
orders
tw_explain(orders)
tw_validate(orders)
```

Preflight checks configuration, required dependencies and supported components.
It does not execute a source callback or predict remote data quality. `tw_run()`
always performs preflight, so a separate call is optional.

```{r failed-attempt}
bad_input <- data.frame(id = 1:2, amount = c(10, -1))
blocked <- tw_trial(orders, data = bad_input)
tw_quality_report(blocked)
stopifnot(blocked$status == "blocked")
```

`data` replaces the submitted delivery for this attempt. For several sources,
use explicit `sources = list(name = new_data)` values. Unknown or ambiguous
names fail before execution. `tw_add_source()` edits the definition itself.
A trial returns a failed result for diagnosis by default; `tw_run()` and
`tw_publish()` stop on failure unless explicitly told to return that result.

## 5. Enrich a table with a checked lookup

```{r checked-lookup}
customers <- data.frame(customer = c(1L, 2L), region = c("North", "South"))
sales <- data.frame(customer = c(1L, 2L), amount = c(100, 250))
regional <- tw_product("regional_orders", sales) |>
  tw_add_lookup(customers, by = dplyr::join_by(customer), name = "customers")
tw_collect(tw_trial(regional))
```

The primary table remains a table. Parent keys must be unique and non-missing,
and child keys must match by default. The join cannot silently multiply or lose
orders. `engine = "dm"` selects optional dm checks with the same contract.
Only equality lookups are supported; general joins remain ordinary R functions.
Several explicitly named primary sources still support complex transformations:
`tw_add_transform()` receives their named list and returns one table.

The lookup name is also its correction name. For example:

```{r correct-reference}
corrected_customers <- customers
corrected_customers$region[corrected_customers$customer == 2L] <- "North"
corrected_result <- tw_trial(regional, sources = list(customers = corrected_customers))
tw_collect(corrected_result)
```

This uses the same preparation and checks with the corrected reference table.
The call does not rewrite the original definition. Supply all changed deliveries
in `sources`, or use `tw_replace_sources()` to retain corrected defaults in a new
definition. [Three deliveries](three-deliveries.html) follows both a payment
correction and a reference correction through issued reports.

## 6. Reuse a product as an input

```{r nested-product}
summary <- tw_product("regional_summary", regional) |>
  dplyr::summarise(total = sum(amount))
summary_result <- tw_trial(summary)
tw_collect(summary_result)
stopifnot(tw_collect(summary_result)$total == 350)
```

There is one product class for both source preparation and derived tables.
Within one execution, shared product dependencies run once. Preflight rejects
cycles and conflicting definitions with the same product identity. A product
definition is not a frozen input. Pass a successful lake result directly when you
need its exact immutable release: `tw_product("export", published)`. Other successful
results reuse their retained submitted table or lazy query. A live query keeps
its backend's mutability; a DBI append result represents the submitted batch.

## 7. Choose a destination

| Destination | Add to the product | Important boundary |
|---|---|---|
| R session | Nothing | Result stays in memory |
| Lake with history | `tw_set_target(tw_target_lake("reporting-lake"))` | Immutable releases; one coordinated writer |
| Database table | `tw_set_target(tw_target_database(con, "orders"))` | Driver transaction and isolation guarantees |
| Local Parquet file | `tw_set_target(tw_target_parquet("orders.parquet"))` | Single file; Arrow optional; no release registry |
| Pin | `tw_set_target(tw_target_pins(board, "orders"))` | Versioning depends on the pins board |

`tw_run()` executes the configured target. `tw_publish()` supplies a local lake target
if the product has none. Publishing means writing data to that destination,
not publishing it on the internet.

```{r durable-example, eval=requireNamespace("duckdb", quietly=TRUE), purl=FALSE}
local({
  root <- tempfile("orders-lake-")
  on.exit(unlink(root, recursive = TRUE))
  published <- orders |> tw_publish(to = root)
  stopifnot(published$status == "published")
  tw_collect(published)
})
```

The lake gate checks the complete candidate before making it current. It also
archives inputs and preserves earlier releases. The lake adapter materializes
its input for archival; it is not an end-to-end lazy warehouse executor.

## 8. Keep database work lazy

```{r lazy-example, eval=requireNamespace("duckdb", quietly=TRUE), purl=FALSE}
local({
  con <- DBI::dbConnect(duckdb::duckdb())
  on.exit(DBI::dbDisconnect(con, shutdown = TRUE))
  DBI::dbWriteTable(con, "orders", input)
  remote <- tw_product("large_orders") |>
    tw_add_source(tw_source_database(con, table = "orders")) |>
    dplyr::filter(amount >= 50) |>
    tw_add_quality(~ amount >= 0)
  result <- tw_run(remote)
  tw_collect(result)
})
```

Supported dplyr transformations and native checks can run against a lazy DBI
table. Data collection is explicit at `tw_collect()` or required by a selected
adapter. Arbitrary R functions and SQL translations are not automatically
portable between backends. `tw_capabilities()` and `tw_explain()` show declared
boundaries. Arrow datasets use the same idea with `tw_source_parquet()`.

Keep a caller-supplied connection open while using its lazy result. A source
factory is useful when tidyweave should own the connection; factory-backed
sources return materialized data and close their owned connection. Requesting
lazy output from a factory is rejected.
See [resource ownership](workflow-design.html#resource-ownership).

## 9. Keep execution evidence separately

```{r evidence}
evidence_path <- tempfile("orders-runs-")
result <- tw_run(orders, evidence = evidence_path)
tw_run_history(evidence_path)
tw_read_run(evidence_path, result$run_id)$status
unlink(evidence_path, recursive = TRUE)
```

Evidence stores run metadata and check counts, not source rows or credentials.
It also records failed and blocked attempts. Catalog delivery is separate from
writing data: a catalog outage does not undo a published output. Use
`tw_retry_catalogs()` with freshly supplied adapters to retry pending deliveries.
OpenLineage events are delivered after execution, not as live progress updates.

Keep executable definitions in version-controlled R files. Inspection metadata
cannot recreate closure environments, live connections or installed packages.
Use `tw_init_project()` for a small project template and optional targets, renv
or Connect files. Continue with [monthly reporting](getting-started.html),
[optional integrations](optional-integrations.html),
[workflow design](workflow-design.html), or [extensions](extending-tidyweave.html).


[Back to Learn](learn.html) · [Troubleshooting](troubleshooting.html)
