---
title: "From one table to a reusable data product"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From one table to a reusable data product}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(lakefold)
```

## Why does lakefold exist?

You have data, some preparation code, and checks that must pass before anyone
uses the result. At first, one R script is enough. Later you need to rerun it,
change its source, explain a failure, keep an earlier result, or reuse the same
preparation in several reports.

lakefold gives these operations one consistent structure. Describe a **data
product**, add the parts it needs, then run it. Your data remain ordinary R
tables and your preparation remains ordinary R code. You can introduce durable
storage and specialist integrations later.

A data product is simply a named table plus the instructions for producing and
checking it. It does not require a server, an owner field, a catalog, a database,
or a manually maintained version number.

lakefold does not provide a scheduler, a database driver, a complete enterprise
catalog, authentication, or a replacement for dbt or pointblank. Those tools
retain their own responsibilities.

## Step 1: make a useful result

```{r minimal}
orders <- data.frame(id = 1:3, amount = c(25, 75, 50))

product <- dl_product("orders") |>
  dl_add_source(orders)

result <- dl_run(product)
dl_collect(result)
stopifnot(sum(dl_collect(result)$amount) == 150)
```

There are four ideas here:

1. `dl_product("orders")` names the result you want.
2. `dl_add_source(orders)` supplies its input.
3. `dl_run()` executes the instructions.
4. `dl_collect()` returns a tibble you can use in ordinary R code.

**Nothing has been saved to disk.** The result and its execution evidence are
in memory. DuckDB and other optional integrations are unnecessary for this
workflow. Creating or printing the product does not read files or run callbacks.

## Step 2: add an ordinary transformation

```{r transform}
clean_orders <- function(data) {
  transform(data, amount = round(amount, 2))
}

product <- product |>
  dl_add_transform(clean_orders, name = "round_amount")
```

Each transformation receives a data frame or tibble and returns one. Add more
transformations in the order you want them applied. An R function needs no
special specification class. A formula such as
`~ dplyr::filter(.x, amount >= 0)` also works.

Preparation is your decision. Quality checks never silently delete bad rows.
For example, rounding an amount is preparation; rejecting a negative amount is
a quality decision.

## Step 3: describe the table you expect

```{r contract}
product <- product |>
  dl_add_contract(c(id = "integer", amount = "numeric"))
```

This requires two columns with the declared types and non-missing values.
`numeric` accepts integers and decimals; `integer` stays strict. Explicit
contracts reject extra columns and empty data unless you opt into them.

For reusable rules or optional metadata, create a contract explicitly:

```{r richer-contract}
order_contract <- dl_contract(
  columns = list(id = integer(), amount = double()),
  key = "id",
  owner = "Analytics",
  description = "One row per order"
)
product <- product |> dl_add_contract(order_contract)
```

`integer()` and `double()` describe types; they are zero-length prototypes,
not sample values. Use `as.Date(character())` for a Date column.
The anonymous contract automatically gets the product's identity when added.
Ownership, description, column metadata, keys and a manual version are optional.
A plain type vector is often enough.

Without an explicit contract, lakefold checks inferred structure and rejects
empty outputs. It does not guess business keys or non-missing requirements.
A lake target remembers the first published automatic schema for future writes.
An in-memory run has no persistent previous schema to compare with.

## Step 4: add a quality rule

```{r quality}
product <- product |>
  dl_add_quality(~ amount >= 0, name = "nonnegative")

result <- dl_run(product)
dl_quality(result)
stopifnot(all(dl_quality(result)$status == "passed"))
```

A formula evaluates one logical result per row. Missing results count as
failures. A function can instead return one logical decision or
`dl_quality_counts(failed, total)` for a custom aggregate check.

For tolerances or warning-only checks, use the optional explicit rule object:

```{r warning-rule}
dl_rule("small_amounts", ~ amount >= 10,
  severity = "warning", max_failure = 0.05)
```

The native engine has no pointblank dependency. `dl_pointblank()` delegates
advanced checks to pointblank and retains its existing threshold policies.
Both engines return the same quality evidence. Failed, errored, skipped and
empty checks block publication; warnings permit it.

To validate an existing table immediately, use
`dl_validate(orders, order_contract)`. It returns a **quality report**, not the
input data. To compose preparation and publication, use the product verbs.

## Step 5: inspect before running

```{r inspect}
product
product |> dl_validate()
product |> dl_plan()
product |> dl_explain()
```

These operations do different jobs:

| Call | What you learn |
|---|---|
| Print the product | A concise overview |
| `dl_inspect(product)` | Its source, transformations, contract and target |
| `dl_validate(product)` | Whether the definition and required adapters are configured |
| `dl_plan(product)` | The execution order |
| `dl_explain(product)` | A plain-language explanation of the planned work |

Preflight checks missing files, missing dependencies, missing sources, duplicate
rules, unsupported components and read-only targets. It does not run your source
function, query a remote table or predict data quality. `dl_run()` repeats
preflight automatically.

The product stays a definition. The run is a separate result with timestamps,
status, quality evidence, source and output information, schema, row count and
lineage. No mutable platform object owns the whole system.

## Step 6: publish when you need durable history

Install the optional `duckdb` package (version 1.5.5 or later), then choose a
local folder:

```r
published <- product |> dl_publish(to = "reporting-lake")
dl_collect(published)
```

Or save the target in the product and use the same execution verb:

```r
product <- product |> dl_add_target("reporting-lake")
published <- dl_run(product)
```

`dl_publish()` defaults to the `lakefold` folder if no target was supplied.
Publication makes data available in your selected storage; it does not upload
it to the public internet.

With a lake target, each successful publication creates an immutable release.
Original files are archived before parsing. Data-frame and DBI sources are
archived as RDS snapshots of the received table. The full candidate, including
retained partitions, is checked before the release marker is committed.
Earlier releases remain readable after a correction. Failed checks leave the
current successful release available.

```{r publication, eval=requireNamespace("duckdb", quietly=TRUE)}
local({
  root <- tempfile("lakefold-guide-")
  on.exit(unlink(root, recursive = TRUE))
  first <- product |> dl_publish(to = root)
  stopifnot(sum(dl_collect(first)$amount) == 150)
  bad <- product |> dl_add_source(data.frame(id = c(1L, 1L), amount = c(10, -1)))
  blocked <- bad |> dl_publish(to = root, stop_on_failure = FALSE)
  stopifnot(blocked$status == "blocked")
  stopifnot(sum(dl_collect(first)$amount) == 150)
  dl_status(blocked)
})
```

For a table you already prepared, the immediate shortcut is:

```r
orders |> dl_publish("orders", to = "reporting-lake")
```

The existing `dl_open()`, `dl_write()` and `dl_read()` interface remains useful
when you want to manage one connection explicitly.

## Sources: change acquisition without rewriting the workflow

| Input | Example |
|---|---|
| Table in R | `dl_add_source(orders)` |
| CSV, TSV, RDS or Excel file | `dl_add_source("orders.xlsx")` |
| Custom file reader | `dl_add_source(path, reader = my_reader)` |
| API/client callback | `dl_add_source(fetch_orders)` |
| DBI table | `dl_add_source(dl_source_database(con, table = "orders"))` |
| Parameterized query | `dl_source_database(con, query = "SELECT * FROM orders WHERE id > ?", params = list(100))` |

Excel uses optional readxl. An existing API, S3 or HANA client can be called from
a source function; lakefold does not invent another connection protocol.
`dl_source_database()` accepts ordinary DBI connections and `DBI::Id()` names.
A connection factory defers opening until execution and gives lakefold ownership
of closing that connection. Existing connections remain caller-owned.

## An advanced workflow with optional components

This complete example combines a DBI source, DuckDB SQL transformation,
pointblank quality checks, governed storage and a metadata callback. Only the
adapter configuration is technical; the product grammar is unchanged.

```{r advanced, eval=requireNamespace("duckdb", quietly=TRUE) && requireNamespace("pointblank", quietly=TRUE)}
local({
  root <- tempfile("lakefold-advanced-")
  dir.create(root)
  on.exit(unlink(root, recursive = TRUE), add = TRUE)

  source_con <- DBI::dbConnect(duckdb::duckdb())
  on.exit(DBI::dbDisconnect(source_con, shutdown = TRUE), add = TRUE)
  DBI::dbWriteTable(source_con, "orders", orders)

  target <- dl_target_lake(file.path(root, "published"))
  save_metadata <- function(metadata) {
    jsonlite::write_json(metadata, file.path(root, "run.json"),
      auto_unbox = TRUE, pretty = TRUE, null = "null", na = "null")
  }

  advanced <- dl_product("orders", code_version = "example-v1") |>
    dl_add_source(dl_source_database(source_con, table = "orders")) |>
    dl_add_transform(dl_sql("SELECT id, round(amount, 2) AS amount FROM data")) |>
    dl_add_contract(c(id = "integer", amount = "numeric")) |>
    dl_add_quality(dl_pointblank("amounts", function(data) {
      pointblank::create_agent(data) |>
        pointblank::col_vals_gte("amount", 0)
    })) |>
    dl_add_target(target) |>
    dl_add_catalog(save_metadata)

  result <- advanced |> dl_validate() |> dl_run()
  stopifnot(result$status == "published")
  stopifnot(file.exists(file.path(root, "run.json")))
  stopifnot(sum(dl_collect(result)$amount) == 150)
  stopifnot(DBI::dbIsValid(source_con))
  dl_status(result)
})
```

To use DuckLake, replace only the target with a `dl_target_lake(dl_config(...,
backend = "ducklake"))`. PostgreSQL catalog and S3 storage details belong inside
that configuration. They do not become arguments to every transformation.
A backend switch selects a destination; it does not migrate previous data.

The SQL adapter executes trusted SQL against a private temporary DuckDB table
named `data`. For a remote SQL query use the DBI source, including bound
parameters. These are intentional escape hatches, not SQL sandboxing.

## dbt and other specialist tools

Use `dl_run(dl_dbt_project(...))` or `dl_dbt_build()` for dbt's own model graph,
materializations and tests. Use `dl_dbt_publish()` to snapshot and validate a
successful model as a governed release. A dbt model is not presented as a
function of an in-memory R table: dbt owns its connections and project inputs.
Close local catalog connections before invoking dbt in another process.
The [dbt guide](dbt-workflows.html) contains a runnable starter.

Existing multi-input derived products still use named pinned releases and a
builder through `dl_product(inputs = ..., build = ..., contract = ...,
code_version = ...)`. They remain executable with `dl_run()` or `dl_build()`.
This preserves established lazy database behavior for larger workflows.

## Metadata, versions and practical limits

Most metadata comes from execution: timestamps, input hashes, source descriptors,
output references, schema, row counts, checks and lineage. Business ownership
and descriptions remain optional. Catalog callbacks receive metadata, not input
rows or live connections. Custom descriptors and written code can still include
sensitive literals: keep credentials in runtime configuration, not source code.

An in-memory run retains its evidence in the returned result. A lake run also
records durable ingestion and release evidence. Source-factory and preflight
failures occurring before a lake ingestion starts are returned locally and do
not create a registry run. Catalog delivery occurs after successful data
publication; a catalog failure produces a warning without undoing the data.

Composed products rerun callbacks by default. Enable `cache = TRUE` only with
an explicit `code_version` covering imported functions, captured values and
dependencies. Cache reuse selects only the current matching release.
Explicit product and contract versions reject changed registered definitions.
Do not assume an inspected JSON description can recreate a closure environment
or live DBI connection. Keep executable product definitions in versioned R code.

The new composition path materializes source data in R and gives ordinary R
tables to transformations. For workloads larger than R memory, use the existing
lazy `dl_pipeline()` and derived-product APIs or delegate processing to dbt.
The lake registry still requires one coordinated writer. Custom targets have
only the guarantees their adapter provides.

## Why this architecture

The external API is intentionally small: **name, add, run, collect, publish**.
It follows tidyverse principles of readable verbs, pipes and ordinary objects.
A beginner does not need to learn adapters, execution plans or metadata schemas.

Internally, lakefold separates normalization, specifications, validation,
planning and execution. This takes structural inspiration from tidymodels:
components can be exchanged while the composition remains stable. There are no
model-fitting concepts or copied machine-learning APIs.

![The user describes a product with simple verbs. Internally it is normalized and validated, then executed through native R, lake storage or a custom adapter.](figures/composition-architecture.svg)

Ordinary functions remain functions. Explicit objects exist where they add
meaning: a contract, a database query with connection ownership, a SQL
transformation, a product or a storage target. The lake adapter reuses the
proven publication engine; the native executor shares the same source,
transformation and quality interfaces with custom targets.

See the [extension guide](extending-lakefold.html) for a complete S3 adapter,
method contracts and tests you can use in another R package.
