---
title: "Create a DuckLake and reopen it tomorrow"
description: "Create a real local DuckLake, save data and reopen the folder."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Create a DuckLake and reopen it tomorrow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Use this after [Publish and correct data](publish-data.html) when you want
the actual DuckLake backend. The basic lessons use ordinary DuckDB. A **layer**
is a named storage area; custom layers are optional.

One folder holds your local DuckLake configuration, metadata, stored data and
received deliveries. No PostgreSQL server or S3 account is needed for this local
example. Install `tidyweave` and optional `duckdb` (>= 1.5.5) first. DuckDB must
be able to load its DuckLake extension; it installs it automatically if needed.

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
  eval = requireNamespace("duckdb", quietly = TRUE) &&
    identical(Sys.getenv("TIDYWEAVE_TEST_DUCKLAKE"), "true"))
```

## Create it once

```{r create}
library(tidyweave)
folder <- tempfile("my-ducklake-") # Use a durable project folder for real work.
lake <- open_lake(folder, backend = "ducklake",
  layers = c("raw", "staging", "core", "marts"))
lake
```

Choose an empty or new folder. `backend = "ducklake"` explicitly selects DuckLake.
Without it, a new folder uses ordinary DuckDB, which needs no extension download.
Reopening always uses the saved backend. Custom layers are optional; the defaults
are `raw`, `validated`, `products`.

## Save your first checked table

```{r save}
orders <- product("orders", data.frame(id = 1:3, amount = c(100, 200, 50))) |>
  add_quality(~ amount >= 0)
first <- publish(orders, to = lake, layer = "core")
collect(first)
stopifnot(sum(collect(first)$amount) == 350)
close_lake(lake)
```

`publish()` checks and stores the table in the chosen layer. It does not publish
on the internet. Close the connection when finished, including before opening
the same local catalog from another program.

## Reopen using only the folder

```{r reopen}
lake <- open_lake(folder)
lake
read_release(lake, "orders")
stopifnot(identical(lake$config$backend, "ducklake"),
  identical(lake$config$layers, c("raw", "staging", "core", "marts")))
```

The backend and layers are remembered, including their order and any named
roles. You do not need to rebuild a configuration or manage a database connection
string. Repeating settings is allowed only if they agree with the saved choices.

## Save another delivery in the same layer

```{r next-delivery}
corrected <- data.frame(id = 1:3, amount = c(100, 200, 80))
second <- publish(orders, data = corrected, to = lake, layer = "core")
collect(second)
stopifnot(sum(collect(second)$amount) == 380,
  sum(read_release(lake, "orders", first$release_id)$amount) == 350)
close_lake(lake)
```

The second complete delivery totals 380. The earlier saved result remains 350.
You can keep `first` to refer to that original result; the assertion above checks
its stored version explicitly.

## Open just for reading

```{r read-only}
lake <- open_lake(folder, read_only = TRUE)
read_release(lake, "orders")
stopifnot(sum(read_release(lake, "orders")$amount) == 380)
close_lake(lake)
```

This does not change the saved folder configuration or create extra layers.

## If you already use another entry point

`setup_lake(path = folder)` opens the same folder with the same saved settings.
For a definition you can pass around without opening a connection, use
`lake_config(path = folder)`. These are the same local layout, not different
types of lake. Remote storage and split catalog/data paths still use explicit
`lake_config()` constructors followed by `connect_lake()`.

Older folders did not save their layer choices. Standard folders upgrade
automatically on a writable open. If an older folder has custom schemas, the
error shows their names: supply your original `layers` once in the intended order.
That first successful writable open remembers them for later. A read-only open
can use explicit layers but does not save or migrate the configuration.

```{r cleanup, include=FALSE}
unlink(folder, recursive = TRUE) # Only remove this disposable tutorial folder.
```


[Back to Learn](learn.html) · [Troubleshooting](troubleshooting.html)
