---
title: "Use dbt for SQL transformations"
description: "Connect an existing or generated dbt project to checked data products."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use dbt for SQL transformations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Use this when you already need dbt SQL models. Read [Publish and correct
data](modular-workflows.html) first. You need DuckDB and a separately installed
dbt CLI; installation is described below. Credential-dependent snippets are
configuration examples, not steps required for the beginner lessons.

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tidyweave)
root <- tempfile("tidyweave-dbt-guide-")
```

Use ordinary R functions until SQL models, dbt tests or incremental strategies
make dbt useful. tidyweave connects a product to dbt; dbt keeps responsibility for
its SQL graph. Choose the entry point that matches your work:

| Your task | Entry point | Result |
|---|---|---|
| Run an existing dbt project | `tw_dbt_project(...) |> tw_run()` | dbt result and invocation artifacts |
| Transform a product's incoming table with a dbt model | `tw_transform_dbt()` | Staged input, selected dbt build and materialized output |
| Inspect current dbt relations and declared keys | `tw_dbt_model()` | Lazy `dm` model borrowing your lake connection |
| Publish one successful dbt relation as an immutable release | `built |> tw_publish("model")` | A quality-gated governed release |

The first three operations do not make dbt tables immutable. A product can add a
lake target for a governed snapshot of its final output. Data quality failures
prevent that publication; they do not undo dbt's earlier external writes.

## Install dbt separately

The integration uses the CLI through processx and reads its JSON artifacts. A
tested engine combination is `dbt-core==1.12.5`, `dbt-duckdb==1.10.1` and Python
`duckdb==1.5.5`. These are test pins, not a claim that they are the latest releases.

```sh
python -m venv .venv
.venv/bin/python -m pip install dbt-core==1.12.5 dbt-duckdb==1.10.1 duckdb==1.5.5
.venv/bin/dbt --version
```

On Windows use `.venv/Scripts/python.exe` and `.venv/Scripts/dbt.exe`. Keep the R
and Python DuckDB engines compatible. Configure other dbt engines and adapters
independently; the local starter below generates dbt-duckdb profiles, not a dbt v2
catalog configuration. Their production behavior requires separate verification.

## Start with an existing project or generate an example

```{r existing-project}
config <- tw_lake_config(path = "reporting-lake", backend = "duckdb",
  layers = c("raw", "staging", "core", "marts"))
existing <- tw_dbt_project("analytics", lake = config)
existing
```

Construction creates no profile, source file or connection. Supply
`sources = list(orders = accepted_result)` for managed bindings to exact releases
under the `inputs` source group. At execution, tidyweave prepares the local
DuckDB/DuckLake profile and managed source YAML. Existing SQL continues to use
ordinary `source('inputs', 'orders')` expressions.

For other adapters, or a profile managed by your deployment, use
`tw_dbt_project("analytics", profiles_dir = "analytics")` instead. The following
optional starter creates example SQL and project files:

```{r scaffold, eval=requireNamespace("duckdb", quietly=TRUE) && requireNamespace("yaml", quietly=TRUE)}
config <- tw_lake_config(
  path = file.path(root, "lake"), backend = "duckdb"
)
project <- tw_dbt_init(file.path(root, "dbt"), config)
list.files(project$path, recursive = TRUE)
```

The starter has synthetic orders, a staging model, customer revenue and key/null
tests. Its profile attaches the configured catalog as `lake`. Its schema macro
uses exact schema names; keep development and production in separate catalogs.

These CLI examples are displayed without execution during vignette builds.
Set `executable` in `tw_dbt_project()` when dbt is not on your `PATH`.

```{r build, eval=FALSE}
# Initialize schemas, then release the local catalog lock before dbt starts.
lake <- tw_connect_lake(config)
tw_close_lake(lake)

built <- project |> tw_run(echo = FALSE)
tw_dbt_status(built)
tw_dbt_lineage(built)
tw_dbt_test(project, select = "customer_revenue", echo = FALSE)
```

`select` and `exclude` use dbt selection expressions, such as
`"+customer_revenue"`. `vars` accepts a named R list; credentials belong in the
profile's environment settings. `full_refresh = TRUE` requests a full refresh,
and `timeout` limits the child process. dbt owns dependencies, test selection and
incremental logic.

## Put dbt inside a product

`tw_transform_dbt()` requires an explicit staging table. This makes its input
meaningful: incoming rows are written there before the selected model runs.
The model must read that staging table. Use a dedicated table that no seed or dbt
model overwrites, and coordinate writers to it.

For the generated project, add a new source definition and model. These are dbt
files, with no hidden copy of dbt's dependency engine inside tidyweave:

```{r staged-model, eval=requireNamespace("duckdb", quietly=TRUE) && requireNamespace("yaml", quietly=TRUE)}
yaml::write_yaml(
  list(version = 2L, sources = list(list(
    name = "tidyweave_input", database = "lake", schema = "raw",
    tables = list(list(name = "orders", identifier = "staged_orders"))
  ))),
  file.path(project$path, "models", "product_sources.yml")
)
writeLines(c(
  "select customer_id, sum(amount) as revenue",
  "from {{ source('tidyweave_input', 'orders') }}",
  "group by customer_id"
), file.path(project$path, "models", "marts", "staged_customer_revenue.sql"))
```

Define a connection factory returning a **new** connection each time. The adapter
closes its staging connection before invoking dbt, then opens another connection
to read the successful relation. All factory-owned connections close, including
after failure. The returned output is an ordinary table.

```{r dbt-transform, eval=FALSE}
connection <- function() {
  DBI::dbConnect(duckdb::duckdb(), dbdir = config$catalog$path,
    bigint = "integer64")
}
orders <- data.frame(order_id = 1:3, customer_id = c(101L, 101L, 102L),
  amount = c(25, 75, 50))

revenue <- tw_product("customer.revenue") |>
  tw_add_source(orders) |>
  tw_add_transform(tw_transform_dbt(
    project, model = "model.tidyweave_demo.staged_customer_revenue",
    connection = connection,
    input = DBI::Id(schema = "raw", table = "staged_orders")
  ), name = "dbt") |>
  tw_add_quality(~ revenue >= 0)

result <- tw_run(revenue, evidence = file.path(root, "runs"))
tw_collect(result)
result$metadata$transformations$dbt
```

The adapter currently accepts model IDs of the form `model.package.name`. Use
`tw_dbt_build()` directly for versioned-model selection and other specialized dbt
selectors. The requested materialized model must succeed in the current invocation.
Failed selected models or tests fail the transform. The metadata
records the model, invocation ID and artifacts directory, even if later R
transformations follow. `input` is replaced every run; dbt controls whether its
output model uses an incremental strategy.

Staging and dbt writes are external side effects. A later failed product contract
cannot roll them back. Add `tw_set_target(tw_target_lake(config))` when the final checked
result also needs an immutable release; do not interpret that as an atomic commit
of the entire dbt project.

## Start from a pinned release

A governed release can be another product's source. The reference resolves to one
specific release and is recorded in the input metadata.

```{r release-source, eval=requireNamespace("duckdb", quietly=TRUE) && requireNamespace("yaml", quietly=TRUE)}
released <- tw_product("orders") |>
  tw_add_source(data.frame(order_id = 1:3, amount = c(25, 75, 50))) |>
  tw_set_target(tw_target_lake(config)) |>
  tw_run()

prepared <- tw_product("orders.prepared", released) |>
  dplyr::filter(amount >= 50) |>
  tw_run()
input_for_dbt <- tw_collect(prepared)
prepared$inputs[, c("asset", "release_id")]
```

The result supplies its exact release reference. Execution opens and closes its
own read connection, so no manual connection is needed for this step.

Pass `input_for_dbt` to a product with an appropriate `tw_transform_dbt()` step.
Collect and close borrowed local catalog connections before dbt starts. For a managed dbt project, pass the successful result through
`sources = list(orders = released)`. It binds the exact physical relation and
records its release lineage. Rebuilding with a correction requires selecting
that new result explicitly; models keep their logical source names.

## Inspect current relations or publish a snapshot

```{r relational-model, eval=FALSE}
lake <- tw_connect_lake(config)
relations <- tw_dbt_model(lake, built,
  tables = c(orders = "model.tidyweave_demo.stg_orders",
    customers = "model.tidyweave_demo.customer_revenue"),
  primary_keys = list(orders = "order_id", customers = "customer_id"),
  foreign_keys = list(list(table = "orders", columns = "customer_id",
    ref_table = "customers", ref_columns = "customer_id")))
dm::dm_examine_constraints(relations)
dplyr::collect(relations$customers)

schema <- tw_contract(columns = c(customer_id = "integer", revenue = "numeric"))
release <- tw_publish(built, "model.tidyweave_demo.customer_revenue",
  to = lake, contract = schema,
  asset = "revenue.snapshot", code_version = "reporting-v1")
tw_read_release(lake, "revenue.snapshot", release$release_id)
tw_close_lake(lake)
```

`dm` checks explicit relational keys; SQL dependencies alone do not establish
primary or foreign keys. `tw_dbt_model()` borrows the connection and reads current
relations. `tw_publish()` validates and snapshots one relation visible now,
using the dbt invocation as provenance. Coordinate writers between build and
snapshot. `tw_model()` provides relational access to pinned governed releases.

## Diagnose a failed invocation

```{r failure-inspection, eval=FALSE}
built <- project |> tw_run(stop_on_failure = FALSE)
if (!built$success) {
  tw_dbt_status(built)
  built$stderr
  built$artifact_error
}
```

Each invocation writes separate artifacts under `.tidyweave/runs/<id>` in the dbt
project. Old successful artifacts cannot stand in for a failed invocation. A
nonzero exit code, failed/skipped node or invalid artifact pair makes the result
unsuccessful. There is no automatic CLI retry because it could repeat writes.
Anonymous dbt telemetry is disabled for the child process. Logs can contain SQL,
paths and database diagnostics; store them with the operating project.

The bundled synthetic artifacts work without installing dbt:

```{r artifacts}
artifacts <- system.file("extdata", "dbt-artifacts", package = "tidyweave")
tw_dbt_status(artifacts)
tw_dbt_lineage(artifacts)
```

For details, consult dbt's
[model selection](https://docs.getdbt.com/reference/node-selection/methods),
[run results](https://docs.getdbt.com/reference/artifacts/run-results-json),
[manifest](https://docs.getdbt.com/reference/artifacts/manifest-json) and the
[dbt-duckdb adapter](https://github.com/duckdb/dbt-duckdb).

```{r cleanup, include=FALSE}
unlink(root, recursive = TRUE)
```
