---
title: "dbt workflows and relational models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dbt workflows and relational models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(lakefold)
knitr::opts_chunk$set(eval = requireNamespace("duckdb", quietly = TRUE))
```

This storage guide uses optional DuckDB >= 1.5.5. For a database-free start,
see [composing products](composing-products.html).


Use R to configure and ingest, dbt to build and test SQL models, and `dm` to
inspect and query their relationships. dbt owns its DAG and incremental model
strategies. lakefold invokes its CLI through `processx`, never an internal
Python API. Existing dbt projects can be used without changing their SQL.

## Install an external engine

Install dbt separately from the R package. A tested classic engine combination
is `dbt-core==1.12.5`, `dbt-duckdb==1.10.1` and `duckdb==1.5.5`:

```sh
python -m venv .venv
.venv/bin/python -m pip install dbt-core==1.12.5 dbt-duckdb==1.10.1 duckdb==1.5.5
.venv/bin/dbt --version
```

On Windows use `.venv/Scripts/python.exe` and `.venv/Scripts/dbt.exe`.
These are reproducibility pins, not a claim to be the latest versions. The
wrapper uses standard CLI arguments and JSON artifacts. It disables anonymous
usage telemetry for the child process using both classic and v2 dbt settings. dbt v2 projects can be
supplied explicitly, but their engine, adapter and catalog combinations require
separate integration validation. The generated starter uses dbt-duckdb profiles;
it does not generate a v2 `catalogs.yml` or promise interchangeable engines.

## Generate a local starter

The following chunks execute without dbt; generating files is independent of
running the external engine.

```{r scaffold}
root <- tempfile("lakefold-dbt-guide-")
config <- dl_config(
  dl_catalog_duckdb(file.path(root, "lake.duckdb")),
  dl_storage_local(file.path(root, "data")),
  landing = file.path(root, "landing"), backend = "duckdb"
)
project <- dl_dbt_init(file.path(root, "dbt"), config)
project
list.files(project$path, recursive = TRUE)
```

The starter builds `raw.raw_orders` from a synthetic seed,
`staging.stg_orders`, then `marts.customer_revenue`, with five key/null tests.
It attaches the configured catalog under the same `lake` name used in R. The
schema macro uses exact layer names; use separate catalogs for dev and production.

## Build, test and inspect

The next chunks require your installed dbt executable and are shown without
execution during ordinary vignette builds. The repository integration tests run
these operations against real dbt. Replace `"dbt"` with the executable path if it
is not on `PATH`.

```{r build, eval=FALSE}
# Create catalog folders, then release the local database lock.
lake <- dl_connect(config)
dl_disconnect(lake)

result <- project |> dl_execute(echo = FALSE)
dl_dbt_status(result)
dl_dbt_test(project, select = "customer_revenue", echo = FALSE)
dl_dbt_lineage(result)
```

`select` and `exclude` accept dbt expressions such as `"+customer_revenue"` and
`"tag:reporting"`. `vars` is a named R list; use environment-based credentials in
your profile instead of passing secrets as variables. `full_refresh = TRUE`
requests dbt's full refresh. `timeout` is measured in seconds.

Each invocation writes to its own `.lakefold/runs/<id>` directory. The result
retains the exit code, stdout/stderr, node statuses and parsed manifest. Old
artifacts cannot become the outcome of a new failed invocation. dbt warnings
remain visible; configure warning severity in dbt itself if they should fail CI.

```{r failure-inspection, eval=FALSE}
result <- dl_dbt_build(project, stop_on_failure = FALSE)
if (!result$success) {
  dl_dbt_status(result)
  result$stderr
  result$artifact_error
}
```

A nonzero exit code, failed/skipped node or invalid artifact pair marks the
result unsuccessful. By default `dl_dbt_failed` carries the result in
`condition$result`. Process-start or timeout errors use `dl_dbt_process_error`.
No automatic retry is performed because a retry could repeat writes.

## Understand the returned tables offline

These are explicitly synthetic artifact fixtures bundled for documentation.

```{r artifacts}
artifacts <- system.file("extdata", "dbt-artifacts", package = "lakefold")
dl_dbt_status(artifacts)
dl_dbt_lineage(artifacts)
```

`run_results.json` lists executed nodes, while the manifest describes the whole
project. A selected build may leave some manifest relations absent. Choose those
you want to open by their dbt unique IDs.

## Consume a relational model

```{r model, eval=FALSE}
lake <- dl_connect(config)
model <- dl_dbt_model(
  lake, result,
  tables = c(orders = "model.lakefold_demo.stg_orders",
             customers = "model.lakefold_demo.customer_revenue"),
  primary_keys = list(orders = "order_id", customers = "customer_id"),
  foreign_keys = list(list(
    table = "orders", columns = "customer_id",
    ref_table = "customers", ref_columns = "customer_id"
  ))
)
dm::dm_examine_constraints(model)
dplyr::collect(model$customers)
# Optional diagram: dm::dm_draw(model)
dl_disconnect(lake)
```

The model borrows the connection. `check = TRUE` validates declared keys, which
can be expensive on large tables. SQL dependencies alone cannot prove keys.
The default `database = "lake"` maps manifest schema/table names to the R
attachment, so ensure both R and dbt point to the same catalog.

## Use your own ingested data

`dl_ingest()` returns a pinned release. To make that exact release a dbt source,
read its physical schema/table from `dl_registry(lake, "releases")` and declare
those values in your dbt source YAML. This makes lineage explicit. A subsequent
ingestion creates a new release; updating dbt's source reference is a deliberate
project change. The package does not silently repoint dbt sources or copy dbt
outputs into governed releases.

## DuckLake and deployment

Set `backend = "ducklake"` in the starter config to generate a local DuckLake
attachment. Close R connections before dbt and reopen afterwards. This also
avoids DuckDB version and local catalog locking conflicts between processes.
For S3 or PostgreSQL catalogs, supply an independently verified profile through
`dl_dbt_project(path, profiles_dir, target, executable)`. Keep credentials in
environment variables or a secret provider. Automatic remote profile generation
is not implemented.

A dbt build is not an atomic lakefold publication: earlier models may remain
written when a later model or test fails. `dl_dbt_model()` opens current relations,
whereas `dl_model()` pins immutable governed releases. Do not interchange their
reproducibility guarantees. The registry still requires a single writer.

```{r cleanup}
unlink(root, recursive = TRUE)
```

Artifact contracts follow the official
[run results reference](https://docs.getdbt.com/reference/artifacts/run-results-json)
and [manifest reference](https://docs.getdbt.com/reference/artifacts/manifest-json).
Attachment configuration follows the
[dbt-duckdb adapter](https://github.com/duckdb/dbt-duckdb).
