---
title: "Operation engines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Operation engines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
```

An engine chooses the implementation of one operation. The specification keeps
its predicate or relationship. This guide's executed examples need only the
native implementation; alternatives need the named optional package at execution.

## Quality engines

```{r}
library(tidyweave)
positive <- tw_quality_rule("positive", ~ amount > 0)
native <- positive |> tw_set_engine("native")
alternative <- positive |> tw_set_engine("pointblank")
native
alternative

tw_run_quality(native, data.frame(amount = c(10, -1, NA)))
```

The predicate fails for the negative and missing values. The native engine
records **two failures**. Selecting pointblank stores the choice without loading
the package or reading data. Its dependency and supported operation are checked
before execution. Engine switching supports formula rules; arbitrary R functions
cannot silently become pointblank agent builders.

```{r eval=FALSE}
# install.packages("pointblank")
tw_run_quality(alternative, data.frame(amount = c(10, -1, NA)))
```

## Relationship engines

```{r}
customers <- data.frame(customer_id = 1:2, region = c("North", "South"))
relationship <- tw_lookup_spec(customers, by = "customer_id", name = "customers")
native_lookup <- relationship |> tw_set_engine("native")
dm_lookup <- relationship |> tw_set_engine("dm")

flow <- tw_workflow() |>
  tw_add_product(tw_product("orders")) |>
  tw_add_recipe(tw_recipe() |> tw_step_transform(native_lookup))
tw_collect(tw_trial(flow, data = data.frame(customer_id = c(1L, 2L, 1L))))
```

Both choices describe a checked lookup with unique reference keys. The dm engine
uses dm's relationship checking and requires dm at execution. The native engine
requires no optional relationship package.

## Supported choices

| Specification | Engines | Attach with |
|---|---|---|
| Formula `tw_quality_rule()` | `"native"`, `"pointblank"` | `tw_add_quality()` |
| `tw_lookup_spec()` | `"native"`, `"dm"` | `tw_step_transform()` |

`tw_execution_config()` can supply defaults for an assembled workflow. Explicit
component engine choices take precedence. These are operation engines; storage
is chosen separately with `tw_set_target()` or `tw_publish(to = ...)`.

See [Quality rules](quality-gates.html) for severity and tolerances,
[Integrations](optional-integrations.html) for storage capabilities, and
[Write an adapter](extending-tidyweave.html) for custom implementations.
