---
title: "Everyday workflows: publish, correct and explain"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Everyday workflows: publish, correct and explain}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
library(tidyweave)
```

You have a table, another delivery next month, and a report that must stay
reproducible. Follow one path: define the work, try it, publish a checked result,
then use the same definition for a correction. Only the publication and reporting
sections need optional DuckDB >= 1.5.5.

## Define the work once

The synthetic orders use one row per order. `id` is the business key; an amount
must be non-negative. Those are explicit business decisions.

```{r define-orders}
delivery <- data.frame(id = c(1L, 2L), amount = c(10, 20))
orders <- product("orders", delivery) |>
  dplyr::mutate(amount = round(amount, 2)) |>
  add_contract(contract("orders.schema",
    columns = c(id = "integer", amount = "numeric"), key = "id")) |>
  add_quality(~ amount >= 0)
```

The product stores instructions, not a database connection. Ordinary dplyr verbs
are deferred until execution. A contract describes the expected table; a quality
rule checks its contents. Neither silently repairs the input.

## Try it and explore an ordinary table

```{r try-orders}
preview <- orders |> trial()
collect(preview)
collect(preview) |> dplyr::summarise(total = sum(amount))
```

This needs no storage, metric registration or approval. `trial()` keeps the result in memory and disables configured framework writers
throughout the dependency graph. Use ordinary R analysis while deciding what matters.
`explain(orders)` shows the plan without reading a source.

## Understand a failed delivery

```{r failed-delivery}
bad_delivery <- delivery
bad_delivery$amount[1] <- -10
failed <- orders |> trial(data = bad_delivery, stop_on_failure = FALSE)
status(failed)
quality_report(failed)
quality_rows(failed, ~ amount >= 0)
```

The same definition receives new data. `stop_on_failure = FALSE` returns the
failed result for inspection; it does not bypass checks. `quality_report()`
returns a diagnostic table without needing a file path. Use `quality()` for the
underlying checks and `lineage()` for recorded input evidence. A failed gate
does not write a new accepted release.

## Publish a result you want to keep

```{r storage-availability, include=FALSE}
knitr::opts_chunk$set(eval = requireNamespace("duckdb", quietly = TRUE))
```

```{r publish-orders}
root <- tempfile("tidyweave-everyday-")
orders <- orders |> set_target(root)
first <- orders |> publish()
collect(first)
```

The destination is now retained by the same product definition. It avoids
repeating the destination when deliveries arrive. `publish()` writes checked data
to a local lake. It does not publish on the internet. The framework opens and
closes the connection it owns.

For a workflow that needs optional engines, construct the product with
`execution = execution_config(quality = "pointblank", relationships = "dm",
to = root)`. This ordinary R configuration is retained without opening a
connection. Ordinary predicates and lookup syntax stay the same. Explicit choices on individual steps win. Destination
and layer defaults apply only to the root product, not to every dependency.

A configured `run()` may write to its configured target. For an in-memory trial,
use `trial(orders)`, even after storing a destination. `publish()` makes the intention
to persist visible at the call site.

## Accept a corrected complete delivery

```{r correct-orders}
corrected <- data.frame(id = c(1L, 3L), amount = c(12, 30))
second <- orders |> publish(data = corrected)
collect(second)
collect(first)
```

The new table totals 42. The first result still totals 30 because it pins its
original release. The product definition and its checks have not been rewritten.

These deliveries replace the **complete table**. Order 2 is absent from the new
snapshot and is therefore removed; order 3 is added. If your source sends only
changed rows, do not treat that delta as a complete snapshot. Complete-month
replacement requires an explicit partition policy; see the
[monthly reporting guide](getting-started.html). Keys, partition meaning and
stock-versus-flow calculations cannot be safely inferred from column names.

For several sources, use stable names with `sources = list(name = new_data)`.
A nested product's ID selects its input while preserving its transformations
and gates. Unknown or ambiguous names fail before execution. Already published
inputs remain pinned until explicitly replaced. The
[insurance guide](relational-insurance.html) shows this across checked lookups
and a separate dbt build.

## See what changed

Compare the two exact results directly:

```{r compare-orders}
difference <- compare(first, second)
difference$numeric_summary
difference$changed
```

The contract supplies the key. Result comparison pins both selected releases
and opens and closes its own connection. Lake-based comparison remains available. Order 1 changed from
10 to 12, order 2 was removed, and order 3 added 30. The overall increase is 12.
Counts run in the database; row previews are limited to 100 by default.

## Define related calculations together

```{r exploratory-metrics}
exploratory <- metric_set("orders",
  total = sum(amount),
  count = dplyr::n())
preview <- orders |> trial() |> measure(metrics = exploratory)
collect(preview)
quality(preview)
```

These exploratory calculations require neither a code version nor approval and
do not register unapproved definitions. Trial measurements retain their execution and formula, and `quality(preview)`
returns the trial checks. They cannot be issued as a report. The same definitions
work on a publication result, which pins its exact source release.

When the definitions are ready for an issued report, explicitly identify the
reviewed version and approval:

```{r approved-metrics}
metrics <- metric_set("orders",
  total = sum(amount),
  count = dplyr::n(),
  units = c(total = "EUR", count = "orders"),
  code_version = "orders-analysis-v1", approved = TRUE)
values <- second |> measure(metrics = metrics)
collect(values)
```

Use shared `dimensions`, `time_column` and `time_behavior` for related metrics.
Stock measures require a single date; flows may aggregate periods explicitly.
Missing inputs are rejected by default, even with `na.rm = TRUE`; change the
missing-value policy only when your formula intentionally handles them.

Approval is your declared business decision, not an external approval system.
Change `code_version` when external values, custom code or dependencies change.
Formula text is fingerprinted, but arbitrary closure environments are not.

## Issue a report and reopen it later

```{r issue-report}
report <- values |> report_release("orders-report-v1",
  code_version = "reporting-v1")
saved <- report_read(root, "orders-report-v1", values_only = TRUE)
saved
```

The destination is inferred from the retained source when all measurements
belong to the same lake. Supply `to` explicitly when a destination cannot be
inferred. The original measurement set preserves evidence; its collected table
is for analysis and presentation. Exploratory unapproved or unversioned metrics
cannot be frozen into an issued report.

A report stores values and their input release, calculation definition and
parameters. `report_read()` never recalculates. Changed inputs or formulas need a
new report ID; retrying identical content is safe.

```{r preserve-report}
later <- orders |> publish(data = data.frame(id = 1L, amount = 99))
preserved <- report_read(root, "orders-report-v1", values_only = TRUE)
stopifnot(identical(saved, preserved))
```

The current table now totals 99. The issued report still contains 42 and two
orders, and `collect(second)` still returns its original corrected data.

## Run a correction through its dependent steps

Define the repeated process once with ordinary named functions. Arguments name
inputs or other steps, making the dependency graph explicit. The report remains
an explicit decision after the workflow succeeds.

```{r monthly-workflow}
monthly <- workflow(
  accepted = function(delivery) orders |> publish(data = delivery),
  measures = function(accepted) measure(accepted, metrics = metrics),
  inputs = list(delivery = corrected), code_version = "monthly-v1")
first_run <- run(monthly)
corrected_run <- run(monthly,
  inputs = list(delivery = data.frame(id = 1L, amount = 45)),
  previous = first_run)
status(corrected_run)
collect(corrected_run$results$measures)
```

Only changed branches rerun. Successful independent branches can be reused.
On failure, `run(..., stop_on_failure = FALSE)` returns step status and retained
conditions in `$errors`; dependent steps are skipped. Supply that result as
`previous` for an explicit retry after fixing the cause. Already committed steps
are not rolled back. Change `code_version` when code or captured settings change;
use `refresh = "accepted"` when an external source changes without a new input.

## Add infrastructure only when needed

Use [checked lookups](composing-products.html) for enrichment,
[optional integrations](optional-integrations.html) for existing clients and
specialist engines, and the [layered guide](layered-data-stack.html) when RAW
receipt evidence and dbt SQL models become useful. `ingest()` belongs to that
receipt workflow; `write_data()` is a low-level lake operation.

For exceptional writer recovery, inspect `interrupted()` and follow the
[operations guide](https://github.com/JanWein/tidyweave/blob/main/docs/OPERATIONS.md).
Recovery is not an ordinary delivery step.

Lake writes require one coordinated writer and materialize a full candidate.
Multi-product publication and dbt execution are not one distributed transaction.
Remote storage, backup restoration and workload sizing need verification in the
actual operating environment.

```{r verify-and-cleanup, include=FALSE}
stopifnot(sum(collect(first)$amount) == 30,
  sum(collect(second)$amount) == 42,
  collect(values)$value[collect(values)$.metric == "total"] == 42,
  difference$counts[["changed"]] == 1,
  difference$numeric_summary$difference == 12)
unlink(root, recursive = TRUE)
```
