---
title: "Repeat a workflow after a correction"
description: "Connect dependent steps and rerun them with a changed delivery."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Repeat a workflow after a correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
  eval = requireNamespace("duckdb", quietly = TRUE))
```

**Use this when:** several dependent steps must repeat after a delivery changes.
**Before you start:** complete [Save a report](save-report.html). This guide needs
optional DuckDB >= 1.5.5 and creates its own synthetic data and temporary storage.

For a single product, `publish(orders, data = corrected)` is enough. Add a
workflow when you need to connect several steps and inspect which ones ran.

## Define the product and calculations

```{r definition}
library(tidyweave)
folder <- tempfile("repeat-workflow-")
delivery <- data.frame(id = 1:3, amount = c(25, 75, 50))
orders <- product("orders", delivery) |>
  add_quality(~ amount >= 0) |>
  set_target(folder)
metrics <- metric_set("orders", total = sum(amount, na.rm = TRUE),
  approved = TRUE, code_version = "orders-v1")
```

## Name the dependent steps

```{r workflow}
monthly <- workflow(
  accepted = function(delivery) orders |> publish(data = delivery),
  measures = function(accepted) measure(accepted, metrics = metrics),
  inputs = list(delivery = delivery), code_version = "monthly-v1")
first <- run(monthly)
status(first)
collect(first$results$measures)
```

Each argument names an input or another step. `measures` receives the successful
result of `accepted`. `run()` executes this complete configuration, including its
publication. The total is 150.

## Apply a correction

```{r correction}
corrected <- data.frame(id = 1:3, amount = c(30, 75, 50))
second <- run(monthly, inputs = list(delivery = corrected), previous = first)
status(second)
collect(second$results$measures)
```

The total is now 155. Passing `previous` lets the workflow reuse successful
independent branches. Here both steps depend on the changed delivery, so both
must run. The first result still identifies the original saved data.

## Inspect a failure before retrying

```{r failure}
bad <- corrected
bad$amount[1] <- -30
failed <- run(monthly, inputs = list(delivery = bad), previous = second,
  stop_on_failure = FALSE)
status(failed)
```

The dependent calculation cannot proceed after publication fails. The workflow
retains conditions in `$errors`. Fix the delivery before an explicit retry:

```{r retry}
retried <- run(monthly, inputs = list(delivery = corrected), previous = failed)
collect(retried$results$measures)
```

Already committed steps are not rolled back. A workflow is not one transaction
across databases or products. Change `code_version` when code or captured
settings change. Use `refresh = "accepted"` when an external source changes
without a new input value.

## Issue the report deliberately

```{r report}
retried$results$measures |>
  report_release("orders-report-v1", code_version = "report-v1")
report_read(folder, "orders-report-v1", values_only = TRUE)
```

The report is an explicit decision after the workflow succeeds. Keeping it
outside the automatic steps avoids issuing a report for every trial delivery.
For a larger flow with receipts and dbt, see [insurance reporting](relational-insurance.html).

```{r verify-and-cleanup, include=FALSE}
stopifnot(collect(first$results$measures)$value == 150,
  collect(second$results$measures)$value == 155,
  collect(retried$results$measures)$value == 155,
  report_read(folder, "orders-report-v1", values_only = TRUE)$value == 155)
unlink(folder, recursive = TRUE)
```

[Back to Learn](learn.html) · [Troubleshooting](troubleshooting.html)
