---
title: "Everyday workflows: import, compare, analyze and recover"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Everyday workflows: import, compare, analyze and recover}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
```

## One small workflow, several useful questions

You already have an R table and want to answer three questions: **what is
current, what changed, and what did the earlier report actually use?** You can
start without a pipeline, owner, custom reader or external service.

This example uses synthetic orders. The same pattern works for laboratory
batches, monthly finance data, inventory snapshots or API extracts. A business
key identifies a record; a partition identifies a complete group to replace.
Choose those meanings for your own domain.

```{r open-and-write}
library(lakefold)
root <- tempfile("lakefold-everyday-")
lake <- dl_open(root)
orders <- data.frame(id = c(1L, 2L), amount = c(10, 20))
first <- dl_write(lake, orders, "orders")
```

An automatic structural schema allows integers and decimals in numeric columns.
It does not invent a unique key or reject missing values. Add `contract` when
you know the business requirements. Column names such as `Reserve amount` are
supported; use backticks or `.data[["Reserve amount"]]` in expressions.

## See exactly what changed

A corrected delivery changes order 1, removes order 2, and introduces order 3:

```{r compare-releases}
corrected <- data.frame(id = c(1L, 3L), amount = c(12, 30))
second <- dl_write(lake, corrected, "orders")
difference <- dl_compare(lake, "orders", key = "id")
difference
difference$numeric_summary
difference$changed
```

| Record | Earlier amount | Current amount | Meaning |
|---|---:|---:|---|
| 1 | 10 | 12 | Changed |
| 2 | 20 | Absent | Removed |
| 3 | Absent | 30 | Added |
| Total | 30 | 42 | Increase of 12 |

The total change includes additions and removals. Missing numeric values are
excluded from totals and counted separately. Missing values compare equal when
deciding whether a matched row changed. Row order does not matter.

By default the comparison uses the latest two releases, runs counts in the
database and collects at most 100 rows per preview. Set `from` and `to` to pin
another pair, or `limit = Inf` when collecting every changed row is appropriate.
When your published contract declares a key, omit `key`. Nonunique or missing
keys produce an error instead of an ambiguous comparison.

A structural change is shown in `$schema`. Added, removed or incompatible
column types mark matched records as changed; compatible integer/decimal
columns compare by value.

## Keep complete batches or months together

Whole-table replacement is the default. Add one optional argument when each
delivery contains a complete partition:

```{r partition-example}
august <- data.frame(id = 1L, month = as.Date("2026-08-31"), amount = 100)
september <- data.frame(id = 1L, month = as.Date("2026-09-30"), amount = 110)
dl_write(lake, august, "monthly_orders", partition_by = "month")
dl_write(lake, september, "monthly_orders", partition_by = "month")
august$amount <- 105
dl_write(lake, august, "monthly_orders", partition_by = "month")
dl_read(lake, "monthly_orders") |> dplyr::arrange(month)
```

The current table contains August 105 and September 110. A correction for
August preserves September. Every supplied month replaces the entire month;
send a complete month rather than individual changed records. Use a batch ID,
region/date combination or another declared partition where that meaning fits.
The resulting full table is validated before publication.

Delivery monitoring checks the current date partitions. For a multi-date full
replacement, pass `date_column = "month"` to `dl_check_delivery()`. If no date
column can be identified, the recorded `business_date` is used. Historical
deliveries removed from the current table do not satisfy a current expectation.

## Plug in an existing source client

A zero-argument function can provide data from a database client, API or other
R package. It must return a data frame:

```{r source-function}
fetch_orders <- function() {
  # Replace this body with your existing authenticated client call.
  data.frame(id = 1:2, amount = c(25, 75))
}
dl_write(lake, fetch_orders, "api_orders")
```

The function runs once per call; its returned table is archived as RDS. A
changed response produces a new input fingerprint. This records the received
R object. Use a file source when the exact original workbook or response bytes
must be retained. Acquisition errors happen before a pipeline run is created.
For very large inputs, this callback is not a streaming interface.

## Calculate once, preserve the explanation

A metric needs its identity, input asset and formula. Business descriptions and
units are optional. A governed calculation still needs an explicit approval
state and a code version:

```{r metric-and-report}
total <- dl_metric("orders.total", "orders", sum(amount, na.rm = TRUE),
  approved = TRUE, code_version = "analysis-v1")
values <- dl_measure(lake, total)
report <- dl_report_release(lake, "orders-report-v1", list(total = values),
  code_version = "reporting-v1")
dl_report_read(lake, "orders-report-v1", values_only = TRUE)
```

The saved result is 42, together with its formula identity, input release,
parameters and calculation time. `approved = TRUE` is your declared approval;
it is not a separate approval system. `code_version` must change when external
values, custom code or dependencies change. Full formula text is fingerprinted,
while arbitrary closure environments are not.

With no time column a metric defaults to flow. Supplying `time_column` defaults
to stock and requires exactly one selected reporting date. A stock metric is
appropriate for balances; a flow may sum transactions over multiple dates.

Missing values are rejected in metric input columns by default, even when the
formula uses `na.rm = TRUE`. Set `na_policy = "expression"` only when the
formula intentionally handles them. Static `.data$amount` and
`.data[["amount"]]` are checked. For unresolved dynamic selection or a custom
function, declare `input_columns`. Custom functions without a declaration
check every input column.

Repeat calculation and saving are safe when the report content is identical:

```{r repeat-report}
again <- dl_measure(lake, total)
saved_again <- dl_report_release(lake, "orders-report-v1", list(total = again),
  code_version = "reporting-v1")
saved_again$id
```

Only the calculation time is ignored when comparing a retry. The first saved
timestamps remain. Changed input releases, metric definitions, parameters,
report code or values require a new report ID. `dl_report_read()` itself never
recalculates a metric and can reopen reports created before 0.6.0. JSON readback
represents date values as ISO strings.

## Analyze with a read-only connection

Close the writer and reopen the existing folder:

```{r read-only}
dl_close(lake)
lake <- dl_open(root, read_only = TRUE)
dl_measure(lake, total)
dl_report_read(lake, "orders-report-v1", values_only = TRUE)
```

The persistent database is attached read-only. No schema migration runs, and
measurements default to `record = FALSE`. Their returned manifests still pin
the release and include the complete metric definition. New transient metrics
are allowed, but reusing a registered ID/version with different content fails.
You can also use `record = FALSE` on a writable connection for exploratory work.
The backend's normal connection locking and storage access rules still apply.

## Diagnose and recover only when needed

For a rule that fails during evaluation, retain its original local exception:

```{r local-errors}
contract <- dl_contract("diagnostic", columns = c(id = "integer"),
  rules = list(dl_rule("configuration", function(data) stop("Missing rule setting"))))
quality <- dl_validate(data.frame(id = 1L), contract, keep_errors = TRUE)
lapply(dl_quality_errors(quality), conditionMessage)
```

The exception remains in memory. Raw exception text is not stored in the
registry, because it may include private values. For pointblank, retain agents
with `keep_agents = TRUE` and use `dl_pointblank_report()` for its native detail.

For a crashed writer, first stop the original process and inspect
`dl_interrupted()`. `dl_recover()` previews explicitly selected runs and staging
slots, refuses known live processes, and leaves releases and quality evidence
intact. Unknown ownership requires `writer_stopped = TRUE` after external
verification. See the [operations guide](https://github.com/JanWein/lakefold/blob/main/docs/OPERATIONS.md)
for the execution and retry procedure. Recovery is an occasional operational
step, not part of an ordinary write.

## Boundaries to plan for

The framework still needs one coordinated writer. It materializes a full
candidate for partition corrections and retains each release table. Multi-table
atomic publication, incremental upserts, distributed locking and automated
retention are separate future work. Run the bundled `benchmark.R` against your
own workload before selecting an operating configuration. Remote S3/PostgreSQL
and backup restoration require verification in that environment.

```{r verify-and-close, include=FALSE}
stopifnot(values$value == 42,
  difference$counts[["changed"]] == 1,
  difference$numeric_summary$difference == 12,
  sum(dl_read(lake, "monthly_orders")$amount) == 215,
  nrow(dl_registry(lake, "reports")) == 1)
dl_close(lake)
unlink(root, recursive = TRUE)
```
