---
title: "Extend lakefold with ordinary S3 methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extend lakefold with ordinary S3 methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(lakefold)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Start with a function. A source callback or transformation often needs no class.
Use a specification when you need reusable configuration, inspection, validation
or several interchangeable implementations. You do not need a plugin registry
or access to a private lakefold function.

## The extension contract

| Interface | Receives | Returns | Responsibility |
|---|---|---|---|
| `dl_read_source(source, ...)` | Source specification | data.frame/tibble | Acquire data; close only owned resources |
| `dl_execute_transform(transform, data, ...)` | Transformation and data frame | data.frame/tibble | Transform using the chosen engine |
| `dl_run_quality(rule, data, ...)` | Rule and ordinary or lazy table | Standard quality tibble | Evaluate checks; preserve failures |
| `dl_write_target(target, data, context, ...)` | Checked frame and run context | Output descriptor list | Write successfully or raise an error |
| `dl_execute_target(target, product, ...)` | Validated product | `dl_run_result` | Optional specialized lifecycle/transaction execution |
| `dl_publish_metadata(catalog, metadata, ...)` | Catalog and descriptive metadata | Ignored | Deliver metadata after successful execution |
| `dl_check_component(x, ...)` | Component specification | `x`, invisibly | Check configuration without execution |
| `dl_inspect(x, ...)` | Component | Safe descriptive list with `type` | Describe configuration, omit connections and rows |

Most storage extensions implement only a writer and preflight. The default
executor already handles sources, transformations, contracts, quality, errors
and run evidence. A specialized executor is justified when the backend needs
different transaction or resource semantics, as the lake compiler does.

Register methods in your extension package with roxygen2 `@export` directives
and import the relevant generics, for example `@importFrom lakefold
dl_write_target`. The methods appear as `S3method(...)` entries in NAMESPACE.
All component methods accept `...` so the interface can evolve. Reject options
that your implementation does not support.

## A complete external file target

This example creates a new directory containing `data.rds`. It never reuses an existing destination. It
provides checked publication, but it does not implement lakefold's release
history, registry, partition replacement or transaction semantics.

```{r target}
rds_target <- function(path) {
  structure(list(path = path), class = "example_rds_target")
}

dl_check_component.example_rds_target <- function(x, ...) {
  if (!is.character(x$path) || length(x$path) != 1L ||
      is.na(x$path) || !nzchar(x$path)) {
    stop("The RDS target needs one non-empty path.")
  }
  if (!dir.exists(dirname(x$path))) stop("Create the destination directory first.")
  if (file.exists(x$path)) stop("Choose a new directory; this adapter never replaces data.")
  invisible(x)
}

dl_inspect.example_rds_target <- function(x, ...) {
  list(type = "new RDS directory", path = x$path)
}

dl_write_target.example_rds_target <- function(target, data, context, ...) {
  # Recheck because availability can change after preflight.
  dl_check_component(target)
  if (!dir.create(target$path, showWarnings = FALSE)) {
    stop("The destination could not be created exclusively.")
  }
  complete <- FALSE
  on.exit(if (!complete) unlink(target$path, recursive = TRUE))
  output <- file.path(target$path, "data.rds")
  saveRDS(data, output)
  complete <- TRUE
  list(path = output, product = context$product, rows = nrow(data))
}
```

Creating a new directory establishes exclusive ownership of this output path.
This example does not promise atomic visibility of partial writes to other
readers. Use an adapter with staging and a storage-native commit protocol when
that guarantee matters.

The following registration is only needed because this article defines methods
interactively. An installed extension package uses its NAMESPACE instead.

```{r register}
registerS3method("dl_check_component", "example_rds_target",
  dl_check_component.example_rds_target, envir = asNamespace("lakefold"))
registerS3method("dl_inspect", "example_rds_target",
  dl_inspect.example_rds_target, envir = asNamespace("lakefold"))
registerS3method("dl_write_target", "example_rds_target",
  dl_write_target.example_rds_target, envir = asNamespace("lakefold"))
```

```{r execute}
path <- tempfile("orders-output-")
product <- dl_product("orders") |>
  dl_add_source(data.frame(id = 1:2, amount = c(10, 20))) |>
  dl_add_quality(~ amount >= 0) |>
  dl_add_target(rds_target(path))

result <- dl_run(product)
stopifnot(result$status == "published")
stopifnot(identical(readRDS(result$outputs$path), as.data.frame(dl_collect(result))))
result$outputs
unlink(path, recursive = TRUE)
```

Replace `rds_target(path)` with a lake target and the higher-level product stays
the same. A PostgreSQL, S3 or external API writer belongs behind the same
interface. Use its existing client, authentication and transaction APIs.

## Quality extensions

Create a rule inheriting from `dl_rule`, add your own class, then implement
`dl_run_quality.your_class()`. A convenient way to produce the standard evidence
is to delegate counts to a native rule:

```{r custom-quality}
# In a real adapter, obtain these counts from your external engine.
dl_run_quality.example_quality <- function(rule, data, ...) {
  native <- dl_rule(rule$name, function(data) {
    dl_quality_counts(sum(data$amount < 0 | is.na(data$amount)), nrow(data))
  }, severity = rule$severity, max_failure = rule$max_failure)
  evidence <- dl_run_quality(native, data)
  evidence$engine <- "example"
  evidence
}
```

The required columns are documented by `dl_run_quality()` and visible from any
native call: `rule`, `status`, `severity`, `n_failed`, `n_total`, `threshold`,
`message`, `engine`, `stage`, `segment`, `details`. Evaluated checks require
valid counts. Empty evidence, unknown states, invalid counts and adapter errors
block publication. An extension must never convert an engine failure into a
pass. Keep row values and raw credential-bearing errors out of these columns.

pointblank already implements advanced assertions and thresholds. Reuse its
existing adapter instead of translating all of its assertions into new classes.
Likewise, dbt owns its project graph and external build lifecycle.

## Metadata and orchestration

A catalog function receives a descriptive metadata list after successful data
execution. Implement an S3 catalog adapter only when configuration and preflight
justify one. This is also the integration point for an external catalog such as
OpenMetadata: use a client compatible with the server's schema and API version.
There is no bundled, production-certified OpenMetadata client.

Catalog delivery and data publication are separate operations. A catalog error
is returned as a warning while the data remain published. Design your catalog
upsert to be idempotent using product and release identity; retries must not
create duplicate catalog entities.

Scheduling is an external concern. A CI job, scheduler or Posit Connect task can
call `dl_run(product)`. Result status, timestamps and quality evidence provide a
consistent way to observe that execution. Remote multi-system transactions and
multi-writer locking are not invented by the extension interface.

## Verification checklist for an adapter

Test configuration without acquisition, both successful and failed reads,
resource ownership, invalid output types, missing dependencies, writer failures,
and metadata behavior. Prove that a failed quality gate never invokes your
writer. Run the same product with a native and your custom implementation and
compare data and quality evidence. State the adapter's atomicity, persistence,
retry and concurrency guarantees explicitly.

The package's `test-extension.R` demonstrates separately registered S3 methods;
`test-target-lake.R` proves substitution between native and governed storage
while retaining original inputs, release history and publication gates.
