---
title: "Write an adapter"
description: "Implement and verify an S3 adapter for a new component."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Write an adapter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

For package developers. Read [Architecture](workflow-design.html) first.
The core extension example is self-contained and needs no external service.

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
```

An ordinary function is usually enough for a source or transformation. Introduce
an adapter when configuration, inspection, resource ownership or a different
implementation needs a reusable boundary. An extension package uses ordinary
S3 methods; it does not modify tidyweave's core or register with a plugin system.

## The small set of extension interfaces

| Generic | Input | Required result |
|---|---|---|
| `tw_read_source()` | Source configuration | An ordinary or supported lazy table |
| `tw_execute_transform()` | Transform and table; a named list for multiple inputs | One ordinary or supported lazy table |
| `tw_run_quality()` | Rule and candidate table | Standard quality evidence |
| `tw_write_target()` | Target, checked table and run context | Output descriptor list, or an error |
| `tw_publish_metadata()` | Catalog and descriptive run metadata | Ignored; raise an error on failure |
| `tw_check_component()` | Component configuration | Component invisibly; no acquisition |
| `tw_inspect()` | Component | Safe descriptive list with a `type` field |
| `tw_capabilities()` | Component | Six standard capability flags, optionally additional fields |

The product executor already coordinates acquisition, transforms, checks and
run evidence. A writer is responsible for its destination's persistence,
atomicity and cleanup. It must not claim guarantees it does not provide.

## A source, a writer and a quality engine

This small example is intentionally local and runnable. The source wrapper
stands in for an external reader. The target writes an RDS file inside a new
directory. Neither requires private tidyweave functions.

```{r definitions}
library(tidyweave)

guide_source <- function(data) structure(list(data = data), class = "guide_source")
tw_read_source.guide_source <- function(source, ...) source$data
tw_check_component.guide_source <- function(x, ...) {
  if (!is.data.frame(x$data)) stop("Supply an R data frame.")
  invisible(x)
}
tw_inspect.guide_source <- function(x, ...) {
  list(type = "example reader", columns = names(x$data))
}
tw_capabilities.guide_source <- function(x, ...) {
  tw_component_capabilities(read = TRUE, write = FALSE, lazy = FALSE,
    transactions = FALSE, partition = FALSE, immutable = FALSE)
}

guide_target <- function(path) structure(list(path = path), class = "guide_target")
tw_check_component.guide_target <- function(x, ...) {
  if (!is.character(x$path) || length(x$path) != 1L ||
      is.na(x$path) || !nzchar(x$path)) stop("Supply one destination path.")
  if (!dir.exists(dirname(x$path))) stop("Create the parent directory first.")
  if (file.exists(x$path)) stop("Choose a new directory for this output.")
  invisible(x)
}
tw_inspect.guide_target <- function(x, ...) list(type = "new RDS directory", path = x$path)
tw_capabilities.guide_target <- function(x, ...) {
  tw_component_capabilities(read = FALSE, write = TRUE, lazy = FALSE,
    transactions = FALSE, partition = FALSE, immutable = FALSE)
}
tw_write_target.guide_target <- function(target, data, context, ...) {
  tw_check_component(target)
  if (!dir.create(target$path, showWarnings = FALSE)) {
    stop("Could not create the new destination.")
  }
  complete <- FALSE
  on.exit(if (!complete) unlink(target$path, recursive = TRUE))
  path <- file.path(target$path, "data.rds")
  saveRDS(as.data.frame(data), path)
  complete <- TRUE
  list(path = path, product = context$product, rows = nrow(data))
}

guide_quality <- function() {
  rule <- tw_quality_rule("nonnegative", ~ amount >= 0)
  class(rule) <- c("guide_quality", class(rule))
  rule
}
tw_run_quality.guide_quality <- function(rule, data, ...) {
  # A real external engine would supply these aggregate counts.
  failed <- sum(is.na(data$amount) | data$amount < 0)
  total <- nrow(data)
  native <- tw_quality_rule(rule$name, function(data) tw_quality_counts(failed, total),
    severity = rule$severity, max_failure = rule$max_failure)
  evidence <- tw_run_quality(native, data)
  evidence$engine <- "example"
  evidence
}
```

The writer owns its newly created directory and removes incomplete output on
failure. It does **not** promise atomic visibility to other readers, immutable
history, partition replacement or distributed locking. A production adapter
should use the existing storage client's commit protocol when those are needed.

The quality adapter preserves the class from `tw_quality_rule()` and returns its
standard evidence shape. Unknown states, empty evidence, errors and malformed
counts cannot authorize publication. Never turn an engine failure into a pass.

## Register methods in the extension package

In an installed package, add roxygen2 `@export` and `@importFrom tidyweave
read_source` (or the relevant generic) above each method. roxygen2 writes its
S3 registration to NAMESPACE. This article instead registers the methods
interactively so the example can run by itself:

```{r register-methods}
methods <- list(
  tw_read_source.guide_source = tw_read_source.guide_source,
  tw_check_component.guide_source = tw_check_component.guide_source,
  tw_inspect.guide_source = tw_inspect.guide_source,
  tw_capabilities.guide_source = tw_capabilities.guide_source,
  tw_check_component.guide_target = tw_check_component.guide_target,
  tw_inspect.guide_target = tw_inspect.guide_target,
  tw_capabilities.guide_target = tw_capabilities.guide_target,
  tw_write_target.guide_target = tw_write_target.guide_target,
  tw_run_quality.guide_quality = tw_run_quality.guide_quality
)
for (name in names(methods)) {
  registerS3method(sub("[.].*$", "", name), sub("^[^.]+[.]", "", name),
    methods[[name]], envir = asNamespace("tidyweave"))
}
```

Keep `...` in extension methods so the interface is consistent. Reject options
that your method does not support. Preflight must not open a remote connection,
execute a query or write a file just to decide whether a definition is valid.

## Prove that substitution leaves the workflow intact

```{r substitution}
guide_workflow <- function(source, target = NULL, check = ~ amount >= 0) {
  tw_product("orders") |>
    tw_add_source(source) |>
    tw_add_transform(function(data) transform(data, amount = amount * 2)) |>
    tw_add_quality(check) |>
    tw_set_target(target)
}
input <- data.frame(id = 1:2, amount = c(10, 20))
path <- tempfile("custom-target-")
native <- tw_run(guide_workflow(input))
extended <- tw_run(guide_workflow(guide_source(input), guide_target(path), guide_quality()))
stopifnot(identical(tw_collect(native), tw_collect(extended)))
stopifnot(identical(readRDS(extended$outputs$path), as.data.frame(tw_collect(native))))

flags <- tw_capabilities(guide_target(tempfile()))
stopifnot(setequal(names(flags), names(tw_component_capabilities())),
  all(vapply(flags, function(x) is.logical(x) && length(x) == 1L, logical(1))))
```

The same function composes both implementations. `TRUE` means a declared
capability is supported, `FALSE` means it is unsupported, and `NA` means it is
unknown or depends on user code. A capability declaration is not evidence that
a remote database supports a particular transaction or SQL translation.

```{r gate}
blocked_path <- tempfile("blocked-target-")
bad <- guide_workflow(guide_source(data.frame(id = 1L, amount = -1)),
  guide_target(blocked_path), guide_quality())
blocked <- tw_run(bad, stop_on_failure = FALSE)
stopifnot(blocked$status == "blocked", !file.exists(blocked_path))
```

The failed gate never invokes the writer. These are useful conformance tests,
alongside invalid inputs, failed reads, failed writes, owned-resource cleanup,
missing dependencies and lazy/materialized behavior.

## Add specialist integrations without replacing them

Use the existing pointblank adapter for pointblank assertions. Let dbt own its
SQL graph and tests. Let targets own dependency caching. Their existing clients
and conventions belong inside adapters, not in dozens of product arguments.

Metadata delivery is another boundary. A callback is sufficient for a small
integration; OpenLineage and OpenMetadata adapters provide HTTP implementations.
Delivery failures are visible without undoing published data. Durable retries
receive freshly supplied adapters rather than serialized credentials.

```{r cleanup}
unlink(path, recursive = TRUE)
method_table <- get(".__S3MethodsTable__.", envir = asNamespace("tidyweave"))
rm(list = names(methods), envir = method_table)
```

The [optional integrations guide](optional-integrations.html) demonstrates built-in
adapters together. The package tests include independent S3 substitution and built-in adapter
checks. Real deployment verification still belongs to the integration's owner:
a local HTTP fixture cannot certify an enterprise catalog or IAM configuration.


[Back to Learn](learn.html) · [Troubleshooting](troubleshooting.html)
