---
title: "1. Build a product"
description: "Prepare a table once and use the same instructions on another delivery."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Build a product}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
```

**Goal:** turn a table into a reusable preparation step. **Needs:** tidyweave;
no storage or optional engine. Start in a fresh R session.

## Start with a delivery

```{r delivery}
library(tidyweave)
delivery <- data.frame(id = 1:3, amount = c(25, 75, 50))
```

There is one row per order. `id` identifies it, and `amount` is its value.
The total is 150. For now, the data is already in R; reading a file can come later.

## Describe the work

```{r product}
orders <- product("orders", delivery) |>
  dplyr::mutate(amount = round(amount, 2))
```

A **product** is the instructions for a named table. This call describes a
rounding step; it has not run the step or written anything yet. Supported dplyr
verbs work on a product in the same order as in your pipe.

## Try it, then use the result

```{r trial}
result <- orders |> trial()
collect(result)
collect(result) |> dplyr::summarise(total = sum(amount))
```

`trial()` executes the instructions without configured publication or catalog
writers. The **result** records that attempt. `collect()` gives you an ordinary
tibble for further analysis. The total is **150**.

## Reuse the instructions

```{r next-delivery}
next_delivery <- data.frame(id = 1:3, amount = c(30, 75, 50))
next_result <- orders |> trial(data = next_delivery)
collect(next_result) |> dplyr::summarise(total = sum(amount))
```

The new total is **155**. You passed new data to the same definition; you did not
copy the preparation code. Neither result has been saved beyond this R session.

## What you learned

`product()` defines the work, `trial()` tries it, and `collect()` returns the
usable table. Before saving results, add the checks that make a delivery acceptable.

```{r verify, include=FALSE}
stopifnot(sum(collect(result)$amount) == 150,
  sum(collect(next_result)$amount) == 155)
```

## Continue

[Previous: Get started](get-started.html) · [Next: Check a delivery](check-delivery.html) · [All five lessons](get-started.html)
