---
title: "Product specifications"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Product specifications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
```

A product specification describes the identity and requirements of an output.
It can be shared independently of its preparation and delivery. This guide
needs only tidyweave. Read [Get started](get-started.html) for the full lifecycle.

## Name the output

```{r}
library(tidyweave)
orders <- tw_product("orders", owner = "Analytics",
  description = "One row per accepted order")
orders
```

Use a stable product name. A code version is useful when a workflow enables
caching; update it whenever code or captured dependencies change.

## Define shape and grain

A contract describes columns, required values and keys. The key expresses row
identity. Checks apply to the prepared candidate, after recipe steps.

```{r}
schema <- tw_contract(
  columns = c(id = "integer", amount = "numeric"), key = "id"
)
orders <- orders |> tw_add_contract(schema)
```

For a contract inferred from a sample, `tw_contract_from()` creates a draft.
Review and confirm it with `tw_contract_confirm()` before using it as a gate.
`tw_contract_update()` creates a revised contract while retaining unchanged
requirements. See the reference for extra-column and empty-table policies.

## Add quality requirements

```{r}
nonnegative <- tw_quality_rule("nonnegative", ~ amount >= 0) |>
  tw_set_engine("native")
orders <- orders |> tw_add_quality(nonnegative)
```

Named rules make failures recognizable in run evidence. Formula checks use the
candidate's columns. Engine selection changes how a supported operation runs;
it does not change the meaning of its predicate. See [Engines](engines.html).

## Assemble and verify

```{r}
flow <- tw_workflow() |>
  tw_add_product(orders) |>
  tw_add_recipe(tw_recipe() |> tw_step_mutate(amount = round(amount, 2)))
result <- tw_trial(flow, data = data.frame(id = 1:2, amount = c(12.345, 20)))
tw_collect(result)
tw_quality(result)
```

The prepared amounts are **12.35** and **20**. A failed blocking requirement
prevents the framework writer from running. Read
[Check a delivery](check-delivery.html) to inspect a rejected attempt.

## Keep responsibilities separate

In a modular workflow, keep transformations in `tw_recipe()`. A product with
embedded transformations cannot be added with `tw_add_product()`. Sources and
destinations can belong to the product or workflow, but cannot be configured in
both places. Supplying `data =` at execution is convenient for a single delivery.

For small scripts, direct `tw_product("orders", delivery)` pipelines with dplyr
verbs remain supported. Relational model products also retain their member-table
relationships; prepare their member tables before assembling the model.

Continue with [Preparation recipes](preparation-recipes.html).
