---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
```

Build a checked table from three independent definitions, then reuse the same
workflow for a new delivery. This introduction needs only tidyweave and its hard
dependencies. Every example on this page runs in a fresh R session.

## Define the product

A product specification names the output and states its requirements. Here,
each order has a unique integer ID and a nonnegative amount.

```{r}
library(tidyweave)
orders <- tw_product("orders") |>
  tw_add_contract(tw_contract(
    columns = c(id = "integer", amount = "numeric"), key = "id"
  )) |>
  tw_add_quality(~ amount >= 0)
orders
```

## Describe preparation

A recipe is an ordered collection of deferred steps. Its expressions use dplyr
data masking: `amount` refers to the column when the recipe is executed.

```{r}
preparation <- tw_recipe() |>
  tw_step_mutate(amount = round(amount, 2))
preparation
```

The product and recipe are ordinary R objects. Neither holds this delivery or
opens a connection. The same recipe can serve several products.

## Assemble the workflow

```{r}
flow <- tw_workflow() |>
  tw_add_product(orders) |>
  tw_add_recipe(preparation)
flow
```

The workflow coordinates the components. An optional source or destination can
be attached later. Keeping data outside the definition makes repeated deliveries
explicit.

## Try a delivery

```{r}
delivery <- data.frame(id = 1:3, amount = c(25, 75, 50))
result <- tw_trial(flow, data = delivery)
tw_collect(result)
tw_status(result)
```

Preparation runs before the output checks. The result records the attempt;
`tw_collect()` returns an ordinary tibble. Its amount totals **150**.
`tw_trial()` disables framework publication and catalog writers, including those
on dependencies. User-supplied functions remain responsible for their own effects.

## Reuse and change components

```{r}
corrected <- data.frame(id = 1:3, amount = c(30, 75, 50))
next_result <- tw_trial(flow, data = corrected)
sum(tw_collect(next_result)$amount)

whole_amounts <- tw_recipe() |> tw_step_mutate(amount = round(amount))
alternate <- flow |> tw_update_recipe(whole_amounts)
tw_extract_recipe(flow)
tw_extract_recipe(alternate)
```

The corrected total is **155**. Updating a slot returns a new workflow; `flow`
still has its original recipe. No delivery has been persisted by these trials.

## Publish when ready

With DuckDB >= 1.5.5 installed, save a checked delivery to a local folder:

```{r eval=FALSE}
saved <- tw_publish(flow, data = corrected, to = "data/orders")
tw_collect(saved)
```

Publication means saving data to a destination. It does not make it public
online. A lake result identifies an immutable release. Use `tw_run()` when you
want to execute the destinations already configured on a definition.

## Choose the next topic

Continue with [recipes](preparation-recipes.html) and
[workflows](modular-workflows.html), then [quality checks](quality-gates.html)
and [reports](save-report.html). [Integrations](optional-integrations.html)
explains storage and optional tools. Function arguments and return values live
in the [reference](../reference/index.html).

```{r verify, include=FALSE}
stopifnot(sum(tw_collect(result)$amount) == 150,
          sum(tw_collect(next_result)$amount) == 155)
```
