---
title: "Products, recipes and workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Products, recipes and workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r failure}
library(dataraft)

orders <- dr_product("orders") |>
  dr_add_contract(c(id = "integer", amount = "numeric")) |>
  dr_add_quality(~ amount >= 0)

flow <- dr_workflow() |>
  dr_add_product(orders) |>
  dr_add_recipe(dr_recipe() |> dr_step_mutate(amount = round(amount, 2)))

bad_delivery <- data.frame(id = 1:3, amount = c(25, -75, 50))
checked <- dr_trial(flow, data = bad_delivery)
checked$status
dr_quality_rows(checked)
```

Correct the delivery and reuse the same workflow:

```{r success}
next_delivery <- data.frame(id = 1:3, amount = c(25, 75, 50))
result <- dr_trial(flow, data = next_delivery)
dr_collect(result)
```

A **product** defines identity, contract and quality requirements. A **recipe**
defines preparation in step order. A **workflow** binds these definitions to
sources and a target. Definitions do no I/O. `dr_trial()` checks without writing;
`dr_run()` executes the configured target. `dr_collect()` retrieves the output.


Use `dr_plan()` to inspect a definition before reading sources. Contract failures
and quality failures are available in the result; `dr_quality_rows()` exposes
retained failing rows. Set a destination with `dr_set_target()` and execute with
`dr_run()` once the trial passes. Quality gates run before the destination writer.

Recipes own preparation in modular workflows. Products added with
`dr_add_product()` must not carry embedded transformations. Recipe steps run in
the order added, followed by output contract checks and quality rules.
