---
title: "Monthly deliveries, corrections and reproducible reports"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Monthly deliveries, corrections and reproducible reports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
  eval = requireNamespace("duckdb", quietly = TRUE))
```

This guide uses optional DuckDB >= 1.5.5 and synthetic data. For a first example
without infrastructure, start with [Compose a product](composing-products.html).

We will publish August's total of **350**, accept a correction to **370**,
reject a duplicate, add September's **390**, and read the original issued
report. The same product grammar handles every delivery.

## 1. Describe one monthly delivery

```{r definition}
library(tidyweave)
root <- tempfile("monthly-reserves-")
august_date <- as.Date("2026-08-31")
august <- data.frame(entity = c("North", "South"),
  date = rep(august_date, 2), amount = c(100, 250))

reserves <- product("reserves", august) |>
  add_contract(contract(
    columns = c(entity = "character", date = "Date", amount = "numeric"),
    key = c("entity", "date")
  )) |>
  add_quality(~ amount >= 0, name = "nonnegative") |>
  set_target(target_lake(root, partition_by = "date"))

first <- run(reserves, business_date = "2026-08-31")
sum(collect(first)$amount)
stopifnot(first$status == "published", sum(collect(first)$amount) == 350)
```

The contract requires one row per entity and date. The rule rejects negative
amounts. The target replaces complete delivered dates and retains other dates.
No owner, catalog service or manually assigned definition version is required.

`first$release_id` identifies a checked, immutable table state. A reporting date
is a value inside the table; a release ID identifies its version. They are
different things.

## 2. Record the report when it is issued

A metric gives the number an explicit meaning. A reserve is a stock at one date,
so adding August and September is not a valid reserve total. The stock setting
requires selection of one reporting date.

```{r issued-report}
total_reserve <- metric("total_reserve", "reserves",
  expr = sum(amount, na.rm = TRUE), time_column = "date", time_behavior = "stock",
  unit = "EUR", approved = TRUE, code_version = "report-v1")
as_reported <- measure(first, total_reserve, at = august_date)
lake <- open_lake(root)
report_release(lake, "august-report-v1",
  results = list(total_reserve = as_reported), code_version = "report-v1")
close_lake(lake)
stopifnot(as_reported$value == 350)
```

`approved = TRUE` records your declared business approval; it does not perform
that review. The report manifest stores the value, metric definition and input
release. It does not render a PDF or presentation. Your existing reporting tools
can use the value.

## 3. Accept a correction without changing the issued report

```{r correction}
corrected <- august
corrected$amount[corrected$entity == "South"] <- 270
reserves <- reserves |>
  add_source(corrected, replace = TRUE)
correction <- run(reserves, business_date = "2026-08-31")
sum(collect(correction)$amount)
sum(collect(first)$amount)
stopifnot(sum(collect(correction)$amount) == 370,
  sum(collect(first)$amount) == 350)
```

`replace = TRUE` replaces the sole primary source in the product definition. The lake
target separately controls how the resulting delivery updates stored data.
The current release now totals 370. Reading the first run still selects its
exact release, not whatever happens to be current.

## 4. Reject a duplicate delivery

```{r duplicate}
bad <- rbind(corrected, corrected[2, ])
blocked <- reserves |>
  add_source(bad, replace = TRUE) |>
  run(stop_on_failure = FALSE)
incidents(blocked)
stopifnot(blocked$status == "blocked")
```

The repeated South row violates the contract key. It never becomes current.
The default raises an error on a failed run; `stop_on_failure = FALSE` lets
this lesson inspect the outcome. Correcting data belongs in preparation or the
upstream source, not inside a quality check that silently deletes a row.

## 5. Add a complete September partition

```{r september}
september <- data.frame(entity = c("North", "South"),
  date = rep(as.Date("2026-09-30"), 2), amount = c(110, 280))
latest <- reserves |>
  add_source(september, replace = TRUE) |>
  run(business_date = "2026-09-30")
collect(latest) |> dplyr::arrange(date, entity)
stopifnot(nrow(collect(latest)) == 4L)
```

Both dates are now in the current table. Each supplied date must contain its
complete replacement: sending only South for September would remove North from
September. This is partition replacement, not a row upsert. The gate validates
the complete candidate, including retained August rows.

## 6. Reuse a pinned release in another product

```{r derived-product}
monthly_totals <- product("monthly_totals", latest) |>
  dplyr::group_by(date) |>
  dplyr::summarise(total = sum(amount, na.rm = TRUE), .groups = "drop")
totals <- collect(run(monthly_totals)) |> dplyr::arrange(date)
totals
stopifnot(identical(totals$total, c(370, 390)))
```

This is the same kind of product as the delivery workflow. Its input happens
to be a checked release rather than a spreadsheet. The successful result supplies its exact release directly. The framework owns
and closes the read connection; no manual release-ID plumbing is needed.

## 7. Read the issued report, without recalculating it

```{r read-report}
lake <- open_lake(root, read_only = TRUE)
issued <- report_read(lake, "august-report-v1", values_only = TRUE)
current <- measure(lake, total_reserve, at = august_date)
issued$total_reserve$value
current$value
stopifnot(issued$total_reserve$value == 350, current$value == 370)
close_lake(lake)
```

| Event | Current data | Issued August report |
|---|---|---|
| First August delivery | August 350 | 350 |
| August correction | August 370 | 350 |
| Duplicate delivery | August 370, unchanged | 350 |
| September delivery | August 370; September 390 | 350 |

The point is not to make 350 and 370 agree. It is to preserve which question
each number answers, with the input and calculation needed to explain it.

## Use your own file next

Replace only the source. A custom Excel reader can convert Excel timestamps to
R dates so they match the contract:

```{r real-file, eval=FALSE, purl=FALSE}
read_reserves <- function(path) {
  readxl::read_excel(path, sheet = "Reserves") |>
    dplyr::mutate(date = as.Date(date))
}
reserves |>
  add_source("incoming/reserves.xlsx",
    reader = read_reserves, replace = TRUE) |>
  run(evidence = "runs")
```

Passing a file path lets the lake archive the original workbook. Passing an R
table archives that table as RDS instead. Use one coordinated writer per lake.
For shared infrastructure, configure a target once rather than adding backend
arguments to every step. See [workflow design](workflow-design.html) and the
[operations guide](https://github.com/JanWein/tidyweave/blob/main/docs/OPERATIONS.md).

```{r cleanup}
unlink(root, recursive = TRUE)
```
