---
title: "Getting started: open, write and read"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started: open, write and read}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(lakefold)
```

Start with data already in R. Everything in this guide runs locally, without
credentials, extension downloads or optional integrations.

## Open, write and read

The example uses a temporary folder. In your project, use `dl_open("my-lake")`;
the folder and its backend are remembered across sessions.

```{r minimal}
root <- tempfile("lakefold-guide-")
lake <- dl_open(root)
orders <- data.frame(id = 1:3, amount = c(25, 75, 50))
first <- dl_write(lake, orders)
dl_read(lake, "orders")
```

The variable name supplies the asset name. For expressions, supply it yourself:
`dl_write(lake, data.frame(id = 1:3), "orders")`. `dl_write()` returns a run
result; `dl_read()` returns a tibble. No configuration, source specification,
contract or manual code version is required for this workflow.

## What happens automatically

Each write preserves a delivery snapshot, checks a candidate and publishes an
immutable release. The first successful delivery establishes the column names
and types for future writes. Missing values are allowed. Empty tables and
schema changes block publication. No key, owner, business rule or freshness
deadline is inferred.

```{r defaults}
dl_write(lake, orders) # unchanged built-in workflow: cached
bad <- data.frame(id = c("a", "b"), amount = c(25, 75))
blocked <- dl_write(lake, bad, "orders", stop_on_failure = FALSE)
blocked$status
dl_quality(lake, run_id = blocked$run_id)
dl_read(lake, "orders") # the successful release is still available
```

Normally a blocked write throws an error after recording its quality results.
`stop_on_failure = FALSE` returns the result for interactive handling. Automatic
schema checks protect structure; they cannot establish business correctness.
Integer columns accept integers; numeric columns also accept integer inputs.
Factors, list columns and unsupported classes require explicit conversion when
establishing a schema. Valid column names start with a letter and contain only
letters, digits and underscores; names are never silently repaired.

## Read files directly

CSV, TSV and RDS files have built-in readers. Original file bytes are archived
before parsing. RDS preserves R dates and other supported column types. CSV/TSV
use base R parsing; use a custom reader when dates, locale or column types need
explicit handling.

```{r file}
path <- file.path(root, "deliveries.csv")
utils::write.csv(orders, path, row.names = FALSE)
dl_write(lake, path)
dl_read(lake, "deliveries")
```

For other formats, pass `reader`, for example
`reader = function(path) readxl::read_excel(path)`. Install that reader's package
separately. Readers must return a data frame and must not modify the archived
file. The first file can be parsed twice while its schema is established.
A parsing or type-inference failure at this preparation stage raises an error
before a run is created; the archived original is retained.

## Add a contract when you need one

A contract requires an identifier and column types. Metadata and additional
rules are optional. Explicit contracts require non-missing declared columns by
default; set `required = character()` when missing values are acceptable.

```{r contract}
contract <- dl_contract("orders.checked",
  columns = c(id = "integer", amount = "numeric"), key = "id")
dl_write(lake, orders, contract = contract)
```

Continue supplying this contract on later writes. Omitting it raises an error
instead of silently removing its rules. To change a registered contract, change
its `version`; `dl_contract_diff()` helps review the change. To intentionally
change the automatic schema, supply an explicit contract for the new schema.

Use `dl_contract_from(orders, "orders.checked")` to draft a larger contract.
Review and confirm that draft with `dl_contract_confirm()` before use. This is
separate from the automatic structural baseline used by the minimal workflow.

Keys, R rules, pointblank rules, input checks, ownership and freshness monitoring
can be added independently. See `vignette("quality-gates")` for pointblank.

## Query large data or an exact historical release

```{r reads}
dl_read(lake, "orders", lazy = TRUE) |>
  dplyr::filter(amount >= 50) |>
  dplyr::collect()

dl_read(lake, "orders", release = first$release_id)
dl_releases(lake, "orders")
```

Ordinary `dl_read()` collects all rows into R. Lazy reads keep the work in the
database until collection and require an open connection.

`dl_write()` means that a successful write becomes the current data. Writing an
older payload again creates a new release if a different release is current.
The advanced `dl_run()` keeps its historical-cache policy for idempotent job
retries; use `cache = "current"` to select the write policy explicitly.

## Reopen later

```{r reopen}
dl_close(lake)
lake <- dl_open(root)
dl_read(lake, "orders")
dl_close(lake)
```

`dl_open()` creates a local DuckDB lake by default. For a new DuckLake folder,
choose `backend = "ducklake"` once. Its extension must be available. Reopening
remembers that choice and rejects accidental backend changes. Existing lakes
created with `dl_setup()` still use their original `dl_config()`; the simple
opener does not guess their layout or migrate them.

## Optional complexity

| When you need it | Add |
|---|---|
| Custom storage, schemas or credentials | `dl_config()` and `dl_connect()` |
| Checks before Raw is written | `input_contract` |
| Owner, row meaning, units or deadlines | Contract metadata |
| Advanced quality reports | pointblank and quality-report functions |
| Transformations or partition replacement | `dl_pipeline()` and explicit steps |
| Reusable derived tables | `dl_product()` |
| SQL dependency graphs | The dbt integration |
| Relationships between tables | dm models |
| Reproducible business measures | Metrics and report manifests |

The simple path chooses definition versions automatically. Built-in readers and
structural rules can reuse the current release. Custom readers and rules run
again by default, because function bodies cannot identify every captured value
or external dependency. Supply `code_version` to enable reuse and change it
when any relevant behavior changes. Set `cache = FALSE` to always check again.

The same release registry, diagnostics and final gate support both simple and
explicit workflows. The framework requires one coordinated writer; optional
components do not alter that boundary.

```{r cleanup, include=FALSE}
unlink(root, recursive = TRUE)
```
