---
title: "Work through a reporting month"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Work through a reporting month}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
  eval = requireNamespace("duckdb", quietly = TRUE))
```

**Goal:** combine specifications, checks and publication in a monthly workflow.
**Needs:** tidyweave and optional DuckDB >= 1.5.5. The example is self-contained.
Read [Save a report](save-report.html) first if report issuance is new to you.

This adds two decisions: a delivery replaces one complete month, and a reserve
is measured at one date rather than added across dates.

We will publish August's total of **350**, accept a correction to **370**,
reject a duplicate, add September's **390**, and read the original issued
report. The same product grammar handles every delivery.

## 1. Describe one monthly delivery

```{r definition}
library(tidyweave)
root <- tempfile("monthly-reserves-")
august_date <- as.Date("2026-08-31")
august <- data.frame(entity = c("North", "South"),
  date = rep(august_date, 2), amount = c(100, 250))

reserves <- tw_product("reserves", august) |>
  tw_add_contract(tw_contract(
    columns = c(entity = "character", date = "Date", amount = "numeric"),
    key = c("entity", "date")
  )) |>
  tw_add_quality(~ amount >= 0, name = "nonnegative") |>
  tw_set_target(tw_target_lake(root, partition_by = "date"))

first <- tw_publish(reserves, business_date = "2026-08-31")
sum(tw_collect(first)$amount)
stopifnot(first$status == "published", sum(tw_collect(first)$amount) == 350)
```

The contract requires one row per entity and date. The rule rejects negative
amounts. The target replaces complete delivered dates and retains other dates.
No owner, catalog service or manually assigned definition version is required.

`first` retains the exact saved table. Its reporting date is a value inside
the data; a correction for that date creates another saved version.

## 2. Record the report when it is issued

A metric gives the number an explicit meaning. A reserve is a stock at one date,
so adding August and September is not a valid reserve total. The stock setting
requires selection of one reporting date.

```{r issued-report}
metrics <- tw_metric_set("reserves", total_reserve = sum(amount, na.rm = TRUE),
  time_column = "date", time_behavior = "stock", units = "EUR",
  approved = TRUE, code_version = "reserves-v1")
as_reported <- tw_measure(first, metrics = metrics, at = august_date)
as_reported |> tw_report_release("august-report-v1", code_version = "report-v1")
stopifnot(tw_collect(as_reported)$value == 350)
```

`approved = TRUE` records your declared business approval; it does not perform
that review. The report manifest stores the value, metric definition and input
release. It does not render a PDF or presentation. Your existing reporting tools
can use the value.

## 3. Accept a correction without changing the issued report

```{r correction}
corrected <- august
corrected$amount[corrected$entity == "South"] <- 270
correction <- tw_publish(reserves, data = corrected, business_date = "2026-08-31")
sum(tw_collect(correction)$amount)
sum(tw_collect(first)$amount)
stopifnot(sum(tw_collect(correction)$amount) == 370,
  sum(tw_collect(first)$amount) == 350)
```

`data = corrected` submits the replacement delivery to the existing definition.
The target replaces only the complete dates present in that delivery.
The current release now totals 370. Reading the first run still selects its
exact release, not whatever happens to be current.

## 4. Reject a duplicate delivery

```{r duplicate}
bad <- rbind(corrected, corrected[2, ])
blocked <- tw_publish(reserves, data = bad, stop_on_failure = FALSE)
tw_quality_report(blocked)
stopifnot(blocked$status == "blocked")
```

The repeated South row violates the contract key. It never becomes current.
The default raises an error on a failed run; `stop_on_failure = FALSE` lets
this guide inspect the outcome. Correcting data belongs in preparation or the
upstream source, not inside a quality check that silently deletes a row.

## 5. Add a complete September partition

```{r september}
september <- data.frame(entity = c("North", "South"),
  date = rep(as.Date("2026-09-30"), 2), amount = c(110, 280))
latest <- tw_publish(reserves, data = september, business_date = "2026-09-30")
tw_collect(latest) |> dplyr::arrange(date, entity)
stopifnot(nrow(tw_collect(latest)) == 4L)
```

Both dates are now in the current table. Each supplied date must contain its
complete replacement: sending only South for September would remove North from
September. This is partition replacement, not a row upsert. The gate validates
the complete candidate, including retained August rows.

## 6. Reuse a pinned release in another product

```{r derived-product}
monthly_totals <- tw_product("monthly_totals", latest) |>
  dplyr::group_by(date) |>
  dplyr::summarise(total = sum(amount, na.rm = TRUE), .groups = "drop")
totals <- tw_collect(tw_trial(monthly_totals)) |> dplyr::arrange(date)
totals
stopifnot(identical(totals$total, c(370, 390)))
```

This is the same kind of product as the delivery workflow. Its input happens
to be a checked release rather than a spreadsheet. The successful result supplies its exact release directly. The framework owns
and closes the read connection; no manual release-ID plumbing is needed.

## 7. Read the issued report, without recalculating it

```{r read-report}
issued <- tw_report_read(root, "august-report-v1", values_only = TRUE)
current <- tw_measure(latest, metrics = metrics, at = august_date)
issued
tw_collect(current)
stopifnot(issued$value == 350, tw_collect(current)$value == 370)
```

| Event | Current data | Issued August report |
|---|---|---|
| First August delivery | August 350 | 350 |
| August correction | August 370 | 350 |
| Duplicate delivery | August 370, unchanged | 350 |
| September delivery | August 370; September 390 | 350 |

The point is not to make 350 and 370 agree. It is to preserve which question
each number answers, with the input and calculation needed to explain it.

## Use your own file next

Replace only the source. A custom Excel reader can convert Excel timestamps to
R dates so they match the contract:

```{r real-file, eval=FALSE, purl=FALSE}
read_reserves <- function(path) {
  readxl::read_excel(path, sheet = "Reserves") |>
    dplyr::mutate(date = as.Date(date))
}
reserves |>
  tw_add_source("incoming/reserves.xlsx",
    reader = read_reserves, replace = TRUE) |>
  tw_publish(evidence = "runs")
```

Passing a file path lets the lake archive the original workbook. Passing an R
table archives that table as RDS instead. Use one coordinated writer per lake.
For shared infrastructure, configure a target once rather than adding backend
arguments to every step. See [workflow design](workflow-design.html) and the
[operations guide](https://github.com/JanWein/tidyweave/blob/main/docs/OPERATIONS.md).

```{r cleanup}
unlink(root, recursive = TRUE)
```

## Continue

You can now process deliveries, reject a bad input, accept a correction and
reopen the original report. Continue with [three related deliveries](three-deliveries.html)
for named inputs, grouped reporting and reference corrections, or choose a task
in [Learn](learn.html).

[Previous: Save a report](save-report.html) · [Get started](get-started.html)
