---
title: "Getting started: from file to a governed release"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started: from file to a governed release}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(lakefold)
```

lakefold connects ordinary R ingestion, quality contracts, DuckDB/DuckLake,
dbt and relational analysis. Start locally with DuckDB. This example executes
without network access, database credentials or a dbt installation.

## Configure storage

A specification can be printed and saved before it touches a database. Paths
are resolved when the specification is created. A config stores environment
variable names for credentials, never resolved secrets.

```{r configuration}
root <- tempfile("lakefold-guide-")
config <- dl_config(
  dl_catalog_duckdb(file.path(root, "lake.duckdb")),
  dl_storage_local(file.path(root, "data")),
  landing = file.path(root, "landing"), backend = "duckdb"
)
config
lake <- dl_connect(config)
```

In a long-running function, register `on.exit(dl_disconnect(lake))` immediately
after connecting. Here we explicitly clean up at the end of the example.

## Describe the source and its contract

A reader is a normal R function receiving the immutable landed file path.
The contract documents the meaning of one row and gates publication.

```{r source}
path <- file.path(root, "orders.csv")
utils::write.csv(data.frame(order_id = 1:3, amount = c(25, 75, 50)),
  path, row.names = FALSE)
source <- dl_source("orders.file", path, reader = utils::read.csv)
contract <- dl_contract(
  "orders.contract", "1.0.0", "Analytics", "Order amounts", "One order",
  columns = c(order_id = "integer", amount = "numeric"), key = "order_id"
)
contract
```

`numeric` permits integer or double input, while `integer` is strict. Missing
values are rejected for every declared column by default. Undeclared columns,
duplicate keys and empty deliveries also fail by default. See `?dl_contract`.

## Execute and query

```{r ingestion}
release <- dl_ingest(lake, source, contract, "orders", code_version = "v1")
release
orders <- dl_tbl(lake, "orders", release = release$release_id)
dplyr::collect(orders)
```

The result contains a run ID and an immutable release ID. Queries remain lazy
until `collect()`. `dl_tbl(lake, "orders")` resolves the latest published release;
pass a release ID for a reproducible report. Keep the connection open while
using lazy tables.

## Repeat safely and inspect quality

```{r history}
repeated <- dl_ingest(lake, source, contract, "orders", code_version = "v1")
repeated$status
dl_registry(lake, "quality_results") |>
  dplyr::select(rule, status, n_failed)
```

Identical bytes, definitions and business date reuse a prior release. A changed
reader, captured value or dependency requires a new definition version and
`code_version`. Fingerprints cannot detect all changes in a function's environment.

Failed validation does not replace the published release. By default it raises
an error after recording diagnostics. For interactive inspection use
`stop_on_failure = FALSE` and inspect the returned status.

## Close resources

```{r cleanup}
dl_disconnect(lake)
unlink(root, recursive = TRUE)
```

Next read `vignette("workflow-design")` for transforms or
`vignette("dbt-workflows")` for SQL models and relational analysis.
