---
title: "Build a monthly reporting workflow, step by step"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Build a monthly reporting workflow, step by step}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
```

This storage guide uses optional DuckDB >= 1.5.5. For a database-free start,
see [composing products](composing-products.html).


## What you are building

You receive a monthly spreadsheet with reserve amounts from two fictional
entities. You want reports to read a reliable dataset, corrections to remain
traceable, and last month's published report to remain explainable.

This walkthrough follows that entire story. The numbers are deliberately small:
**August starts at EUR 350, a correction changes it to EUR 370, and September
adds a separate EUR 390.** No production data, credentials or external services
are needed. If the purpose or terminology is unfamiliar, read
[Why lakefold?](why-lakefold.html) first.

| You can stop after... | What you will have |
|---|---|
| Step 2 | A local dataset that you can write and read |
| Step 3 | Current data and reproducible historical versions |
| Step 4 | A business check that prevents duplicate records |
| Step 5 | Several months in one current table |
| Steps 6 to 8 | Optional prepared products, reusable metrics and report records |

Read the steps in order and reuse the same R session. Later steps build on the
earlier objects. Step 6 is optional even if you continue to metrics in Step 7.
Step 8 builds on the metric from Step 7.

## Before you start

Run this installation once. The online guide works without installing local
vignettes or any optional integration.

```{r installation, eval=FALSE, purl=FALSE}
install.packages("remotes")
remotes::install_github("JanWein/lakefold")
```

The package requires R >= 4.2 and DuckDB >= 1.5.5. The core example uses only
lakefold and its normal dependencies. pointblank, dbt, dm, Shiny and Posit
services are optional.

## Step 1. Open a place for your reporting data

```{r open-lake}
library(lakefold)
knitr::opts_chunk$set(eval = requireNamespace("duckdb", quietly = TRUE))
root <- tempfile("lakefold-walkthrough-")
lake <- dl_open(root)
```

**What changed:** a local folder now contains a DuckDB database, a delivery
archive and a small configuration file. `lake` is the connection you pass to
the next functions. This example uses a fresh temporary folder so it is safe to
rerun. For a real project, use a dedicated path such as
`dl_open("my-reporting-lake")`; reopening that folder reads the same data.

You do not need to select layers, create a server, configure S3 or request a
PostgreSQL database to try the workflow. Close the connection when you finish.

## Step 2. Store the first August delivery

Imagine that these rows have just been read from an Excel workbook. We create
them directly in R so everyone can run the example without an input file.
`amount` is a reserve amount in EUR, measured at the date in `date`.

```{r first-delivery}
august_date <- as.Date("2026-08-31")
august <- data.frame(
  entity = c("North", "South"),
  date = rep(august_date, 2),
  amount = c(100, 250)
)

first <- dl_write(lake, august, "reserves")
first$status
dl_read(lake, "reserves")
sum(dl_read(lake, "reserves")$amount)
```

**Expected result: `published`, two rows, and a total of 350.**

`"reserves"` is the dataset's stable name, called its **asset ID**. Your report
can keep reading that name even when next month's input filename changes.
`first` contains the outcome of writing the data; `dl_read()` returns the data
itself as a tibble, an ordinary R table.

Without a custom contract, lakefold remembers the column names and types from
the first successful delivery. Future schema changes and empty tables are
blocked. Missing values are allowed. This is structural protection; it cannot
infer which business records should be unique or whether an amount is plausible.

Because we submitted an R data frame, the archive contains an RDS snapshot of
that object. The [real-file example below](#use-a-real-excel-or-csv-file)
explains how to preserve the original workbook instead.

## Step 3. Correct August without losing the first version

South sends a correction: its amount should have been 270, not 250.

```{r correction}
corrected <- august
corrected$amount[corrected$entity == "South"] <- 270
correction <- dl_write(lake, corrected, "reserves")

sum(dl_read(lake, "reserves")$amount)
sum(dl_read(lake, "reserves", release = first$release_id)$amount)
```

**Expected result: the current total is 370; the original total is still 350.**

A successful write creates a **release**, a checked version of the entire
published table. The default read selects the current release. Supplying
`release` selects one exact historical version. Save the release ID when an
analysis needs to keep referring to that version.

```{r repeat-delivery}
again <- dl_write(lake, corrected, "reserves")
again$status
```

**Expected result: `cached`.** The unchanged built-in workflow can reuse the
current release. It does not create another copy just because you reran the
same call. `dl_write()` only reuses the current release: deliberately writing an
older payload after a different release makes it current through a new release.
The advanced job runner has a separate historical retry policy, explained in
[Quality and history](quality-history.html).

## Step 4. Prevent a business mistake

A file with South listed twice still has the correct column types. A schema
check alone cannot know that it is wrong. Our business rule is:
**one row per entity and reporting date**.

Add a **contract**, the requirements that a proposed table must meet:

```{r business-contract}
contract <- dl_contract("reserves.contract",
  columns = c(entity = "character", date = "Date", amount = "numeric"),
  key = c("entity", "date"),
  max_age_hours = NULL
)
dl_write(lake, corrected, "reserves", contract = contract)
```

The key makes each entity/date combination unique. Explicit contracts also
require non-missing values in all declared columns by default. No owner or
long description is required. We leave the age deadline unset because monthly
deliveries need their own schedule, which can be added later.

Now simulate a duplicate South record:

```{r blocked-delivery}
bad <- rbind(corrected, corrected[2, ])
blocked <- dl_write(lake, bad, "reserves", contract = contract,
  stop_on_failure = FALSE)
blocked$status

dl_quality(lake, run_id = blocked$run_id) |>
  dplyr::filter(status == "failed") |>
  dplyr::select(rule, status, n_failed)

sum(dl_read(lake, "reserves")$amount)
```

**Expected result: `blocked`, a failed `unique_key` check, and a current total
of 370.** The duplicate would have inflated the total to 640. It never became
the published dataset.

For this lesson, `stop_on_failure = FALSE` lets us inspect the result. The
default raises an error after recording the failure, which is useful for jobs
that should stop. Once an execution attempt uses an explicit contract, keep
supplying one on later writes, even if the first contracted attempt failed.
This prevents silently dropping the checks you chose.

You can stop here. You already have a useful local workflow for complete
replacement deliveries, corrections and a business quality rule.

## Step 5. Keep several months in the current table

A release is a version of a table; a reporting date is a value inside that
table. Those are different things.

`dl_write()` replaces the complete current data state. Writing only September
with it would make September the current table; August would still be available
in older releases. If each delivery contains one month and you want **August
and September together in the current table**, use replacement by month.

Create September's complete delivery and add one argument:

```{r september-delivery}
september <- data.frame(
  entity = c("North", "South"),
  date = rep(as.Date("2026-09-30"), 2),
  amount = c(110, 280)
)
september_run <- dl_write(lake, september, "reserves", contract = contract,
  partition_by = "date", business_date = "2026-09-30")
dl_read(lake, "reserves") |>
  dplyr::arrange(date, entity)
```

**Expected result: four rows. August contains 100 and 270; September contains
110 and 280.** Both months now belong to the current release.

`partition_by = "date"` means: replace every date present in this delivery and
retain the other dates. `business_date` separately records the delivery's
reporting date in metadata; it does not create a column or select a partition.
The same argument works with CSV files and a custom date reader.

Each supplied month must be complete. A later September delivery replaces all
of September, so sending only South would remove North from September's
current state. This is month replacement, not a row-level upsert. The full
resulting candidate is checked, including the retained months.

No pipeline definition is needed for this common case. Introduce an explicit
[pipeline](workflow-design.html) when you need named transformations or more
control over execution. It uses the same landing, quality and release behavior.
To inspect a change, try `dl_compare(lake, "reserves")`: the contract already
declares its business key.

## Step 6. Reuse a prepared reporting table (optional)

Suppose several reports need monthly totals. Rather than repeating the same
aggregation in each report, define a **product**: a prepared table built from
recorded input releases.

```{r prepared-product}
totals_contract <- dl_contract("reserve_totals.contract",
  columns = c(date = "Date", total_reserve = "numeric"),
  key = "date", max_age_hours = NULL)

totals <- dl_product("reserve_totals",
  inputs = c(reserves = "reserves"),
  build = function(inputs) {
    inputs$reserves |>
      dplyr::group_by(date) |>
      dplyr::summarise(total_reserve = sum(amount, na.rm = TRUE), .groups = "drop")
  },
  contract = totals_contract, code_version = "tutorial-v1")

dl_build(lake, totals)
dl_read(lake, "reserve_totals") |> dplyr::arrange(date)
```

**Expected result: 370 for August and 390 for September.**

The builder receives the input tables and returns one output table. lakefold
records the exact input releases used and checks the output contract. In larger
workflows, the builder can join several inputs or perform other dplyr operations.
A new source release does not automatically rebuild this product; your job must
call `dl_build()` again when the prepared table should be refreshed.

## Step 7. Define what a reported number means (optional)

A reserve is a **stock** measured at a date. Adding August's 370 to September's
390 produces 760, but that is not the reserve at either reporting date.

A **metric** records the formula, unit, allowed grouping and time meaning. It
can use the published `"reserves"` asset directly; Step 6's product is optional.

```{r reusable-metric}
total_reserve <- dl_metric("total_reserve", "reserves",
  expr = sum(amount, na.rm = TRUE), dimensions = "entity",
  time_column = "date", time_behavior = "stock",
  unit = "EUR", owner = "Finance",
  description = "Total reserve at one reporting date",
  approved = TRUE, code_version = "tutorial-v1")

current_august <- dl_measure(lake, total_reserve, at = august_date)
current_august

dl_measure(lake, total_reserve, by = "entity", at = august_date)
```

**Expected result: 370 overall, or North 100 and South 270 when grouped.**

The stock setting requires exactly one selected reporting date. Once the table
contains both months, omitting `at` raises an error instead of adding the two
stocks. You explicitly set `approved = TRUE` here because the synthetic metric's
meaning has been reviewed for the example. This flag records a declared
approval; it does not perform a business review for you.

## Step 8. Preserve the number used in an earlier report (optional)

Suppose the August report was issued before South's correction. Its figure was
350. Recalculate against the original release, then record the result and its
provenance:

```{r report-record}
as_reported <- dl_measure(lake, total_reserve,
  at = august_date, release = first$release_id)
report <- dl_report_release(lake, "august-report-v1",
  results = list(total_reserve = as_reported), code_version = "tutorial-v1")

as_reported$value
current_august$value
report$id
```

**Expected result: 350 as originally reported, 370 in the corrected view, and
report ID `august-report-v1`.** This reconstruction assumes the same formula
was used originally. In a real workflow, freeze the metric results when the
report is issued so both the original formula and inputs are recorded.

The report record contains values, metric versions and input release IDs. It
does not render a document. Use those metric results in your existing Quarto,
Shiny, Word or other reporting workflow. Treat issued metric definitions and
report IDs as versioned records when business meaning changes.

## Check the whole story

| Event | Current published table | August in the original pinned release |
|---|---|---|
| First August delivery | August: 350 | 350 |
| South's correction | August: 370 | 350 |
| Duplicate South row | Unchanged: August 370 | 350 |
| Complete September delivery through the monthly pipeline | August: 370; September: 390 | 350 |

The value of the framework is this consistent behavior across steps. Your
choice of an Excel reader, a dplyr transformation or an optional SQL model
can change while the questions about inputs, checks and versions stay clear.

```{r verify-results, include=FALSE, purl=TRUE}
stopifnot(
  first$status == "published",
  correction$status == "published",
  again$status == "cached",
  blocked$status == "blocked",
  any(dl_quality(lake, run_id = blocked$run_id)$rule == "unique_key"),
  nrow(dl_read(lake, "reserves")) == 4L,
  sum(dl_read(lake, "reserves", release = first$release_id)$amount) == 350,
  current_august$value == 370,
  as_reported$value == 350,
  dl_measure(lake, total_reserve, at = as.Date("2026-09-30"))$value == 390,
  identical(dl_read(lake, "reserve_totals") |>
    dplyr::arrange(date) |> dplyr::pull(total_reserve), c(370, 390))
)
```

## Use a real Excel or CSV file

For your own workbook, run this after creating the contract in Step 4. The
workbook must have a `Reserves` sheet with columns `entity`, `date`, `amount` in
that order. This block needs your file and the optional readxl package, so it
is not run as part of the bundled example.

```{r real-excel, eval=FALSE, purl=FALSE}
install.packages("readxl") # once
read_reserves <- function(path) {
  readxl::read_excel(path, sheet = "Reserves",
    col_types = c("text", "date", "numeric")) |>
    dplyr::mutate(date = as.Date(date))
}
dl_write(lake, "incoming/august.xlsx", "reserves",
  contract = contract, reader = read_reserves)
```

Passing the path preserves the original workbook before parsing. If you
instead read a workbook yourself and pass the resulting data frame, lakefold
archives that R object, not the workbook. Excel dates are converted to R `Date`
values here to match the contract. Use the same reader in `dl_source()` if you
need Step 5's monthly replacement workflow for workbooks.

CSV, TSV and RDS have built-in readers. For the dated reserve schema, use an
explicit CSV reader as in Step 5. Other CSV files can use
`dl_write(lake, "orders.csv")` directly. Semicolon-separated files can use
`reader = utils::read.csv2` when their number format matches that reader.
Custom readers and rules are re-evaluated by default on the simple path; supply
`code_version` only when you deliberately want versioned cache reuse.

## Add pointblank checks when they help (optional)

The contract key already catches duplicates. If you also want amounts to be
nonnegative, pointblank can express that rule without writing a custom check
function. Install pointblank separately to run this block. It validates an R
data frame and needs no changes to the earlier published releases.

```{r pointblank-rule, eval=requireNamespace("pointblank", quietly=TRUE), purl=FALSE}
pb_contract <- dl_contract("reserves.with_amount_check",
  columns = c(entity = "character", date = "Date", amount = "numeric"),
  key = c("entity", "date"), max_age_hours = NULL,
  rules = list(dl_pointblank("nonnegative", function(data) {
    pointblank::create_agent(data) |>
      pointblank::col_vals_gte(columns = "amount", value = 0)
  })))
negative <- corrected
negative$amount[2] <- -270
amount_check <- dl_validate(negative, pb_contract)
amount_check |>
  dplyr::filter(status != "passed") |>
  dplyr::select(rule, status)
stopifnot(any(amount_check$status == "failed"))
```

The default rule policy blocks a failed check. For graded warning/blocking
thresholds, segments, an input gate before Raw and quality reports, continue
with [Quality gates](quality-gates.html). To apply this new contract to future
writes, supply it explicitly; updating an already registered definition requires
a new version. Installing pointblank alone does not add business rules.

## Inspect the history and close the connection

```{r history}
dl_releases(lake, "reserves") |>
  dplyr::select(release_id, business_date, quality)
dl_close(lake)

lake <- dl_open(root)
sum(dl_read(lake, "reserves", release = first$release_id)$amount)
dl_close(lake)
```

**Expected result after reopening: the pinned August release still totals 350.**
The data lives in the lake folder, not in the lifetime of the `lake` R variable.
A lazy read with `lazy = TRUE` needs an open connection; the default read collects
all rows into R. Filter or aggregate lazy tables before collecting large data.

The downloadable script runs the core example in a temporary folder, checks
these outcomes and cleans up its own demo data:

```{r run-script, eval=FALSE, purl=FALSE}
source(system.file("examples", "monthly_reporting.R", package = "lakefold"),
  echo = TRUE)
```

## Turn the prototype into a recurring job

Keep your contract and processing definitions in version-controlled R files.
Wrap the run in a function, open the lake inside it, and register
`on.exit(dl_close(lake), add = TRUE)` so errors also close the connection.
Point the source at the next delivery and run that function from your existing
scheduler. One coordinated process should write to a given lake.

| When the prototype needs... | Next addition |
|---|---|
| An existing Posit environment | Develop in Workbench; schedule the R job and publish reports on Connect |
| A visible history for colleagues | `dl_catalog_export()` and the optional read-only Shiny `dl_catalog()` |
| Expected-delivery alerts | `dl_check_delivery()` from a scheduler with explicit dates and deadlines |
| Shared Parquet storage | Create a DuckLake target with the required local or S3 storage configuration |
| Remote catalog metadata | Configure PostgreSQL after it has been provisioned; this does not enable concurrent writers by itself |
| A graph of SQL models | Add dbt, then explicitly publish checked snapshots when lakefold releases are needed |
| Relationships across tables | Declare keys with dm |

These are deployment or workflow choices, not prerequisites for the local
example. See [Operations](https://github.com/JanWein/lakefold/blob/main/docs/OPERATIONS.md),
[Workflow design](workflow-design.html), [dbt workflows](dbt-workflows.html)
and [Products and metrics](products-metrics.html) for the next relevant step.

```{r cleanup, include=FALSE, purl=FALSE}
unlink(root, recursive = TRUE)
```
