---
title: "Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
```

A workflow brings one product specification and one optional recipe together
with sources and a destination. Components remain independently reusable. This
guide runs without optional packages; publication examples identify their needs.

## Assemble components

```{r}
library(tidyweave)
spec <- tw_product("orders") |>
  tw_add_contract(c(id = "integer", amount = "numeric")) |>
  tw_add_quality(~ amount >= 0)
preparation <- tw_recipe() |> tw_step_mutate(amount = round(amount, 2))
flow <- tw_workflow() |> tw_add_product(spec) |> tw_add_recipe(preparation)
flow
```

An empty workflow can be inspected while it is being assembled. Execution needs
a product and usable inputs. A recipe is optional when the input already has
the required shape.

## Manage slots

| Operation | Product | Recipe |
|---|---|---|
| Fill an empty slot | `tw_add_product()` | `tw_add_recipe()` |
| Replace an occupied slot | `tw_update_product()` | `tw_update_recipe()` |
| Clear a slot | `tw_remove_product()` | `tw_remove_recipe()` |
| Retrieve its definition | `tw_extract_product()` | `tw_extract_recipe()` |

Add operations reject occupied slots. Updates require an existing component.
Extraction returns the definition, not executed data. Edits return a new value.

```{r}
whole_amounts <- tw_recipe() |> tw_step_mutate(amount = round(amount))
alternate <- flow |> tw_update_recipe(whole_amounts)
tw_extract_recipe(alternate)
tw_extract_recipe(flow)
```

## Bind deliveries

Supply a primary delivery when executing a reusable definition:

```{r}
delivery <- data.frame(id = 1:2, amount = c(10.123, 20))
result <- tw_trial(flow, data = delivery)
tw_collect(result)
```

Alternatively, add a named source. A table, a reader function or a source
adapter can occupy the slot.

```{r}
bound <- flow |> tw_add_source(delivery, name = "orders")
tw_plan(bound)
tw_collect(tw_trial(bound))
```

Sources are configured in one place: a workflow rejects sources that conflict
with sources already on its product. `data =` binds or replaces a single primary
delivery. Use named `sources =` replacements for several input deliveries,
including checked lookup dependencies. See [Compose products](composing-products.html).

## Choose execution deliberately

| Call | Runs preparation and checks | Runs framework writers |
|---|---|---|
| `tw_trial()` | Yes | No |
| `tw_run()` | Yes | Configured writers |
| `tw_publish()` | Yes | Chosen or configured destination |

`tw_trial()` also disables dependency writers. User callbacks can still have
their own effects. `tw_run()` is useful for configured orchestration;
`tw_publish()` makes saving an explicit action. All three produce execution
results, separate from the workflow definition.

```{r eval=FALSE}
# Requires DuckDB >= 1.5.5. The folder uses the local DuckDB backend.
saved <- tw_publish(flow, data = delivery, to = "data/orders")
tw_collect(saved)
```

Output checks gate publication. Lake publication retains immutable releases;
database, Parquet and pins targets retain their own documented guarantees.
[Publish and correct data](publish-data.html) demonstrates a complete correction.

## Inspect definitions and results

Use `tw_inspect()` for component descriptions and `tw_plan()` for the execution
plan without reading data. After execution, use `tw_status()`, `tw_quality()`
and `tw_lineage()` for evidence, and `tw_collect()` for usable data.

```{r}
tw_inspect(flow)$recipe
tw_status(result)
```

## How this relates to tidymodels

[parsnip](https://parsnip.tidymodels.org/) separates a specification from its
implementation; tidyweave applies that pattern to product requirements and
operation engines. [recipes](https://recipes.tidymodels.org/) provides the
ordered-step pattern. [workflows](https://workflows.tidymodels.org/) provides the
assembly and component-management pattern. tidyweave executes data products,
so it exposes trial and publication instead of model fitting and prediction.

For sequential orchestration, `tw_workflow()` also accepts named functions whose
arguments declare dependencies. That interface reuses unchanged branches after
corrections; see [Repeat a workflow](everyday-workflows.html). It is a separate
composition mode and does not accept product or recipe slots.
