---
title: "Specifications, recipes and workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Specifications, recipes and workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Goal and prerequisites

Create a reusable cancellation reporting workflow from synthetic policy,
customer and broker deliveries. You need tidyweave and its ordinary R
dependencies. The main example needs no database or optional package.
Expected result: two cancellations among four policies in force at the start
of the period, a cancellation rate of **0.5**.

## Three independently reusable objects

The design follows [parsnip](https://parsnip.tidymodels.org/),
[recipes](https://recipes.tidymodels.org/) and
[workflows](https://workflows.tidymodels.org/):

| Pattern | tidyweave object | Responsibility |
|---|---|---|
| Specification | `product()`, `contract()`, `quality_rule()` | Identity, schema, grain and checks |
| Preparation | `recipe() |> step_*()` | Ordered, deferred transformation instructions |
| Composition | `workflow() |> add_product() |> add_recipe()` | Bind inputs and execute the assembled components |
| Operation engine | `set_engine()` | Choose native/pointblank quality or native/dm relationship checking |

The analogy is structural. These are data product workflows, not statistical
model fits. There is no training phase, fitted preprocessing state, or implicit
`prep()`/`bake()`. The explicit lifecycle is define, trial, publish, collect.
Storage belongs in `publish(to = ...)` or `set_target()`; it is not a quality
engine. Existing dplyr, pointblank and dm adapters do the actual work.

```{r}
library(tidyweave)

customers <- data.frame(
  customer_id = 1:3,
  region = c("North", "South", "North")
)
brokers <- data.frame(
  broker_id = 1:2,
  channel = c("Broker", "Direct")
)
policies <- data.frame(
  policy_id = 1:5,
  customer_id = c(1L, 1L, 2L, 3L, 3L),
  broker_id = c(1L, 1L, 2L, 2L, 1L),
  at_start = c(TRUE, TRUE, TRUE, TRUE, FALSE),
  cancelled = c(TRUE, FALSE, FALSE, TRUE, FALSE)
)
```

## 1. Specify meaning and requirements

The input contract checks the delivery before filtering or enrichment. A
cancellation must belong to the opening population for this particular metric.
Other definitions, such as exposure-weighted rates, need a different contract
and measure. Product and contract definitions can be authored by their users.

```{r}
delivery_contract <- contract(
  columns = c(
    policy_id = "integer", customer_id = "integer", broker_id = "integer",
    at_start = "logical", cancelled = "logical"
  ),
  key = "policy_id",
  grain = "One policy observed over the reporting period",
  rules = list(valid_population = ~ !cancelled | at_start)
)

delivery <- product("policy_delivery", policies, contract = delivery_contract)

report_contract <- contract(
  columns = c(
    policy_id = "integer", customer_id = "integer", broker_id = "integer",
    at_start = "logical", cancelled = "logical",
    region = "character", channel = "character"
  ),
  key = "policy_id",
  grain = "One policy in force at the start of the reporting period"
)

report_spec <- product("cancellation_base", contract = report_contract)
```

## 2. Describe preparation once

Lookups check unique reference keys and complete input matches. They remain
explicit dependencies, so corrections and shared product execution work through
the same core as direct product pipelines.

```{r}
preparation <- recipe() |>
  step_filter(at_start) |>
  step_lookup(customers, by = "customer_id", name = "customers") |>
  step_lookup(brokers, by = "broker_id", name = "brokers") |>
  step_arrange(policy_id)

preparation
```

`step_mutate()`, `step_select()`, `step_rename()`, `step_distinct()` and
`step_summarise()` use normal dplyr expressions. `step_transform()` accepts
ordinary functions, formulas using `.x`, or existing transform adapters. Lazy
inputs remain lazy when their steps support that backend. Unsupported operations
do not silently collect data.

## 3. Assemble and try

```{r}
flow <- workflow() |>
  add_product(report_spec) |>
  add_recipe(preparation) |>
  add_source(delivery, name = "policies")

flow
plan(flow)
result <- trial(flow)
collect(result)

rate <- metric(
  "cancellation_rate", "cancellation_base",
  expr = mean(cancelled), dimensions = c("region", "channel"), unit = "ratio"
)
measure(result, rate)
stopifnot(nrow(collect(result)) == 4L, mean(collect(result)$cancelled) == 0.5)
```

The workflow contains one product and one optional recipe. `add_product()`
rejects products with embedded transformations so preparation has one clear
home. Direct product pipelines still support dplyr verbs. A recipe can also be
attached to a direct product with `add_recipe()`.

## 4. Reuse, replace and correct

```{r}
extract_product(flow)
extract_recipe(flow)

compact <- preparation |> step_select(policy_id, cancelled)
compact_spec <- product("compact_base") |>
  add_contract(c(policy_id = "integer", cancelled = "logical"))

compact_flow <- flow |>
  update_recipe(compact) |>
  update_product(compact_spec)

collect(trial(compact_flow))
stopifnot(identical(extract_recipe(flow), preparation))

corrected <- policies
corrected$cancelled[2] <- TRUE
corrected_result <- trial(flow, data = corrected)
stopifnot(mean(collect(corrected_result)$cancelled) == 0.75)
stopifnot(mean(collect(result)$cancelled) == 0.5)
```

`data =` follows the single primary-input chain and preserves the input contract.
Named `sources = list(customers = new_customers)` replaces reference deliveries.
Each execution returns a new result; the definition and earlier results stay
unchanged. `remove_recipe()` removes preparation and `remove_product()` clears
the specification slot. Duplicate additions and ambiguous inputs fail with
repair instructions instead of silently overriding another component.

## 5. Choose an engine for an operation

```{r}
opening_population <- quality_rule("opening_population", ~ at_start)
pointblank_rule <- opening_population |> set_engine("pointblank")

customer_relationship <- lookup_spec(customers, by = "customer_id", name = "customers")
dm_relationship <- customer_relationship |> set_engine("dm")

alternative_preparation <- recipe() |>
  step_filter(at_start) |>
  step_transform(dm_relationship) |>
  step_lookup(brokers, by = "broker_id", name = "brokers")
```

These definitions need no optional engine installed. Executing the pointblank
rule requires pointblank; executing the dm relationship requires dm. The
predicate and relationship keys stay the same. Explicit engine choices override
workflow `execution_config()` defaults. A missing engine fails at preflight.
Functions used as quality checks are not silently reinterpreted as pointblank
agent builders.

## Publication and relational products

Use `publish(flow, to = destination, data = next_delivery)` with an existing
lake configuration or target adapter. Failed checks block publication.
`trial(flow)` disables configured framework writers throughout the compiled
product dependency graph. User-supplied callbacks remain ordinary R code and
can perform their own side effects.

A model product keeps a dm model's tables and relationships. Prepare member
table products with recipes, then assemble the model as in the
[cancellation report](cancellation-report.html). Add the resulting model product
to a workflow without a whole-model recipe. Model products retain their existing
in-memory execution limit and consistent publication behavior.

The function-based `workflow(step = function(input) ..., inputs = ...,
code_version = ...)` API still describes a dependency graph across several
operations. It is useful for sequencing receipts and dbt builds; each modular
product workflow can be executed inside one of those functions. It does not
provide a transaction across independently committed steps.

Next: [publish checked data](publish-data.html) or
[configure a shared DuckLake](shared-ducklake.html).
