---
title: "Add integrations only when you need them"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Add integrations only when you need them}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tidyweave)
root <- tempfile("tidyweave-integrations-")
dir.create(root)
```

Start with [composing products](composing-products.html) if you are new to
tidyweave. A data frame and a function are enough for useful work. This guide
explains what to add when a workflow needs another storage system, an HTTP
source, dependency-aware execution or a deployment environment.

## Keep the product, change its destination

The business definition stays small. This product does not choose infrastructure.

```{r definition}
orders_definition <- product("orders", data.frame(id = 1:3, amount = c(25, 75, 50))) |>
  dplyr::mutate(amount = amount * 2) |>
  add_quality(~ amount >= 0)

collect(run(orders_definition))
```

Choose a target when you need to write the checked result. `set_target()` replaces
only that component; the transformations and quality gate stay the same.

```{r sqlite, eval=requireNamespace("RSQLite", quietly=TRUE)}
con <- DBI::dbConnect(RSQLite::SQLite(), ":memory:")
result <- orders_definition |>
  set_target(target_database(con, "orders")) |>
  run(evidence = file.path(root, "evidence"))
DBI::dbReadTable(con, "orders")

# A caller-owned connection permits lazy database transformations.
from_database <- product("large.orders") |>
  add_source(source_database(con, table = "orders")) |>
  dplyr::filter(amount >= 100) |>
  add_quality(~ amount >= 0) |>
  run()
collect(from_database)
DBI::dbDisconnect(con)
```

`source_database()` accepts a `DBI::Id()` or SQL query as well as a plain table
name. Parameterized queries use `params` and return materialized data.
Caller-owned connections stay open; close them after collecting lazy results.
A zero-argument connection factory opens a fresh connection during execution,
materializes the source and closes it, including after failure.

```{r parquet, eval=requireNamespace("arrow", quietly=TRUE)}
parquet_path <- file.path(root, "orders.parquet")
orders_definition |>
  set_target(target_parquet(parquet_path)) |>
  run()

from_parquet <- product("parquet.orders") |>
  add_source(source_parquet(parquet_path)) |>
  dplyr::filter(amount >= 100) |>
  add_quality(~ amount >= 0) |>
  run()
collect(from_parquet)
```

An Arrow source stays lazy through supported dplyr transformations and native
aggregate checks. `collect()` and targets that require ordinary R tables create
materialization boundaries. A lazy workflow can still issue count/schema queries;
"lazy" does not mean "no queries until collect".

```{r pins, eval=requireNamespace("pins", quietly=TRUE)}
board <- pins::board_folder(file.path(root, "pins"), versioned = TRUE)
pinned <- orders_definition |>
  set_target(target_pins(board, "orders")) |>
  run()

read_source(source_pins(board, "orders", version = pinned$outputs$version))
```

Version retention and authentication belong to the pins board. Reading a reported
pin version is distinct from the immutable-release guarantees of a governed lake.

## Know each adapter's boundaries

| Component | Lazy or materialized? | History and write boundary | Resource ownership |
|---|---|---|---|
| R data frame/function | Ordinary R objects; a function may return a supported lazy table | No persistence without a target | Your R session |
| DBI source | Lazy with an open caller connection; factories and parameterized queries collect | Reads the current table/query, without a source snapshot guarantee | Caller closes borrowed connections; factories close automatically |
| DBI target | Collects the checked candidate | Transaction required by default; driver determines isolation and transactional DDL; no immutable history | Caller or factory, as above |
| Arrow/Parquet source | Lazy by default | Reads current files/dataset; remote support and credentials belong to Arrow | Arrow manages dataset handles |
| Parquet target | Collects to one local file | Temporary sibling file plus rename; replacement depends on filesystem support; no version history | No retained connection |
| pins source/target | Materialized | Board-defined storage/versioning; no package-wide atomicity promise | Configured pins board |
| Lake target | Materializes for governed publication | Immutable releases and full-candidate gates; one coordinated writer | Owned connections close; caller connections remain open |
| httr2 API source | Collects pages into one table | No consistent remote snapshot unless the API supplies one | Request factory obtains runtime authentication |
| dbt transform | Stages input and materializes selected output | dbt owns model writes; failed final product checks do not undo them | Fresh connections close before/after the CLI |

`capabilities(component)` describes declared support; it does not prove remote
permissions or a driver's transaction guarantees. For DBI append, the adapter
checks the combined existing and incoming table inside its transaction. This
requires collecting the existing table. `collect(result)` still represents the
incoming batch; `result$outputs$rows` reports the full destination size.

`target_parquet()` writes a **single local file**, not a remote object or a
partitioned dataset. It refuses replacement unless `overwrite = TRUE`. For larger
storage requirements choose an appropriate backend or implement a target adapter.

## PostgreSQL and SAP HANA through DBI

Keep credentials and driver choices in a factory. The product grammar is the same.
These examples require your installed driver, configured service and permissions;
local SQLite tests do not verify a PostgreSQL or HANA deployment.

```{r enterprise-dbi, eval=FALSE}
postgres_connection <- function() {
  DBI::dbConnect(RPostgres::Postgres(),
    host = Sys.getenv("PGHOST"), dbname = Sys.getenv("PGDATABASE"),
    user = Sys.getenv("PGUSER"), password = Sys.getenv("PGPASSWORD"))
}
hana_connection <- function() {
  DBI::dbConnect(odbc::odbc(), dsn = Sys.getenv("HANA_DSN"))
}

product("contracts") |>
  add_source(source_database(hana_connection,
    table = DBI::Id(schema = "RISK", table = "CONTRACTS"))) |>
  add_quality(~ amount >= 0) |>
  set_target(target_database(postgres_connection,
    DBI::Id(schema = "analytics", table = "contracts"))) |>
  run(evidence = "runs")
```

Factories prioritize clear connection ownership and return materialized sources.
Keep a caller-owned connection open for database-side dplyr execution. Supported
SQL translation depends on [dbplyr](https://dbplyr.tidyverse.org/) and the driver.

## HTTP requests, pagination and credentials

Use [httr2](https://httr2.r-lib.org/) for authentication, retry policy and throttling.
The following API example expects `data` and `next_cursor` fields. Adapt the
parser and pagination callback to the actual API contract.

```{r api, eval=FALSE}
make_request <- function(cursor = NULL) {
  httr2::request(Sys.getenv("ORDERS_API_URL")) |>
    httr2::req_auth_bearer_token(Sys.getenv("ORDERS_API_TOKEN")) |>
    httr2::req_url_query(cursor = cursor) |>
    httr2::req_retry(max_tries = 4) |>
    httr2::req_throttle(rate = 5)
}
api <- source_api(
  request = function() make_request(),
  parse = function(response) {
    body <- httr2::resp_body_json(response, simplifyVector = TRUE)
    tibble::as_tibble(body$data)
  },
  next_request = function(response) {
    cursor <- httr2::resp_body_json(response)$next_cursor
    if (is.null(cursor)) NULL else make_request(cursor)
  },
  max_pages = 100
)
api_definition <- product("api.orders") |>
  add_source(api) |>
  add_quality(~ amount >= 0)
run(api_definition, evidence = "runs")
```

The factory reads fresh credentials when execution starts. The callback rebuilds
subsequent requests with the same policies. Repeated requests, inconsistent page
schemas and exceeding `max_pages` fail the run instead of silently truncating
it. Without a configured retry bound, the adapter uses up to three httr2 attempts.

## DuckLake, S3 and external catalogs

A lake's technical registry records releases and execution. An external catalog
receives descriptive metadata. Configure these independently.

```{r remote-storage, eval=FALSE}
config <- lake_config(
  catalog = registry_postgres("DUCKLAKE_PG_CONNECTION"),
  storage = storage_s3(
    bucket = Sys.getenv("TIDYWEAVE_S3_BUCKET"),
    prefix = "tidyweave/production",
    endpoint = Sys.getenv("TIDYWEAVE_S3_ENDPOINT")
  ),
  landing = "landing-cache",
  backend = "ducklake"
)
orders_definition |>
  set_target(target_lake(config)) |>
  run(evidence = "runs")
```

This configuration provisions no services and migrates no existing local lake.
Verify DuckLake, S3 conditional writes, PostgreSQL permissions, backups and restore
in your own environment. The full configuration example is
`system.file("examples", "setup_s3.R", package = "tidyweave")`.
See the [operations guide](https://github.com/JanWein/tidyweave/blob/main/docs/OPERATIONS.md)
for OpenLineage, OpenMetadata and retrying pending metadata deliveries.

## Let targets manage product dependencies

A nested product is an upstream dependency. `as_targets()` converts the same
composition to real [targets](https://books.ropensci.org/targets/) targets.
Save the following as `_targets.R`; this example is displayed without executing
a build inside the vignette.

```{r targets-script, eval=FALSE}
library(tidyweave)

orders_definition <- product("orders") |> add_source("orders.csv")
totals_definition <- product("totals") |>
  add_source(orders_definition) |>
  add_transform(function(data) dplyr::summarise(data, total = sum(amount)))

as_targets(totals_definition, evidence = "runs")
```

Then run:

```{r targets-run, eval=FALSE}
targets::tar_make()
collect(targets::tar_read(totals))
```

Each product executes once. Downstream products consume the completed upstream
output and retain its run/release reference. Local file changes invalidate the
affected products. Unchanged targets are reused, so they keep the original run
ID and timestamps and create no new execution-evidence record. Ordinary statically
referenced function globals are tracked; dynamic lookups, mutable environments
and remote database/API changes need an explicit cue:

```{r targets-cue, eval=FALSE}
as_targets(api_definition, cue = targets::tar_cue(mode = "always"),
  evidence = "runs")
```

Use connection factories in stored pipelines. A live DBI connection or lazy-table
handle cannot be safely serialized into a targets cache. Definition variables
such as `orders_definition` avoid colliding with the target named `orders`.
Targets provides dependency scheduling and caching; a clock-based schedule still
belongs to your CI runner, scheduler or deployment platform.

## Start a project without configuring everything now

```{r project}
project_path <- file.path(root, "project")
init_project(project_path, targets = TRUE, renv = TRUE, connect = TRUE)
list.files(project_path)
```

The generated project runs with `source("run.R")` from its directory. Optional
files have separate purposes:

| File | What you do next |
|---|---|
| `_targets.R` | Install targets and call `targets::tar_make()` |
| `init-renv.R` | Install renv and explicitly run the script to create a lockfile |
| `job.qmd` | Render locally, publish to Posit Connect, then configure its schedule |

Project creation installs nothing and publishes nothing. Keep the resulting
`renv.lock` under version control, and restore it on another machine. Set deployment
credentials in the runtime, and choose persistent storage for `TIDYWEAVE_EVIDENCE`.
See [renv](https://rstudio.github.io/renv/) and
[Connect scheduling](https://docs.posit.co/connect/user/scheduling/) for their
respective lifecycle responsibilities.

For an independently runnable synthetic example combining SQLite, Parquet, pins,
quality, evidence and a metadata callback:

```{r standalone-example, eval=FALSE}
source(system.file("examples", "advanced_integrations.R", package = "tidyweave"))
```

```{r cleanup, include=FALSE}
unlink(root, recursive = TRUE)
```
