---
title: "Preparation recipes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparation recipes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
```

A recipe stores ordered preparation instructions. Build it once and attach it
to multiple workflows. It does not own product requirements, primary deliveries
or destinations. This guide runs with tidyweave's hard dependencies.

## Add steps

```{r}
library(tidyweave)
preparation <- tw_recipe() |>
  tw_step_mutate(amount = round(amount, 2)) |>
  tw_step_select(id, region, amount)
preparation
```

Each call returns an updated recipe. Steps execute in pipe order with ordinary
dplyr data masking and tidy selection. Expressions retain their R environments;
inject a small value with `!!` if you need to freeze it at definition time.

| Purpose | Steps |
|---|---|
| Calculate or rename columns | `tw_step_mutate()`, `tw_step_rename()` |
| Select rows and columns | `tw_step_filter()`, `tw_step_select()` |
| Order or deduplicate rows | `tw_step_arrange()`, `tw_step_distinct()` |
| Aggregate within groups | `tw_step_summarise(..., .by = group)` |
| Enrich by a checked relationship | `tw_step_lookup()` |
| Use an ordinary function or adapter | `tw_step_transform()` |

## Execute through a workflow

```{r}
delivery <- data.frame(id = 1:3, region = c("North", "North", "South"),
  amount = c(10.124, 20, 30))
flow <- tw_workflow() |>
  tw_add_product(tw_product("orders")) |>
  tw_add_recipe(preparation)
tw_collect(tw_trial(flow, data = delivery))
```

A recipe has no fitted training state. There is no `prep()` or `bake()` phase:
execution acquires the input, applies the instructions, and checks the output.

## Reuse and extend

```{r}
regional <- preparation |>
  tw_step_summarise(total = sum(amount), .by = region)
regional_flow <- tw_workflow() |>
  tw_add_product(tw_product("regional_orders")) |>
  tw_add_recipe(regional)
totals <- tw_collect(tw_trial(regional_flow, data = delivery))
totals
```

`preparation` still has two steps. `regional` adds a third. The totals are
**30.12** for North and **30** for South. Put output checks on the product that
matches each recipe's output shape.

## Add a checked lookup

```{r}
regions <- data.frame(region = c("North", "South"), currency = c("EUR", "EUR"))
enriched <- preparation |>
  tw_step_lookup(regions, by = "region", name = "regions")
enriched_flow <- flow |> tw_update_recipe(enriched)
tw_collect(tw_trial(enriched_flow, data = delivery))
```

The lookup checks reference-key uniqueness and, by default, requires every input
key to match. `unmatched = "keep"` retains unmatched input rows explicitly.
For a reusable relationship, create `tw_lookup_spec()` and attach it with
`tw_step_transform()`. See [Engines](engines.html) for optional dm checking.

## Extend with ordinary R

```{r}
custom <- tw_recipe() |>
  tw_step_transform(\(data) dplyr::mutate(data, amount = amount * 1.2),
    name = "add_tax")
tw_inspect(custom)
```

The function receives the preceding step's result. It must support the data
representation it receives. Backend-supported operations remain lazy; tidyweave
does not silently collect a query to make an unsupported step work. Write an
[adapter](extending-tidyweave.html) when you need reusable configuration or
declared capabilities beyond an ordinary function.

Continue with [Workflows](modular-workflows.html).

```{r verify, include=FALSE}
stopifnot(length(preparation$steps) == 2L,
          isTRUE(all.equal(sort(totals$total), c(30, 30.12))))
```
