---
title: "Group metrics and inspect relationships"
description: "Add dimensions, explicit time meaning and optional dm exploration."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Group metrics and inspect relationships}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Use this after [Save a report](save-report.html) for grouped calculations
and relationships across tables. Storage sections require DuckDB; relationship
exploration additionally uses dm. The example is self-contained.

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
library(tidyweave)
```

A product gives a useful result a name and a repeatable definition. A metric
gives a business calculation explicit meaning. Add each only when you need
it: an ordinary `dplyr::summarise()` remains useful for exploration.

## Start with one table and a checked lookup

Keep the primary row grain explicit. A lookup adds descriptive attributes without
multiplying rows; its parent key must be unique and non-missing.

```{r multiple-sources}
orders <- data.frame(id = 1:3, customer_id = c("a", "a", "b"),
  amount = c(10, 20, 30))
customers <- data.frame(customer_id = c("a", "b"), region = c("EU", "US"))
regional <- product("regional_orders", orders) |>
  add_lookup(customers, by = dplyr::join_by(customer_id))
collect(run(regional))
```

The native lookup rejects unmatched child keys by default. Select `engine = "dm"`
for the same promises using dm validation. General joins remain ordinary R code;
review their cardinality and the meaning of one row. A product publishes one table;
multiple sources do not imply an atomic multi-table publication. A product can also be the source of another product:
shared dependencies run once within a single execution.

## Use governed releases as interchangeable sources

This section requires DuckDB and creates a temporary local lake. A contract is
optional until you need a promise beyond the inferred structure:

```{r source-releases, eval=requireNamespace("duckdb", quietly = TRUE)}
root <- tempfile("tidyweave-products-")
lake <- open_lake(root)
orders_release <- orders |> publish("orders", to = lake)
customers_release <- customers |> publish("customers", to = lake)
definition <- product("regional_orders", orders_release) |>
  add_lookup(customers_release, by = dplyr::join_by(customer_id)) |>
  add_contract(contract(columns = c(id = "integer", customer_id = "character",
    amount = "numeric", region = "character"), key = "id")) |>
  set_target(lake)
published <- run(definition)
read_release(lake, "regional_orders")
lineage(lake, "regional_orders")
```

The inputs changed from ordinary R tables to pinned lake releases; the lookup
stayed the same. Successful lake results preserve their exact release IDs.
Their normalized configuration-based sources materialize the selected values.
For lazy access through a caller-owned connection, use
`source_release(lake, "orders", orders_release$release_id)` and keep the lake open.
Omitting the release ID selects the current release once during execution.
Later publication does not automatically rerun this product; use your scheduler
or the optional targets integration when you need that automation.

## Add relationship exploration with dm

`quality_reference()` is sufficient for an individual foreign-key check. Use
optional `dm` when you also want a native relational model for several tables:

```{r relational-model, eval=requireNamespace("duckdb", quietly = TRUE) && requireNamespace("dm", quietly = TRUE)}
relationships <- model(
  lake,
  tables = c(orders = "orders", customers = "customers"),
  primary_keys = list(orders = "id", customers = "customer_id"),
  foreign_keys = list(list(table = "orders", columns = "customer_id",
    ref_table = "customers", ref_columns = "customer_id")),
  releases = c(orders = orders_release$release_id,
    customers = customers_release$release_id)
)
dm::dm_examine_constraints(relationships)
```

Declared keys and relationships are checked. The result is a normal dm object
containing lazy tables. It does not choose your joins or replace review of the
intended row grain.

## Turn an agreed calculation into a metric

For a governed metric, define its input asset, formula and permitted grouping.
Approval and code identity are explicit because reports reuse this definition;
owner, description and unit are optional:

```{r metric, eval=requireNamespace("duckdb", quietly = TRUE)}
revenue <- metric("orders.revenue", "regional_orders", sum(amount, na.rm = TRUE),
  dimensions = "region", approved = TRUE, code_version = "revenue-v1")
values <- measure(published, revenue, by = "region")
values
```

The values are 30 for EU and 30 for US. `approved = TRUE` records your declared
decision; it is not a separate authorization system. Change `code_version`
when custom code, captured values or dependencies change. Arbitrary closure
environments are not fingerprinted.

| Choice | Meaning |
|---|---|
| No `time_column` | A flow metric may aggregate several dates |
| A `time_column` | A stock metric requires exactly one selected date |
| `dimensions` | Allowed grouping and filter columns |
| `na_policy = "reject"` | Missing relevant inputs block calculation |
| `na_policy = "expression"` | The formula deliberately handles missing inputs |

`sum(amount, na.rm = TRUE)` does not bypass the default missing-input gate.
For ratios or other custom calculations, supply
`compute = function(data, dimensions, params) ...` and declare `input_columns`.
The callback must implement the requested grouping and return its required
columns. Missing or non-finite numeric outputs and empty outputs are rejected.

## Save the values and their explanation together

```{r report, eval=requireNamespace("duckdb", quietly = TRUE)}
report <- report_release(lake, "regional-report-v1", list(revenue = values),
  code_version = "report-v1")
report_read(lake, "regional-report-v1", values_only = TRUE)
```

The measurement's manifest records the formula, parameters, pinned input
release, input quality and result hash. The report stores those values and
their manifests. Reading it does not recalculate the metric. Identical retries
preserve the first saved timestamps; changed data, definitions, parameters or
values need a new report ID. A report ID identifies a published report, while a
metric version identifies its definition.

Use `measure(..., record = FALSE)` for an exploratory calculation without
registry writes, or open a read-only lake. PDF/Word rendering, authentication
and scheduling belong to the surrounding tools. The calculation itself does
not need an LLM or a BI server.

```{r cleanup, include=FALSE, eval=requireNamespace("duckdb", quietly = TRUE)}
stopifnot(sum(values$value) == 60,
  nrow(report_read(lake, "regional-report-v1", values_only = TRUE)$revenue) == 2)
close_lake(lake)
unlink(root, recursive = TRUE)
```


[Back to Learn](learn.html) · [Troubleshooting](troubleshooting.html)
