---
title: "Products, relationships and reproducible metrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Products, relationships and reproducible metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
library(tidyweave)
```

A product gives a useful result a name and a repeatable definition. A metric
gives a business calculation explicit meaning. Add each only when you need
it: an ordinary `dplyr::summarise()` remains useful for exploration.

## Start with ordinary tables and functions

One source gives a transformation one table. Multiple named sources give it a
named list. Both use the same product object and execution function:

```{r multiple-sources}
orders <- data.frame(id = 1:3, customer_id = c("a", "a", "b"),
  amount = c(10, 20, 30))
customers <- data.frame(customer_id = c("a", "b"), region = c("EU", "US"))
regional <- product("regional_orders") |>
  add_source(orders, name = "orders") |>
  add_source(customers, name = "customers") |>
  add_transform(function(inputs) {
    dplyr::left_join(inputs$orders, inputs$customers, by = "customer_id")
  }) |>
  add_quality(~ !is.na(region))
collect(run(regional))
```

The join remains ordinary R code. Review its cardinality and the meaning of one
row. A product publishes one table; multiple sources do not imply an atomic
multi-table publication. A product can also be the source of another product:
shared dependencies run once within a single execution.

## Use governed releases as interchangeable sources

This section requires DuckDB and creates a temporary local lake. A contract is
optional until you need a promise beyond the inferred structure:

```{r source-releases, eval=requireNamespace("duckdb", quietly = TRUE)}
root <- tempfile("tidyweave-products-")
lake <- open_lake(root)
orders_release <- write_data(lake, orders, "orders")
customers_release <- write_data(lake, customers, "customers")
definition <- product("regional_orders") |>
  add_source(source_release(lake, "orders", orders_release$release_id),
    name = "orders") |>
  add_source(source_release(lake, "customers", customers_release$release_id),
    name = "customers") |>
  add_transform(function(inputs) {
    dplyr::left_join(inputs$orders, inputs$customers, by = "customer_id")
  }) |>
  add_contract(contract(columns = c(id = "integer", customer_id = "character",
    amount = "numeric", region = "character"), key = "id")) |>
  set_target(lake)
published <- run(definition)
read_release(lake, "regional_orders")
lineage(lake, "regional_orders")
```

The source implementation changed from R tables to pinned lazy tables; the
transformation stayed the same. Explicit release IDs reproduce exact inputs.
An omitted ID resolves the current release once during execution. Later source
publication does not automatically rerun this product; use your scheduler or
the optional targets integration when you need that automation.

## Add relationship exploration with dm

`quality_reference()` is sufficient for an individual foreign-key check. Use
optional `dm` when you also want a native relational model for several tables:

```{r relational-model, eval=requireNamespace("duckdb", quietly = TRUE) && requireNamespace("dm", quietly = TRUE)}
relationships <- model(
  lake,
  tables = c(orders = "orders", customers = "customers"),
  primary_keys = list(orders = "id", customers = "customer_id"),
  foreign_keys = list(list(table = "orders", columns = "customer_id",
    ref_table = "customers", ref_columns = "customer_id")),
  releases = c(orders = orders_release$release_id,
    customers = customers_release$release_id)
)
dm::dm_examine_constraints(relationships)
```

Declared keys and relationships are checked. The result is a normal dm object
containing lazy tables. It does not choose your joins or replace review of the
intended row grain.

## Turn an agreed calculation into a metric

For a governed metric, define its input asset, formula and permitted grouping.
Approval and code identity are explicit because reports reuse this definition;
owner, description and unit are optional:

```{r metric, eval=requireNamespace("duckdb", quietly = TRUE)}
revenue <- metric("orders.revenue", "regional_orders", sum(amount),
  dimensions = "region", approved = TRUE, code_version = "revenue-v1")
values <- measure(lake, revenue, by = "region", release = published$release_id)
values
```

The values are 30 for EU and 30 for US. `approved = TRUE` records your declared
decision; it is not a separate authorization system. Change `code_version`
when custom code, captured values or dependencies change. Arbitrary closure
environments are not fingerprinted.

| Choice | Meaning |
|---|---|
| No `time_column` | A flow metric may aggregate several dates |
| A `time_column` | A stock metric requires exactly one selected date |
| `dimensions` | Allowed grouping and filter columns |
| `na_policy = "reject"` | Missing relevant inputs block calculation |
| `na_policy = "expression"` | The formula deliberately handles missing inputs |

`sum(amount, na.rm = TRUE)` does not bypass the default missing-input gate.
For ratios or other custom calculations, supply
`compute = function(data, dimensions, params) ...` and declare `input_columns`.
The callback must implement the requested grouping and return its required
columns. Missing or non-finite numeric outputs and empty outputs are rejected.

## Save the values and their explanation together

```{r report, eval=requireNamespace("duckdb", quietly = TRUE)}
report <- report_release(lake, "regional-report-v1", list(revenue = values),
  code_version = "report-v1")
report_read(lake, "regional-report-v1", values_only = TRUE)
```

The measurement's manifest records the formula, parameters, pinned input
release, input quality and result hash. The report stores those values and
their manifests. Reading it does not recalculate the metric. Identical retries
preserve the first saved timestamps; changed data, definitions, parameters or
values need a new report ID. A report ID identifies a published report, while a
metric version identifies its definition.

Use `measure(..., record = FALSE)` for an exploratory calculation without
registry writes, or open a read-only lake. PDF/Word rendering, authentication
and scheduling belong to the surrounding tools. The calculation itself does
not need an LLM or a BI server.

```{r cleanup, include=FALSE, eval=requireNamespace("duckdb", quietly = TRUE)}
stopifnot(sum(values$value) == 60,
  nrow(report_read(lake, "regional-report-v1", values_only = TRUE)$revenue) == 2)
close_lake(lake)
unlink(root, recursive = TRUE)
```
