---
title: "Publish and correct data"
description: "Keep the first accepted delivery when a correction changes the current table."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Publish and correct data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
  eval = requireNamespace("duckdb", quietly = TRUE))
```

**Goal:** save a checked table and retain its earlier version after a correction.
**Needs:** tidyweave and DuckDB >= 1.5.5. Install DuckDB once with
`install.packages("duckdb")`. No database server is needed.

## Define the table and its destination

```{r definition}
library(tidyweave)
folder <- tempfile("orders-history-")
delivery <- data.frame(id = 1:3, amount = c(25, 75, 50))
spec <- tw_product("orders") |>
  tw_add_contract(tw_contract(
    columns = c(id = "integer", amount = "numeric"), key = "id")) |>
  tw_add_quality(~ amount >= 0)
orders <- tw_workflow() |>
  tw_add_product(spec) |>
  tw_add_source(delivery, name = "orders") |>
  tw_set_target(folder)
```

The temporary folder makes this example disposable. For real work, choose a
project folder such as `"data/orders-lake"` instead. `tw_set_target()` remembers
where to save the result. A new folder uses ordinary DuckDB; actual DuckLake is
an explicit option covered in [Create a DuckLake](create-ducklake.html).

## Save the first accepted delivery

```{r first}
first <- orders |> tw_publish()
tw_collect(first)
```

`tw_publish()` writes checked data to the chosen destination. It does not make
your data public on the internet. The connection is opened and closed for you.
A **release** is a saved version of the table. `first` refers to this exact
release, whose total is **150**.

## Publish a corrected complete table

```{r correction}
corrected <- data.frame(id = 1:3, amount = c(30, 75, 50))
second <- orders |> tw_publish(data = corrected)
tw_collect(second)
tw_collect(first)
tw_compare(first, second)$numeric_summary
```

The current total is **155**, and the first release still totals **150**.
The comparison shows an increase of **5**.

This delivery replaces the **complete table**. If you supplied only order 1,
orders 2 and 3 would disappear from the current table. For complete-month
replacement, use the explicit partition choice in [Monthly reporting](getting-started.html).
Do not submit a partial correction as though it were a complete delivery.

## Read it in another session

```{r reopen}
lake <- tw_open_lake(folder, read_only = TRUE)
tw_read_release(lake, "orders")
tw_close_lake(lake)
```

In a new R session, load tidyweave and use your durable folder path here. You do
not need the original R result object to read the current table.

Use `tw_trial()` to experiment even after setting a destination. `tw_publish()` makes
saving explicit. The more general `tw_run()` executes configured writers too;
it is useful for orchestration but is not a substitute for a write-free trial.
Lake writes need one coordinated writer.

```{r verify-and-cleanup, include=FALSE}
stopifnot(sum(tw_collect(first)$amount) == 150, sum(tw_collect(second)$amount) == 155)
unlink(folder, recursive = TRUE)
```

## Continue

[Previous: Check a delivery](check-delivery.html) · [Next: Save a report](save-report.html) · [Get started](get-started.html)
