---
title: "Quality and history: explain a result and reproduce it"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quality and history: explain a result and reproduce it}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
library(tidyweave)
has_duckdb <- requireNamespace("duckdb", quietly = TRUE)
```

A result should tell you what happened. A saved release should let you use the
same data again. These are related capabilities, but they answer different
questions: execution evidence explains an attempt; release history preserves
published data.

## Inspect an unsuccessful attempt

You can inspect a failed quality gate without setting up storage:

```{r failed-attempt}
definition <- product("orders") |>
  add_source(data.frame(id = 1:2, amount = c(10, -1))) |>
  add_quality(~ amount >= 0)
out <- run(definition, stop_on_failure = FALSE)
status(out)
quality(out)
```

Use the default `stop_on_failure = TRUE` for scheduled jobs so the scheduler
sees failure. A condition raised after execution carries the result in
`condition$result`. `stop_on_failure = FALSE` returns the same evidence for
interactive diagnosis; it does not permit publishing failed data.

| Quality status | Meaning | Allows publication? |
|---|---|---|
| `passed` | Check completed within its tolerance | Yes |
| `warning` | A non-blocking rule exceeded its tolerance | Yes |
| `failed` | A blocking rule exceeded its tolerance | No |
| `error` | Evaluation did not complete successfully | No |
| `not_checked` | No evaluable test units or an inactive check | No |

Warnings therefore differ from skipped checks. An empty result is not evidence
that the data passed. See [quality gates](quality-gates.html) for row checks,
contracts, reference checks and pointblank policies.

## Keep evidence beyond the current R session

Choose an evidence directory when you need a durable record for in-memory or
external targets. Data rows, connections and raw exception text are excluded.

```{r durable-evidence}
evidence_dir <- tempfile("tidyweave-runs-")
out <- run(definition, stop_on_failure = FALSE, evidence = evidence_dir)
run_history(evidence_dir)
incidents(evidence_dir)
```

The record includes status, timings, source and output metadata, lineage and
quality results. `read_run(evidence_dir, out$run_id)` retrieves one record.
This local evidence store is optional. Saving an output and saving its evidence
are separate operations; an evidence-write failure does not roll back an
external target. Lake publication additionally records its own transactional
release and quality history.

For a rule's original error, use `validate(data, contract, keep_errors = TRUE)`
and `quality_errors()` locally. The exception may contain private values.

## Distinguish definition versions from data releases

| Identity or time | What it describes |
|---|---|
| Product ID | The named data product consumers use |
| Contract or product version | A definition of structure, rules or behavior |
| `code_version` | Caller-declared identity of code and external dependencies |
| Release ID | One immutable published data state |
| `business_date` | The date represented by the delivery |
| Receipt/publication timestamps | When the delivery arrived and was released |

You can omit explicit definition versions for ordinary product composition.
They become useful when another team needs a stable, reviewed promise. A
correction for the same business date creates a new release; it does not
overwrite the earlier one. A version number is not a substitute for a release
ID when reproducing data.

## Read an exact earlier release

The remaining storage examples require optional DuckDB and use temporary local
files. Two ordinary writes are enough to create history:

```{r release-history, eval=requireNamespace("duckdb", quietly = TRUE)}
root <- tempfile("tidyweave-history-")
lake <- open_lake(root)
first <- write_data(lake, data.frame(id = 1L, amount = 10), "orders")
second <- write_data(lake, data.frame(id = 1L, amount = 12), "orders")
read_release(lake, "orders")
read_release(lake, "orders", release = first$release_id)
releases(lake, "orders") |>
  dplyr::select(asset, release_id, quality)
```

The first read returns 12; the pinned read returns 10. To use the earlier data
inside a new workflow, use the same product grammar:

```{r pinned-source, eval=requireNamespace("duckdb", quietly = TRUE)}
historical_total <- product("historical_total") |>
  add_source(source_release(lake, "orders", first$release_id)) |>
  add_transform(function(data) dplyr::summarise(data, total = sum(amount))) |>
  run()
collect(historical_total)
```

Omitting `release_id` from `source_release()` resolves the current release at
execution and records the chosen immutable identity. Keep the source lake
connection open while using its lazy tables.

## Reuse only when the intended inputs are unchanged

`run()` evaluates a product again by default. For a lake target, opt into
`run(cache = TRUE)` with an explicit product `code_version`; only a matching
current release can be reused. Callbacks can consult external state that their
function body does not reveal, so update `code_version` when that state or its
dependencies change. Dynamic reference checks deliberately reject reuse.

`write_data()` can reuse an identical current delivery automatically. Supplying
older data after a newer delivery publishes a new current release. A result
with status `cached` refers to existing release evidence; it is not a fresh
evaluation of every rule.

## Monitor only deadlines that matter

Contracts do not assume a freshness deadline. Set `max_age_hours`, or monitor
an expected delivery's business date and due time explicitly:

```{r expected-delivery, eval=requireNamespace("duckdb", quietly = TRUE)}
delivery_contract <- contract(columns = c(id = "integer", amount = "numeric"),
  max_age_hours = 48)
check_delivery(lake, "orders", delivery_contract,
  business_date = as.Date("2026-09-30"),
  due_at = as.POSIXct("2026-10-01 09:00:00", tz = "UTC"),
  at = as.POSIXct("2026-10-01 10:00:00", tz = "UTC"))
```

Run monitoring from your existing scheduler. Notification callbacks are
optional; without a transport, no message is sent. A transport should use the
event ID for idempotency, because a crash after delivery and before local
confirmation can cause a duplicate.

For partition corrections, comparisons and read-only analysis, continue with
[everyday workflows](everyday-workflows.html). Published-history retention,
distributed locking and atomic multi-product publication remain separate
operational concerns.

```{r cleanup, include=FALSE}
unlink(evidence_dir, recursive = TRUE)
if (has_duckdb) {
  stopifnot(collect(historical_total)$total == 10)
  close_lake(lake)
  unlink(root, recursive = TRUE)
}
```
