---
title: "Choose and configure quality rules"
description: "Add contracts, tolerances, reference checks and pointblank."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choose and configure quality rules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Use this after [Check a delivery](check-delivery.html). Start with the
requirement you need now. Basic rules need no optional engine; the pointblank
and storage sections state their additional dependencies.

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
library(tidyweave)
```

Quality checks answer a practical question: **can somebody safely use this
result?** Start with the requirement you already know. You do not need a
database, an external quality engine or governance metadata.

```{r first-check}
orders <- data.frame(id = 1:3, amount = c(10, 20, 30))
result <- tw_product("orders", orders) |>
  tw_add_quality(~ amount >= 0) |>
  tw_trial()
tw_collect(result)
tw_quality(result)
```

## Choose the smallest useful tool

| Tool | Question | Does it decide whether publication is allowed? |
|---|---|---|
| `tw_profile_data(data)` | What types, missing counts and ranges occur? | No |
| `tw_contract(columns = ...)` | Which structure, required fields and keys do I promise? | Yes |
| `tw_add_quality(~ amount >= 0)` | Does this business rule hold? | Yes |
| `tw_pointblank_checks()` | Can an existing pointblank plan evaluate these rules? | Yes |

A profile describes data. It does not turn observations into requirements.
Database and Arrow profiles collect aggregate results, not samples; exact
counts can still require scanning the data on the backend.

```{r profile-and-contract}
tw_profile_data(orders)
schema <- tw_contract(
  columns = list(id = integer(), amount = double()),
  key = "id"
)
tw_validate(orders, schema)
```

By default declared fields cannot be missing, and additional fields are
rejected. Use `required = character()` or `allow_extra = TRUE` when that is the
promise you intend. Owners, descriptions, versions and freshness deadlines are
optional. `max_age_hours = NULL` means that age is not monitored; set a deadline
only when the consumer has one.

Factor prototypes describe character labels. The original factor stays intact
in R; a database need not preserve its levels or ordering. `integer64` is a
distinct exact type. For caller-owned DuckDB connections, set
`DBI::dbConnect(duckdb::duckdb(), bigint = "integer64")`; tidyweave configures
its own lake connections this way. A list column needs a target supporting
nested data.

## Missing values and tolerances are explicit

Formulas evaluate one test per row. An ordinary function can do the same:

```{r row-check}
rule <- tw_quality_rule("nonnegative", function(data) data$amount >= 0)
tw_run_quality(rule, data.frame(amount = c(10, -1, NA)))
```

Both `FALSE` and `NA` fail. A scalar function result, such as
`function(data) nrow(data) >= 10`, is one aggregate test. A scalar formula is
repeated over the rows. Longer function results must have exactly one logical
value per row. For custom units, return `tw_quality_counts(failed, total)`.

`max_failure = 0.02` permits two failures out of 100. If a rule breaches its
tolerance, `severity = "error"` blocks publication; `severity = "warning"`
permits it with warning evidence. Empty, skipped or errored checks never count
as passing checks. Contract null checks remain independent of rule tolerances.

## Check relationships without adopting a relational framework

A reference rule answers whether each key occurs in another table. A named
vector maps input columns to reference columns, including composite keys:

```{r reference-check}
customers <- data.frame(id = c("a", "b"), region = c("EU", "US"))
deliveries <- data.frame(
  customer_id = c("a", "b", "missing"),
  region = c("EU", "US", "EU")
)
known_customer <- tw_quality_reference(
  customers,
  by = c(customer_id = "id", region = "region")
)
tw_run_quality(known_customer, deliveries)
```

This finds one unmatched row. Duplicate reference keys do not multiply the
input. Missing keys fail by default; `na_matches = "na"` explicitly permits
missing values to match. The reference can also be a zero-argument function or
source adapter, resolved anew on each run.

Checks between lazy tables on one database use an anti-join and aggregate
counts there. Moving distinct reference keys between backends requires
`copy = TRUE`. Dynamic reference rules cannot use cached lake publication,
because the reference can change independently of the product input.

## Add pointblank only when its capabilities help

The next example runs only when the optional `pointblank` package is installed.
Segments avoid letting a large healthy group hide failures in a small group.

```{r pointblank-check, eval=requireNamespace("pointblank", quietly = TRUE)}
checks <- tw_pointblank_checks("amounts", function(data) {
  pointblank::create_agent(
    data,
    actions = pointblank::action_levels(warn_at = 0.005, stop_at = 0.05)
  ) |>
    pointblank::col_vals_gte(
      "amount", 0, segments = pointblank::vars(entity)
    )
}, policy = "agent")
amounts <- data.frame(
  entity = rep(c("Alpha", "Beta"), each = 100), amount = 10
)
amounts$amount[1:3] <- -1
evidence <- tw_validate(
  amounts,
  tw_contract(columns = c(entity = "character", amount = "numeric"),
    rules = list(checks)),
  keep_agents = TRUE
)
tw_quality(evidence) |>
  dplyr::filter(engine == "pointblank") |>
  dplyr::select(rule, segment, status, n_failed, n_total)
```

Three failures in Alpha cause a warning; five reach the blocking threshold.
`policy = "agent"` uses pointblank's native action levels and rounding.
The default `policy = "rule"` instead uses the rule's `severity` and
`max_failure`. These are alternative policies. Under the agent policy, each
step needs a warning or blocking action level. Inactive or errored steps and
empty plans block under either policy.

## Keep the last good release when a delivery fails

Storage is optional. Add a lake target when readers need durable, historical
releases. This example requires DuckDB and uses only a temporary local folder.

```{r lake-gates, eval=requireNamespace("duckdb", quietly = TRUE)}
root <- tempfile("tidyweave-quality-")
lake <- tw_open_lake(root)
definition <- tw_product("orders") |>
  tw_add_contract(schema) |>
  tw_add_quality(~ amount >= 0) |>
  tw_add_quality(tw_quality_rule("unusually_large", ~ amount < 100,
    severity = "warning")) |>
  tw_set_target(lake)
first <- definition |> tw_add_source(orders) |> tw_run()
unusual <- data.frame(id = 1:3, amount = c(10, 200, 30))
warning_release <- definition |> tw_add_source(unusual) |> tw_run()
bad <- data.frame(id = 1:3, amount = c(10, -1, 30))
blocked <- definition |> tw_add_source(bad) |> tw_run(stop_on_failure = FALSE)
tw_status(blocked)
tw_read_release(lake, "orders")
```

The amount of 200 is still current: its warning permitted publication. The
negative amount blocked the next delivery and did not replace that release.
The original release is also available by `first$release_id`.

```{r reports, eval=requireNamespace("duckdb", quietly = TRUE)}
tw_quality_report(tw_quality(blocked), file.path(root, "blocked.html"))
tw_quality(lake, asset = "orders", release = warning_release$release_id) |>
  dplyr::select(rule, status, n_failed)
```

HTML and JSON reports contain check metadata, without source rows. For a native
pointblank report, call `tw_pointblank_report(evidence, "amounts", path)` while
the result with `keep_agents = TRUE` is still in memory. Agents are not stored
in the registry.

## Put business decisions into tests

`tw_expect_quality()` belongs in your package's testthat tests. It applies the
same fail-closed gate used for publication:

```{r quality-test, eval=FALSE}
testthat::test_that("the example delivery meets its contract", {
  tw_expect_quality(tw_validate(example_delivery, schema))
})
```

For failure diagnosis, historical reads and cache behavior, continue with
[Quality and history](quality-history.html). dbt model tests belong in the dbt
project; see [dbt workflows](dbt-workflows.html).

```{r close-quality-lake, include=FALSE, eval=requireNamespace("duckdb", quietly = TRUE)}
stopifnot(blocked$status == "blocked",
  identical(tw_read_release(lake, "orders")$amount, unusual$amount))
tw_close_lake(lake)
unlink(root, recursive = TRUE)
```


[Back to Learn](learn.html) · [Troubleshooting](troubleshooting.html)
