---
title: "Quality gates: pointblank, dbt and reliable releases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quality gates: pointblank, dbt and reliable releases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE)
library(lakefold)
if (!requireNamespace("pointblank", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

## Three quality responsibilities

| Stage | Tool | Responsibility |
|---|---|---|
| Incoming delivery | pointblank inside a lakefold contract | Plausibility, value ranges, segments and tolerated failures |
| Building persisted SQL models | dbt | Model dependencies, structural tests, SQL rules and materializations |
| Relational analysis in R | dm | Explicit primary/foreign keys and relationships |

lakefold makes the publication decision and records its evidence. A successful
dbt model is initially a mutable relation. A lakefold release references an
immutable candidate snapshot. Scheduling and permissions belong to the
surrounding platform.

## Define a contract with graded pointblank thresholds

This guide requires the optional pointblank package:
`install.packages("pointblank")`. If it is unavailable during documentation
building, the examples are displayed but not executed.

A single global failure ratio can hide problems in small entities. This rule
therefore uses pointblank segments by `entity`. lakefold retains evidence for
each segment and checks all results before publication.

```{r pointblank-contract}
contract <- dl_contract(
  "orders.contract", "1.0.0", "Analytics", "Order amounts", "One order",
  columns = c(id = "integer", entity = "character", amount = "numeric"),
  key = "id", producer = "Data producer", operator = "Platform operations",
  column_metadata = list(amount = list(description = "Order amount", unit = "EUR")),
  rules = list(dl_pointblank("plausibility", function(data) {
    pointblank::create_agent(
      tbl = data,
      actions = pointblank::action_levels(warn_at = 0.005, stop_at = 0.05)
    ) |>
      pointblank::col_vals_gte("amount", 0, segments = pointblank::vars(entity))
  }, policy = "agent"))
)
orders <- data.frame(
  id = seq_len(200), entity = rep(c("Alpha", "Beta"), each = 100),
  amount = rep(10, 200)
)
orders$amount[1:3] <- -1
quality <- dl_validate(orders, contract, keep_agents = TRUE)
dl_quality(quality) |>
  dplyr::filter(engine == "pointblank") |>
  dplyr::select(rule, segment, status, n_failed, n_total, failure_rate)
```

`policy = "agent"` uses action levels as evaluated by pointblank, including its
rounding of proportions to test units. In this example, three failures among
100 units trigger a warning; five reach the blocking threshold. Absolute
thresholds are also supported. The example uses pointblank 0.12.4 argument
names. Newer pointblank versions may also expose `warn`, `error` and `critical`.

`policy = "rule"` remains the compatible default: `max_failure` is the maximum
permitted failure ratio, and `severity` determines whether a breach warns or
blocks. The two policies are alternatives, not combined thresholds. Under the
agent policy, each check needs at least one warning or blocking action level.
Skipped checks, empty plans and evaluation errors block publication regardless
of the chosen tolerance. If multiple blocking action columns are present, any
triggered blocking action prevents publication.

When lakefold handles notifications, configure pointblank action levels without
custom sending functions. The runner records diagnostics before dispatching a
deduplicated event through the supplied transport.

## Validate before Raw and check the full candidate again

```{r ingestion}
root <- tempfile("lakefold-quality-")
config <- dl_config(
  dl_catalog_duckdb(file.path(root, "lake.db")),
  dl_storage_local(file.path(root, "data")),
  landing = file.path(root, "landing"), backend = "duckdb"
)
lake <- dl_connect(config)
first <- dl_ingest_data(lake, orders, contract, "orders", code_version = "v1",
  input_contract = contract, business_date = as.Date("2026-08-31"))
dl_status(first)
orders$amount[4:5] <- -1
blocked <- dl_ingest_data(lake, orders, contract, "orders", code_version = "v1",
  input_contract = contract, business_date = as.Date("2026-08-31"),
  stop_on_failure = FALSE)
dl_status(blocked)
dl_quality(lake, run_id = blocked$run_id) |>
  dplyr::filter(engine == "pointblank") |>
  dplyr::select(stage, rule, status, n_failed)
```

The rejected RDS snapshot remains in landing. This failed attempt writes no Raw
table, and the previous release remains available. `dl_ingest_data()` also works
after `readxl::read_excel()` or an API request made with existing R tools. It
archives the received R object. Use `dl_source()` and
`dl_ingest(..., input_contract = contract)` to archive an original file unchanged.

In an explicit pipeline, `dl_step_precheck()` follows `dl_step_extract()` and
precedes transformations. Its reader must return a data frame. The final
`dl_step_validate()` remains mandatory and checks the complete combined state
when replacing partitions. Input and product contracts may differ.

## Inspect results and export reports

```{r diagnostics}
dl_releases(lake, "orders") |>
  dplyr::select(asset, release_id, quality, business_date)
published_quality <- dl_quality(lake, asset = "orders", release = first$release_id)
dl_expect_quality(published_quality)
html <- file.path(root, "quality.html")
dl_quality_report(dl_quality(blocked), html)
dl_quality_report(published_quality, file.path(root, "quality.json"), format = "json")
dl_pointblank_report(quality, "plausibility", file.path(root, "pointblank.html"))
file.exists(c(html, file.path(root, "pointblank.html")))
```

`dl_quality(lake, asset = "orders")` shows the latest attempt. Supply `release`
to retrieve evidence for a particular published state. Cache-hit diagnostics
resolve the original checks. `dl_status()` and `dl_quality()` also accept dbt
results; process failures remain visible even when individual nodes succeeded.

The compact HTML/JSON report contains check metadata and segment labels, without
full source rows. The native pointblank report needs the in-memory result from
`keep_agents = TRUE`; agents are not stored in the registry. Failed-row extracts
and table samples are disabled. `dl_expect_quality()` applies the runner's gate
decision inside testthat.

## Review a contract draft explicitly

```{r contract-draft}
draft <- dl_contract_from(orders, "orders.next", "Analytics", "Order amounts",
  "One order", key = "id", required = c("id", "entity", "amount"))
draft
reviewed <- dl_contract_confirm(draft)
dl_contract_diff(reviewed, contract) |>
  dplyr::select(field, breaking)
```

The constructor reads column types, using a zero-row prototype for lazy tables.
It does not infer business keys, units or rules from samples. A draft can become
a gate only after `dl_contract_confirm()`. An `NA` in the difference report
indicates a change requiring semantic review, such as a changed R rule or grain.
Confirmation is an explicit review step, not an access-controlled approval system.

## Detect missing deliveries independently of imports

```{r monitoring}
dl_check_delivery(lake, "orders", contract, business_date = as.Date("2026-09-30"),
  due_at = as.POSIXct("2026-10-01 09:00:00", tz = "UTC"),
  at = as.POSIXct("2026-10-01 10:00:00", tz = "UTC"))
dl_freshness(lake) |>
  dplyr::select(asset, published_quality, latest_attempt, delivery_status)
dl_cleanup(lake, older_than_days = 30) # Preview only; no tables are removed
```

Invoke monitoring from the existing scheduler. `notify` can connect an existing
transport, so the package requires no email service. Repeated identical incidents
are suppressed after successful notification; transport failures remain
retryable. The catalog distinguishes publication quality, the latest attempt
and delivery status. The expected business date is compared with the latest
published release.

`dl_cleanup(..., dry_run = FALSE)` removes only expired unpublished Raw/candidate
tables from completed failed attempts. Published releases, reports, landing
files and quality evidence are retained. DuckLake snapshots and physical files
are not cleaned up by this operation.

## Publish dbt results for reproducible reporting

This section needs a configured dbt project and is not executed here. Integration
tests exercise build, test, snapshot publication and the resulting query.

```{r dbt-release, eval=FALSE}
dl_disconnect(lake) # Close the local catalog before the external dbt process.
result <- dl_dbt_build(project)
lake <- dl_connect(config)
release <- dl_dbt_publish(lake, result, "model.shop.customer_revenue",
  revenue_contract, "shop.revenue", code_version = "v1")
dl_tbl(lake, "shop.revenue", release$release_id)
```

The package copies the currently visible relation and validates that exact copy.
An earlier invocation ID alone cannot prove unchanged table contents, so this
operation never approves a cached release solely from the dbt ID. Operators
must exclude competing writers between build and snapshot. One call publishes
one relation, not an atomic bundle of all dbt models.

```{r cleanup, include=FALSE}
dl_disconnect(lake)
unlink(root, recursive = TRUE)
```
