---
title: "Case study: insurance reporting"
description: "Combine three deliveries, checked relationships, SQL models and a correction."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case study: insurance reporting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

An advanced case study. Complete [Get started](get-started.html) and
[Use dbt](dbt-workflows.html) first. The complete runner needs DuckDB,
pointblank, dm and a dbt CLI. Synthetic definitions are included in the appendix;
the downloadable runner includes them so it can execute independently.

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
library(tidyweave)
run_stack <- identical(Sys.getenv("TIDYWEAVE_RUN_INSURANCE_VIGNETTE"), "true")
path <- tempfile("insurance-guide-")
backend <- "ducklake"
executable <- Sys.getenv("TIDYWEAVE_DBT_EXECUTABLE", "dbt")
```

```{r insurance-fixtures, include=FALSE}
insurance_inputs <- function() {
  brokers <- tibble::tibble(
    broker_id = c("B1", "B2", "B3"),
    broker_name = c("Pine Brokers", "Direct Desk", "Harbor Partners"),
    channel = c("broker", "direct", "partner")
  )
  policy_months <- tibble::tibble(
    policy_id = rep(paste0("P", 1:6), 2),
    month = rep(as.Date(c("2026-01-01", "2026-02-01")), each = 6),
    company = rep(
      c("Northstar", "Northstar", "Northstar", "Harbor", "Harbor", "Northstar"),
      2
    ),
    broker_id = rep(c("B1", "B1", "B2", "B3", "B3", "B1"), 2),
    status = c(
      rep("active", 5),
      "pending",
      "active",
      "active",
      "lapsed",
      rep("active", 3)
    ),
    premium_due = c(100, 200, 300, 400, 500, 0, 100, 200, 0, 400, 500, 150)
  )
  payments <- tibble::tibble(
    payment_id = paste0("T", 1:10),
    policy_id = c("P1", "P1", "P2", "P3", "P4", "P1", "P2", "P4", "P5", "P6"),
    month = rep(as.Date(c("2026-01-01", "2026-02-01")), each = 5),
    payment_date = as.Date(c(
      "2026-01-05",
      "2026-01-20",
      "2026-01-08",
      "2026-01-10",
      "2026-01-12",
      "2026-02-05",
      "2026-02-08",
      "2026-02-12",
      "2026-02-15",
      "2026-02-18"
    )),
    cash_amount = c(60, 40, 180, 300, 400, 100, 200, 350, 500, 150)
  )
  list(brokers = brokers, policy_months = policy_months, payments = payments)
}

insurance_corrected_inputs <- function(inputs = insurance_inputs()) {
  inputs$payments <- dplyr::bind_rows(
    inputs$payments,
    tibble::tibble(
      payment_id = "T11",
      policy_id = "P5",
      month = as.Date("2026-01-01"),
      payment_date = as.Date("2026-01-28"),
      cash_amount = 250
    )
  )
  inputs
}

insurance_contracts <- function() {
  policy_columns <- c(
    policy_id = "character",
    month = "Date",
    company = "character",
    broker_id = "character",
    status = "character",
    premium_due = "numeric"
  )
  payment_columns <- c(
    payment_id = "character",
    policy_id = "character",
    month = "Date",
    payment_date = "Date",
    cash_amount = "numeric"
  )
  dimension_columns <- c(
    company = "character",
    channel = "character",
    month = "Date"
  )
  policy_summary <- c(
    dimension_columns,
    active_policies = "integer",
    premium_due = "numeric"
  )
  payment_summary <- c(
    dimension_columns,
    payment_count = "integer",
    cash_collected = "numeric"
  )
  mart_columns <- c(
    policy_summary,
    payment_count = "integer",
    cash_collected = "numeric"
  )
  policy_contract <- contract(
    "insurance.policy_months.schema", columns = policy_columns,
    key = c("policy_id", "month"),
    grain = "One monthly snapshot row per policy, including inactive policies.")
  payment_contract <- contract(
    "insurance.payments.schema", columns = payment_columns,
    key = "payment_id",
    grain = "One cash transaction, assigned to its receipt month.")
  mart_structure <- contract(
    "insurance.monthly_performance.structure", columns = mart_columns,
    grain = "One company-channel-month; all monetary values are synthetic EUR.")
  list(
    brokers = contract(
      "insurance.brokers.schema",
      columns = c(
        broker_id = "character",
        broker_name = "character",
        channel = "character"
      ),
      key = "broker_id",
      grain = "One row per broker; attributes are static in this example."
    ),
    policy_months = policy_contract,
    payments = payment_contract,
    enriched_policy_months = contract_update(policy_contract,
      id = "insurance.enriched_policy_months.schema",
      columns = c(broker_name = "character", channel = "character"),
      required = c(policy_contract$required, "broker_name", "channel")),
    enriched_payments = contract_update(payment_contract,
      id = "insurance.enriched_payments.schema",
      columns = c(company = "character", broker_id = "character",
        broker_name = "character", channel = "character", policy_status = "character"),
      required = c(payment_contract$required, "company", "broker_id", "broker_name",
        "channel", "policy_status")),
    core_policy_monthly = contract(
      "insurance.core_policy_monthly.schema",
      columns = policy_summary,
      grain = "One company-channel-month policy aggregate."
    ),
    core_cash_monthly = contract(
      "insurance.core_cash_monthly.schema",
      columns = payment_summary,
      grain = "One company-channel-month cash aggregate."
    ),
    mart_schema = mart_structure,
    mart = contract_update(mart_structure,
      id = "insurance.monthly_performance.schema",
      key = c("company", "channel", "month"),
      rules = list(
        non_negative_counts = ~ active_policies >= 0 & payment_count >= 0,
        non_negative_amounts = ~ premium_due >= 0 & cash_collected >= 0
      )
    )
  )
}

insurance_metrics <- function() {
  metric_set("insurance.monthly_performance",
    active_policies = sum(active_policies, na.rm = TRUE),
    premium_due = sum(premium_due, na.rm = TRUE),
    cash_collected = sum(cash_collected, na.rm = TRUE),
    cash_to_due = sum(cash_collected, na.rm = TRUE) / sum(premium_due, na.rm = TRUE),
    dimensions = c("company", "channel"),
    time_column = "month",
    time_behavior = c(active_policies = "stock", premium_due = "flow",
      cash_collected = "flow", cash_to_due = "flow"),
    units = c(active_policies = "policies", premium_due = "EUR",
      cash_collected = "EUR", cash_to_due = "ratio"),
    approved = TRUE, code_version = "insurance-example-v3")
}
```

This worked example turns three insurance deliveries into two reusable data
products, a monthly SQL mart and an issued report. The same grammar describes
ordinary tables, Pointblank checks, dm lookups and lake publication. The optional
engines implement individual steps; their configuration does not surround every
business operation.

All organizations, policies and payments are synthetic. Amounts use EUR only.
This is a data engineering example, not a regulatory reporting or accounting
policy. The complete example is executable:

```{r complete-example, eval=FALSE, purl=FALSE}
library(tidyweave)
source(system.file("examples", "relational-insurance.R", package = "tidyweave"))
demo <- run_relational_insurance(executable = "/path/to/dbt")
demo$initial$summary
```

Sourcing the script only defines its synthetic data, contracts, metrics and
runner. Calling the runner starts the external workflow. The actual product
calls appear below; they are not hidden in integration helper functions.

## 1. Begin with the business question and the grain

For January and February 2026, we want active policy counts, premium due and
cash received, both overall and by company and channel. These need two facts:
a policy snapshot and a cash transaction. A policy may have several payments
or no payment in a month.

`month` labels a reporting period using its first day. `2026-01-01` means the
January month-end snapshot, not the policy's state on January 1. For payments it
means receipt month, with `payment_date` retaining the actual transaction date.
`measure(..., at = as.Date("2026-01-01"))` therefore selects January.

| Table | One row means | Unique key | Lookup |
|---|---|---|---|
| Brokers | One broker with fixed attributes | `broker_id` | None |
| Policy months | One policy at one reporting month | `policy_id`, `month` | One broker |
| Payments | One cash receipt | `payment_id` | One policy-month, then one broker |
| Enriched policies | The same policy-month with broker attributes | `policy_id`, `month` | Already checked |
| Enriched payments | The same cash receipt with descriptive attributes | `payment_id` | Already checked |
| Monthly mart | One company, channel and month | `company`, `channel`, `month` | Aggregated facts |

![Two checked many-to-one enrichment paths preserve policy-month and payment grains before separate SQL aggregation.](figures/relational-insurance.svg)

A policy with EUR 100 due and payments of EUR 40 and EUR 60 would appear twice
in a direct policy-to-payments join. Summing its repeated premium would claim
EUR 200 due. Valid relationships alone do not make that aggregation correct.
We enrich the facts independently and combine them after aggregating to the
same monthly grain.

```{r runner-start, include=FALSE, eval=isTRUE(knitr::opts_knit$get("tangle"))}
run_relational_insurance <- function(path = tempfile("tidyweave-insurance-"),
  backend = "ducklake", executable = "dbt") {
  packages <- c("duckdb", "pointblank", "dm", "yaml", "processx")
  missing <- packages[!vapply(packages, requireNamespace, logical(1), quietly = TRUE)]
  if (length(missing)) stop("Install example dependencies: ", paste(missing, collapse = ", "))
  if (!file.exists(executable) && !nzchar(Sys.which(executable)))
    stop("Install dbt-duckdb and supply its dbt executable.")
  backend <- match.arg(backend, c("ducklake", "duckdb"))
  if (file.exists(path) && (!dir.exists(path) ||
    length(list.files(path, all.files = TRUE, no.. = TRUE))))
    stop("Choose a new or empty example directory.")
```

```{r input-data}
inputs <- insurance_inputs()
vapply(inputs, nrow, integer(1))
```

```{r input-preview, purl=FALSE}
knitr::kable(inputs$policy_months)
knitr::kable(inputs$payments)
```

The inputs contain three brokers, twelve policy-month observations and ten
payments. They are full snapshots for these two months. A corrected payment
snapshot retains existing transactions and adds the correction. This example
does not implement CDC, append-only delivery, premium allocation or receivables
ageing. Those require explicit additional business rules.

## 2. Add only the infrastructure this workflow uses

The first product examples in the [introductory lessons](get-started.html)
need no infrastructure. This full walkthrough uses DuckDB or DuckLake, optional
Pointblank and dm, plus a separately installed dbt-duckdb CLI.

```{r install-packages, eval=FALSE, purl=FALSE}
install.packages(c("duckdb", "pointblank", "dm", "yaml", "processx"))
```

Follow the [dbt guide](dbt-workflows.html) for compatible R/Python engines.
`backend = "duckdb"` avoids the DuckLake extension; the product grammar and
business calculations are unchanged. Both engines need exclusive local write
coordination. A configuration opens no connection, so a dbt process can acquire
the catalog after R publication closes its owned connection.

```{r configure, eval=isTRUE(knitr::opts_knit$get("tangle")) || run_stack}
config <- lake_config(path = file.path(path, "lake"), backend = backend,
  layers = c("raw", "staging", "core", "marts"))
execution <- execution_config(quality = "pointblank", relationships = "dm",
  to = config)
```

| Location | Role | Acceptance |
|---|---|---|
| Landing | Original deliveries or received-value snapshots | Receipt only |
| RAW | Input checked in its received shape | Immutable accepted releases |
| STAGING | Two enriched products | Independently checked releases |
| CORE | Reusable dbt aggregates | Mutable SQL build relations |
| MARTS | Reporting model and its approved snapshot | The published snapshot is immutable |

A layer describes a role. Publication describes acceptance. A staging product
can be a governed reusable output, and a dbt table in marts can still fail a test.

## 3. State the promises that matter

```{r contract-definitions}
contracts <- insurance_contracts()
```

```{r payment-contract, purl=FALSE}
contract("insurance.payments.schema",
  columns = c(payment_id = "character", policy_id = "character", month = "Date",
    payment_date = "Date", cash_amount = "numeric"),
  key = "payment_id", grain = "One cash transaction in its receipt month.")
```

A contract describes the expected columns, types, non-null fields and business
key. `contract_update()` extends existing columns and preserves their expectations
under a new explicit identity. The enriched products add descriptive columns;
their keys and row meaning are unchanged. Added columns are optional unless
included in `required`, so this example explicitly requires the new attributes.
The mart uses a new contract because its grain changes. Contracts do not invent keys,
repair orphan references or silently coerce a mistaken identifier.

The appendix contains the complete ordinary R contract definitions. Owner,
freshness and governance fields remain optional. Without a contract, the framework
can infer structure, but it cannot infer what one business row should mean.
Changing a contract's `grain` requires an explicit key decision, including
`key = character()` when no uniqueness promise is intended. Existing business
rules must also be reviewed explicitly when grain, column names or types change.
Contracts are not silently propagated through arbitrary dplyr transformations.
Named formulas in `rules` use the same predicate grammar as `add_quality()`;
there is no separate business-rule language to learn.

## 4. Accept the three deliveries

```{r accept-inputs, eval=isTRUE(knitr::opts_knit$get("tangle")) || run_stack}
deliveries <- list(
  brokers = product("insurance.brokers", inputs$brokers, execution = execution) |>
    add_contract(contracts$brokers) |>
    add_quality(~ channel %in% c("broker", "direct", "partner")),
  policy_months = product("insurance.policy_months", inputs$policy_months,
    execution = execution) |>
    add_contract(contracts$policy_months) |>
    add_quality(~ status %in% c("active", "pending", "lapsed") & premium_due >= 0),
  payments = product("insurance.payments", inputs$payments, execution = execution) |>
    add_contract(contracts$payments) |>
    add_quality(~ cash_amount > 0)
)
raw <- lapply(deliveries, ingest)
quality(raw$payments)
```

The execution configuration is retained by each delivery definition and selects
Pointblank once for all these predicates. Changing this execution default does
not require rewriting the declared contract version; the actual check engine
is still recorded in execution evidence.
Omitting that choice uses the native engine; it does not require learning another grammar for a simple
logical check. Complex Pointblank agents remain available through
`pointblank_checks()`. Missing logical results fail the check.

Input checks run before RAW writes. A rejected delivery stays in landing, while
the previous accepted release remains available. Each successful result can be
used directly as a source: its exact release and lineage travel with it. These
three acceptances are separate commits, not a distributed transaction.

This ingestion path collects data into R for the gate and snapshot. It needs
memory for the delivery. A lazy source alone does not turn it into streaming
ingestion.

## 5. Enrich policies with a checked lookup

```{r policy-product, eval=isTRUE(knitr::opts_knit$get("tangle")) || run_stack}
policy_product <- product("insurance.policy_months.enriched", raw$policy_months,
  source_name = "policy_months", execution = execution) |>
  add_lookup(raw$brokers, by = dplyr::join_by(broker_id)) |>
  add_contract(contracts$enriched_policy_months)
```

The primary table remains one policy-month per row. `add_lookup()` requires
unique, non-missing parent keys and, by default, a match for every child key.
It rejects ambiguous parent rows and orphan children before enrichment. It does
not expand policies into their payment transactions.

The execution configuration selects real dm relationship checks, including
lookups inside dependencies. The native engine provides the same guarantees.
An explicit `engine` on an individual step overrides the execution default. Equality joins are supported; interval and rolling
joins need an explicit R transformation and an appropriate business contract.
Both engines respect compatible lazy sources; neither silently copies unrelated
databases. Use explicit collection or connection-specific code for that boundary.

## 6. Enrich payments at their own grain

```{r payment-product, eval=isTRUE(knitr::opts_knit$get("tangle")) || run_stack}
policy_attributes <- product("insurance.payment_policy_attributes", raw$policy_months) |>
  dplyr::select(policy_id, month, company, broker_id, policy_status = status)

payment_product <- product("insurance.payments.enriched", raw$payments,
  source_name = "payments", execution = execution) |>
  add_lookup(policy_attributes, by = dplyr::join_by(policy_id, month)) |>
  add_lookup(raw$brokers, by = dplyr::join_by(broker_id)) |>
  add_contract(contracts$enriched_payments)
```

The first lookup uses **policy ID and month together**. A policy ID alone would
match two snapshots and could copy attributes from the wrong date. The second
lookup adds the broker attributes. Premium due is deliberately absent from
`policy_attributes`, because it is not additive across payment rows.

These are real dplyr methods on the existing product object. `select()`,
`mutate()`, `filter()`, `group_by()` and `summarise()` capture their ordinary
expressions now and execute dplyr against the actual table later. They do not
read data during definition or translate arbitrary R code into SQL themselves.
Supported lazy behavior is inherited from dplyr/dbplyr and the chosen backend.
An ordinary `add_transform(function(data) ...)` remains the escape hatch.

## 7. Publish independently reusable staging products

```{r publish-staging, eval=isTRUE(knitr::opts_knit$get("tangle")) || run_stack}
published <- list(
  policies = policy_product |> publish(layer = "staging"),
  payments = payment_product |> publish(layer = "staging")
)
stopifnot(nrow(collect(published$policies)) == 12L,
  nrow(collect(published$payments)) == 10L)
```

Each product has its own contract and immutable release. A portfolio report can
reuse the policies, and a cash analysis can reuse the payments without rebuilding
the lookups. The output gates verify that enrichment preserved the promised
columns and keys. Lake publication materializes inputs for archival and checking;
it is not an end-to-end lazy warehouse executor.

## 8. Let dbt aggregate each fact before combining them

The bundled project contains ordinary SQL models, explicit enforced dbt column
contracts exported below, not-null tests, composite-grain tests and reconciliations. Copy it to
a fresh working directory once:

```{r copy-sql-project, eval=isTRUE(knitr::opts_knit$get("tangle")) || run_stack}
template <- system.file("examples", "relational-insurance", "dbt", package = "tidyweave")
project_path <- file.path(path, "analytics")
dir.create(project_path, recursive = TRUE, showWarnings = FALSE)
stopifnot(all(file.copy(list.files(template, full.names = TRUE, all.files = TRUE,
  no.. = TRUE), project_path, recursive = TRUE)))
```

The SQL column contracts come from the same R structural definitions. This
small, explicit schema bridge avoids maintaining the column list twice. Business
formulas and the final composite key stay in the R publication contract; dbt's
SQL tests enforce the corresponding SQL grain and reconciliation promises.

```{r export-sql-contracts, eval=isTRUE(knitr::opts_knit$get("tangle")) || run_stack}
properties <- list(version = 2L, models = c(
  dbt_contract(contracts$core_policy_monthly, "core_policy_monthly")$models,
  dbt_contract(contracts$core_cash_monthly, "core_cash_monthly")$models,
  dbt_contract(contracts$mart_schema, "monthly_performance")$models))
yaml::write_yaml(properties, file.path(project_path, "models", "contracts.yml"))
```

```{r build-and-publish, eval=isTRUE(knitr::opts_knit$get("tangle")) || run_stack}
project <- dbt_project(project_path, lake = config, sources = published,
  executable = executable)
built <- project |> run(echo = FALSE)
approved <- built |> publish("monthly_performance", contract = contracts$mart,
  asset = "insurance.monthly_performance", layer = "marts")
```

The managed project derives its connection profile and binds logical dbt source
names to the selected staging releases. There is no manual release-ID lookup,
profile generator or source-YAML helper. The models use ordinary dbt references:

```sql
-- Policy stock and premium aggregate
select company, channel, month,
  cast(sum(case when status = 'active' then 1 else 0 end) as integer) as active_policies,
  cast(sum(premium_due) as double) as premium_due
from {{ source('inputs', 'policies') }}
group by company, channel, month
```

```sql
-- Independent cash aggregate
select company, channel, month,
  cast(count(*) as integer) as payment_count,
  cast(sum(cash_amount) as double) as cash_collected
from {{ source('inputs', 'payments') }}
group by company, channel, month
```

The mart left-joins these aggregates on company, channel and month. It fills
missing cash groups with zero, retaining policies with no payment. The inactive
Northstar/direct February group remains a row with zero active policies, due and
cash. The SQL tests reconcile source totals with the mart, check unique reporting
grains and require a payment's actual receipt date to belong to its labelled month.

A successful dbt build is separate from consumer approval. dbt can write models
before a later test fails; it does not roll back the whole graph. `publish()`
checks and snapshots the chosen successful model. A failed build cannot replace
the prior approved release. Coordinate writers between building and snapshotting.

```{r inspect-mart, eval=run_stack, purl=FALSE}
collect(approved)
dbt_status(built)
```

## 9. Measure stock, flows and a ratio of sums

```{r metric-definitions}
metrics <- insurance_metrics()
january <- as.Date("2026-01-01")
february <- as.Date("2026-02-01")
```

```{r measure-approved, eval=isTRUE(knitr::opts_knit$get("tangle")) || run_stack}
measures <- measure(approved, metrics = metrics,
  at = c(january, february), period = "each")
summary <- collect(measures)
totals <- measure(approved, metrics = metrics[c("cash_collected", "cash_to_due")],
  at = c(january, february), period = "aggregate")
by_channel <- measure(approved, metrics = metrics["cash_collected"],
  at = january, by = c("company", "channel"))
collect(by_channel)
```

`metric_set()` defines the shared product, dimensions and time column once.
Only formulas, units and stock-versus-flow meaning differ. Its result is an
ordinary named list of metrics, so subsetting remains ordinary R syntax.

`period = "each"` computes every named metric separately for each selected date.
`collect()` returns one ordinary tidy table with `.metric`, `.period`, `.unit`
and `value`, plus any grouping columns. `.period` is a list column: each entry
retains the selected Date value or vector of dates. There is no need to assemble
individual values into a reporting table. Keep the original measurement set for
report issuance; the collected table is for analysis and presentation.

`period = "aggregate"` requests one calculation over all selected dates. It does
not override a stock metric's single-date requirement. The published result pins
every measurement automatically. A later publication
cannot change these calculations. An active-policy **stock** needs exactly one
reporting date: five active policies in each month does not mean ten policies at
one time. Premium due and cash received are **flows** in this example and can be
summed over the selected months.

The cash-to-due ratio is `sum(cash_collected, na.rm = TRUE) / sum(premium_due, na.rm = TRUE)`. Initially this
is `2280 / 2850 = 0.8` across both months. Averaging monthly or channel percentages
would give different weights to different premium bases. A zero denominator is
undefined; the non-finite measurement is rejected rather than converted to zero.
`na_policy = "reject"` still blocks missing metric inputs before calculation;
`na.rm = TRUE` makes the SQL aggregate behavior explicit after that gate.
An `approved = TRUE` metric records a declared business decision, not a separate
authorization system. Code versions identify the reviewed calculation.

## 10. Issue a report, keeping values and their explanation

```{r issue-report, eval=isTRUE(knitr::opts_knit$get("tangle")) || run_stack}
report <- report_release(measures, "insurance.report.initial", to = config,
  code_version = "insurance-report-v1",
  params = list(currency = "EUR", months = c("2026-01", "2026-02")))
initial <- list(build = built, release = approved, data = collect(approved),
  metrics = metrics, measures = measures, totals = totals, by_channel = by_channel,
  report = report, summary = summary)
```

Report issuance remains an explicit registry write. Passing the configuration
lets the framework open and close its writable connection, including on failure.
A caller-supplied connection remains the caller's responsibility. Reading a saved report retrieves its values and manifests without
rerunning SQL or replacing its inputs with the latest release. January initially
reports five active policies, EUR 1,500 due and EUR 980 cash; February reports
five active policies, EUR 1,350 due and EUR 1,300 cash.

## 11. Exercise the boundaries that should reject data

```{r reject-negative, eval=isTRUE(knitr::opts_knit$get("tangle")) || run_stack}
bad_payments <- inputs$payments
bad_payments$cash_amount[1] <- -60
rejected <- product("insurance.payments", bad_payments) |>
  add_contract(contracts$payments) |>
  add_quality(~ cash_amount > 0) |>
  ingest(execution = execution, stop_on_failure = FALSE)
stopifnot(status(rejected)$outcome == "blocked")
```

The example does not model refunds, so negative cash fails before RAW. Supporting
refunds later requires an explicit business change. The rejected receipt is still
retained. Now try a structurally valid but unknown broker identifier:

```{r reject-orphan, eval=isTRUE(knitr::opts_knit$get("tangle")) || run_stack}
orphan_policies <- inputs$policy_months
orphan_policies$broker_id[1] <- "B404"
orphan_run <- policy_product |>
  publish(sources = list(policy_months = orphan_policies),
    layer = "staging", stop_on_failure = FALSE)
stopifnot(status(orphan_run)$outcome == "failed")
lake <- connect_lake(config, read_only = TRUE)
current_policy_release <- releases(lake, published$policies$asset)$release_id[[1]]
close_lake(lake)
stopifnot(identical(current_policy_release, published$policies$release_id))
```

The lookup error explains the unmatched key; no candidate becomes current.
A duplicated or missing parent key similarly fails before it can multiply rows.
A known policy in an unknown month is an orphan of the **composite** key.
For specialist diagnostics, construct a dm object explicitly and inspect
`dm_examine_constraints()`; the simple product path already enforces its checks.

```{r reject-metric-scopes, eval=isTRUE(knitr::opts_knit$get("tangle")) || run_stack}
two_month_stock <- tryCatch(measure(approved, metrics = metrics["active_policies"],
  at = c(january, february), period = "aggregate"), error = identity)
undefined_ratio <- tryCatch(measure(approved, metrics$cash_to_due,
  at = february, filters = list(company = "Northstar", channel = "direct")),
  error = identity)
stopifnot(inherits(two_month_stock, "error"), inherits(undefined_ratio, "error"))
```

## 12. Accept a correction without rewriting the issued report

A late correction adds a January receipt of EUR 250 for policy P5. This is a new
full snapshot with eleven transactions. Brokers and policy snapshots are unchanged.

```{r accept-correction, eval=isTRUE(knitr::opts_knit$get("tangle")) || run_stack}
monthly <- workflow(
  received = function(payments) {
    deliveries$payments |>
      replace_sources(insurance.payments = payments) |> ingest()
  },
  enriched = function(received) {
    payment_product |> publish(sources = list(payments = received), layer = "staging")
  },
  built = function(enriched) {
    project |> run(sources = list(payments = enriched), echo = FALSE)
  },
  released = function(built) {
    built |> publish("monthly_performance", contract = contracts$mart,
      asset = "insurance.monthly_performance", layer = "marts")
  },
  measures = function(released) {
    measure(released, metrics = metrics, at = c(january, february), period = "each")
  },
  inputs = list(payments = inputs$payments), code_version = "insurance-workflow-v1")
corrected_inputs <- insurance_corrected_inputs(inputs)
correction <- run(monthly, inputs = list(payments = corrected_inputs$payments))
status(correction)
corrected_raw <- raw
corrected_raw$payments <- correction$results$received
corrected_products <- published
corrected_products$payments <- correction$results$enriched
rebuilt <- correction$results$built
corrected_release <- correction$results$released
```

The delivery definition keeps its RAW checks. The named `payments` source
reuses the enriched product's lookups and contract, while the dbt project rebinds
only its logical `payments` input. Brokers and policy releases remain pinned.
Unknown or ambiguous names fail early; the original definitions remain reusable.

The workflow records the RAW, enrichment, dbt, publication and measurement
dependencies once. A correction replaces the named delivery and runs its consumers.
Unchanged successful branches can be reused. Failed steps block their consumers;
`stop_on_failure = FALSE` returns all step status for diagnosis and an explicit
retry with `previous`. These writes are not one transaction. Broker and policy
releases stay pinned in this payment-correction workflow. Change `code_version`
when changing the captured definitions. Report issuance stays explicit.

```{r measure-correction, eval=isTRUE(knitr::opts_knit$get("tangle")) || run_stack}
corrected_measures <- correction$results$measures
corrected_totals <- measure(corrected_release,
  metrics = metrics[c("cash_collected", "cash_to_due")],
  at = c(january, february), period = "aggregate")
```

```{r issue-correction, eval=isTRUE(knitr::opts_knit$get("tangle")) || run_stack}
corrected_report <- report_release(corrected_measures, "insurance.report.corrected", to = config,
  code_version = "insurance-report-v1",
  params = list(currency = "EUR", months = c("2026-01", "2026-02")))
preserved <- list(report = report_read(config, report$id, values_only = TRUE),
  data = collect(approved))
corrected_summary <- collect(corrected_measures)
corrected <- list(raw = corrected_raw, products = corrected_products, build = rebuilt,
  release = corrected_release, data = collect(corrected_release), metrics = metrics,
  measures = corrected_measures, totals = corrected_totals,
  report = corrected_report, summary = corrected_summary)
stopifnot(isTRUE(all.equal(preserved$report, summary)),
  identical(dplyr::filter(summary, .metric == "cash_collected")$value, c(980, 1300)),
  identical(dplyr::filter(corrected_summary, .metric == "cash_collected")$value, c(1230, 1300)),
  dplyr::filter(collect(corrected_totals), .metric == "cash_to_due")$value == 2530 / 2850,
  sum(preserved$data$cash_collected[preserved$data$month == january]) == 980)
```

The new report explains the correction. The original still contains its original
EUR 980 January cash and references its original mart release. Consumers can use
`collect(approved)` in Shiny or Quarto, or `product("export", approved)` with a
Parquet, DBI or pins target. BI service deployment and refresh remain external.

```{r runner-end, include=FALSE, eval=isTRUE(knitr::opts_knit$get("tangle"))}
list(context = list(path = path, config = config, inputs = inputs, contracts = contracts),
    raw = raw, definitions = list(policies = policy_product, payments = payment_product),
    products = published, project = project, initial = initial, corrected = corrected,
    failures = list(pointblank = rejected,
      dm = list(result = orphan_run, current_release = current_policy_release,
        previous_release = published$policies$release_id)), preserved = preserved)
}
```

## Why this architecture

The user describes one primary table, optional transformations, checked lookups,
quality and a destination. Internally those inputs become specifications,
dependencies and execution steps. Ordinary dplyr methods describe preparation;
quality and lookup engines are interchangeable; source and target S3 adapters
isolate infrastructure. There is one product object and one execution lifecycle.

Use explicit SQL when the SQL graph is the useful representation. Use a custom
Pointblank agent for specialist checks, an ordinary function for an algorithm,
or direct DBI for backend-specific capabilities. The framework supplies
composition, gates and provenance, without recreating those tools.

The remaining explicit choices are intentional: business keys and time meaning,
metric formulas, the dbt models and contracts, where data belongs, and when a
report is issued. Simpler syntax cannot infer those decisions safely.

## Appendix: synthetic definitions

<details><summary>Inspect the synthetic data, contracts and metric definitions</summary>

```{r show-insurance-fixtures, ref.label="insurance-fixtures", eval=FALSE, purl=FALSE}
```

</details>



```{r cleanup, include=FALSE, purl=FALSE}
if (run_stack) unlink(path, recursive = TRUE)
```


[Back to Learn](learn.html) · [Troubleshooting](troubleshooting.html)
