---
title: "4. Save a report"
description: "Keep reported values and their evidence while later data changes."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. Save a report}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
  eval = requireNamespace("duckdb", quietly = TRUE))
```

**Goal:** save reported values and read them later without recalculating.
**Needs:** tidyweave and DuckDB >= 1.5.5. This example creates its own data and
storage; no objects from lesson 3 are needed.

## Publish the input

```{r input}
library(tidyweave)
folder <- tempfile("orders-report-")
orders <- product("orders", data.frame(id = 1:3, amount = c(25, 75, 50))) |>
  add_quality(~ amount >= 0) |>
  set_target(folder)
first <- orders |> publish()
```

A report must identify the saved input it used. Keep `first` as the input to
calculation rather than passing only its collected table.

## Define related calculations together

```{r metrics}
metrics <- metric_set("orders",
  total = sum(amount, na.rm = TRUE),
  count = dplyr::n(),
  units = c(total = "EUR", count = "orders"),
  code_version = "orders-v1", approved = TRUE)
values <- first |> measure(metrics = metrics)
collect(values)
```

A **metric** names a calculation. The set shares the input table `"orders"`.
The results are **150 EUR** and **3 orders**.

`approved = TRUE` records your decision that the definitions are suitable for
this report; it does not carry out a review or grant access. `code_version`
identifies the reviewed calculation code. Change it when relevant code,
external values or dependencies change. For exploration, omit those two
arguments and use `measure()` or ordinary dplyr; issuing a report requires them.

## Save the report once

```{r save}
report <- values |> report_release("orders-report-v1", code_version = "report-v1")
saved <- report_read(folder, "orders-report-v1", values_only = TRUE)
saved
```

The destination comes from the published input. A **report** stores the values
and the evidence explaining them. Its ID names this issued report; the report
code version identifies your reporting logic.

This produces saved data and evidence, not a PDF, Word document or slide deck.
Use the returned tables with your reporting tools. Reading a report never
recalculates its formulas.

## Check what happens after a correction

```{r correction}
corrected <- data.frame(id = 1:3, amount = c(30, 75, 50))
second <- orders |> publish(data = corrected)
current <- second |> measure(metrics = metrics)
collect(current)
original_report <- report_read(folder, "orders-report-v1", values_only = TRUE)
original_report
```

The current total is **155**; the issued report remains **150**. To issue revised
values, use a new report ID. Reusing an ID with changed content is rejected.
Retrying the same content is safe.

```{r verify-and-cleanup, include=FALSE}
stopifnot(identical(saved, original_report),
  collect(values)$value[collect(values)$.metric == "total"] == 150,
  collect(current)$value[collect(current)$.metric == "total"] == 155)
unlink(folder, recursive = TRUE)
```

## Continue

[Previous: Publish and correct data](publish-data.html) · [Next: Work through a reporting month](getting-started.html) · [All five lessons](get-started.html)
