---
title: "Publish to a shared DuckLake"
description: "Configure PostgreSQL once and coordinate publications from independent R sessions."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Publish to a shared DuckLake}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**Use this when:** several people publish to the same lake. Complete
[the cancellation report](cancellation-report.html) first. Business definitions
remain the same; only the destination changes.

You need a PostgreSQL database dedicated to this DuckLake, shared S3-compatible
storage, DuckDB with DuckLake/PostgreSQL/httpfs extensions, and RPostgres.
Every writer must use this version of tidyweave. The database and bucket must
already exist and each person needs authorized access. This guide does not
provision infrastructure or implement access policies.

## Configure the destination once

Supply the PostgreSQL libpq connection string through the environment variable
`TIDYWEAVE_PG_CONNECTION`, for example through your deployment's secret manager.
Keyword/value strings and `service=your_service` are supported; URI strings are
not supported by the writer coordinator. S3 credentials come from
`AWS_ACCESS_KEY_ID`, `AWS_SECRET_ACCESS_KEY` and optionally `AWS_SESSION_TOKEN`.
Do not put credentials in product definitions, scripts or Git.

```{r config, eval=FALSE, purl=FALSE}
library(tidyweave)
team <- tw_lake_config(
  catalog = tw_registry_postgres("TIDYWEAVE_PG_CONNECTION", lock_timeout = 30),
  storage = tw_storage_s3(
    bucket = "your-team-bucket", prefix = "insurance",
    endpoint = "https://your-s3-endpoint", region = "eu-central-1"
  ),
  landing = "/your-shared-evidence/insurance",
  backend = "ducklake"
)
```

All clients use the same database, data location and layer names. The landing
folder must be durable and accessible where receipt evidence is consumed.
S3 table storage alone does not make local landing files shared. Configure
credentials separately for each person; configuration objects contain their
environment-variable names, not credential values.

## Publish and correct

Use the `portfolio` model product and `reporting_product()` definition from the
cancellation guide. No business definition needs a PostgreSQL-specific rewrite.

```{r publish, eval=FALSE, purl=FALSE}
first_model <- tw_publish(portfolio, to = team)
first_report_data <- tw_publish(reporting_product(first_model), to = team)

second_model <- tw_publish(portfolio, to = team, previous = first_model,
  sources = list(policies = corrected_policies))
second_report_data <- tw_publish(reporting_product(second_model), to = team,
  previous = first_report_data)
```

`previous` is your correction's starting point. If someone has already published
a newer release, the call fails before replacing it. Read the newer state,
reconcile the complete delivery, and publish with that state as `previous`.
Without `previous`, publication is a deliberate new replacement using the
normal ordered-writer policy. Always use `previous` for user-authored corrections.

A model's tables commit together, including their relationship evidence and
model manifest. The derived report is a separate publication. Its source is the
exact model you passed, even if another model is published meanwhile.
Saving report values remains a separate reviewed step with `tw_report_release()`.

## What simultaneous writing means

Independent R sessions may submit work concurrently. PostgreSQL advisory locks
serialize package write operations per catalog database, including setup,
definition registration, publication, report issuance and recovery. A session
releases its lock when the operation returns or fails; an idle open lake does
not reserve the writer. If the wait exceeds `lock_timeout`, the operation fails
with an actionable busy message. There is no unbounded retry loop.

This first implementation prioritizes ordered, consistent writes over parallel
write throughput. Model validation occurs before acquiring the publication lock;
publication rechecks an explicit `previous` while holding it. A model result
contains all tables in memory during validation, so size the session accordingly.

Direct SQL, dbt materialization and unrelated clients do not participate in this
protocol. Use dbt-owned schemas for intermediate work and `tw_dbt_publish()` for
package-managed publication. Do not mutate package-managed release tables.
Concurrent external changes are outside these guarantees. Catalog migrations and
old clients must be coordinated operationally; metadata schema 4 rejects older
package writers after upgrade.

## Verification

The repository includes `scripts/check-postgres.R` and a dedicated CI job using
PostgreSQL 16, DuckLake and independent R processes. It checks concurrent clients,
stale corrections, idempotent report issuance and model quality gates. Local
DuckDB tests cover multi-table rollback and immutable historical models.
These do not replace testing your S3 endpoint, authentication and network-failure
behavior in the deployment environment.

[Back to Learn](learn.html) · [Local DuckLake](create-ducklake.html)

For scheduling with targets, put the whole model publication in an ordinary
`targets::tar_target()` call. `tw_as_targets()` currently expands table-product
dependencies only; it refuses model expansion rather than splitting an atomic
model into independently published tables.
