---
title: "Three deliveries, one report, and a correction"
description: "Correct payments and reference tables by name while retaining issued reports."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Three deliveries, one report, and a correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**Use this when:** your report combines several deliveries and a later correction
may affect either payments or a reference table. Complete [Get started](get-started.html)
first. The example is self-contained; saving and reading history needs optional
DuckDB >= 1.5.5. Run the visible code from top to bottom.


```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
```

You receive payments, contracts and brokers for one reporting period. You need
totals by company, must fix a bad payment, and later accept a correction without
changing the report you already issued. All data here are synthetic. Only saving
and reopening results needs optional DuckDB >= 1.5.5.

The deliveries are **complete snapshots**, not lists of changed rows.
The report sums payments for this one period. A payment has one unique ID, belongs
to one contract and must be non-negative. These are explicit business rules for
this example, not rules the package guesses for you.

## Name the three deliveries

```{r deliveries}
library(tidyweave)
payment_data <- data.frame(
  payment_id = c("T1", "T2", "T3"),
  policy_id = c("P1", "P2", "P3"), amount = c(100, 200, -50))
contract_data <- data.frame(
  policy_id = c("P1", "P2", "P3"),
  broker_id = c("B1", "B1", "B2"), company = c("North", "North", "South"))
broker_data <- data.frame(
  broker_id = c("B1", "B2"), channel = c("Broker", "Direct"))

payments <- tw_product("payments", payment_data) |>
  tw_add_lookup(contract_data, by = "policy_id", name = "contracts") |>
  tw_add_lookup(broker_data, by = "broker_id", name = "brokers") |>
  tw_add_contract(tw_contract(
    columns = c(payment_id = "character", policy_id = "character",
      amount = "numeric", broker_id = "character", company = "character",
      channel = "character"),
    key = "payment_id")) |>
  tw_add_quality(~ amount >= 0, name = "nonnegative")
payments
```

`tw_product()` stores the instructions. `tw_add_lookup()` adds reference columns and
checks that each payment matches exactly one reference row. That protects your
totals from duplicated or lost payments. The names `payments`, `contracts` and
`brokers` are the same names you will use for corrections. Bare table variables
also supply a default name; explicit names keep the script independent of your
variable names. `tw_explain(payments)` lists them without reading any data.

Use dplyr `mutate()`, `filter()` and `summarise()` for preparation as needed.
Checked enrichment uses `tw_add_lookup()`. General joins on ordinary tables belong
inside `tw_add_transform()`; the package does not guess their relationship rules.

## Find and fix the data error

```{r diagnose}
preview <- tw_trial(payments)
preview
tw_quality_report(preview)
tw_quality_rows(preview)
stopifnot(identical(tw_quality_rows(preview)$payment_id, "T3"))
```

The trial returns a blocked result instead of interrupting your session.
`tw_quality_rows()` shows T3 with amount -50. With several failed rules, it lists
their names and asks which one you want to inspect. No row data is copied into
portable check reports. `tw_collect(preview)` refuses to return failed data as
a successful result. Correct the delivery at its source; checks never repair it.

```{r repair}
fixed_payments <- payment_data
fixed_payments$amount[fixed_payments$payment_id == "T3"] <- 50
preview <- tw_trial(payments, sources = list(payments = fixed_payments))
tw_collect(preview)
stopifnot(sum(tw_collect(preview)$amount) == 350)
```

Only the named delivery changes for this call. The definition, other deliveries,
joins and checks are reused. For the main delivery alone, `data = fixed_payments`
is an equivalent shortcut. `tw_trial()` disables configured package writers and
keeps the result in memory. `tw_publish()` and configured `tw_run()` calls still stop
on failure by default.

## Calculate the report you want

```{r explore}
metrics <- tw_metric_set("payments", total = sum(amount, na.rm = TRUE),
  dimensions = "company", units = "EUR")
preview_values <- tw_measure(preview, metrics = metrics, by = "company")
preview_values
stopifnot(identical(tw_collect(preview_values)$value, c(300, 50)))
```

`dimensions` lists the columns you allow people to group or filter by.
`by = "company"` chooses this report's grouping. `by = character()` explicitly
chooses an overall total. If you omit `by`, the package reminds you of the
available grouping instead of silently suggesting the dimension was used.
`tw_collect(preview_values)` gives an ordinary table for plotting or exploration.

## Save the reviewed report

The next steps need DuckDB. A report here means saved numbers and their
calculation history, not a rendered PDF. Use collected values in your usual
reporting tool to create a document.

```{r storage-availability, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(eval = requireNamespace("duckdb", quietly = TRUE))
```

```{r publication}
root <- tempfile("tidyweave-everyday-")
payments <- payments |> tw_set_target(root)
first <- tw_publish(payments, sources = list(payments = fixed_payments))

# Only after business review: approval is your explicit declaration.
metrics <- tw_metric_set("payments", total = sum(amount, na.rm = TRUE),
  dimensions = "company", units = "EUR",
  approved = TRUE, code_version = "payment-totals-v1")
values <- tw_measure(first, metrics = metrics, by = "company")
tw_report_release(values, "august-v1", code_version = "report-v1")
```

Set the storage folder once. `tw_publish()` checks and saves the complete table;
it does not put anything on the internet. Earlier saved results remain available.
No connection management or internal version ID is needed in this workflow.

Approval declares that you reviewed the metric. The code version identifies
the calculation you reviewed; change it when that calculation changes.
The report ID, such as `august-v1`, identifies this issued set of numbers.
Keep `values` for report issuance: `tw_collect(values)` is for ordinary analysis
and does not retain the complete report evidence.

## Accept a corrected payment delivery

```{r correction}
corrected_payments <- fixed_payments
corrected_payments$amount[corrected_payments$payment_id == "T3"] <- 80
second <- tw_publish(payments, sources = list(payments = corrected_payments))
difference <- tw_compare(first, second)
difference
difference$changed

corrected_values <- tw_measure(second, metrics = metrics, by = "company")
corrected_values
tw_report_release(corrected_values, "august-v2", code_version = "report-v1")
```

The contract already states the payment key, so comparison needs no extra
argument. T3 changes from 50 to 80. North stays at 300 and South becomes 80.
The calculation is unchanged, so its code version stays the same. This is a new
report, so it receives a new report ID.

These calls do not rewrite the product definition's original inputs. Each call
receives a complete named delivery. To retain corrected defaults for later
calls, explicitly update the definition with
`payments <- tw_replace_sources(payments, payments = corrected_payments)`.

## Correct a reference delivery using the same names

Suppose contract P3 was assigned to the wrong company. Use the same operation,
now supplying the corrected contracts alongside the current payments.

```{r reference-correction}
corrected_contracts <- contract_data
corrected_contracts$company[corrected_contracts$policy_id == "P3"] <- "North"
third <- tw_publish(payments, sources = list(
  payments = corrected_payments, contracts = corrected_contracts))
reference_values <- tw_measure(third, metrics = metrics, by = "company")
reference_values
stopifnot(identical(tw_collect(reference_values)$value, 380))
```

The broker delivery and all checks are unchanged. Unmatched or duplicated
reference keys still block the result. Replace one or several deliveries in the
same `sources` list, without rebuilding the preparation steps.

## Reopen exactly what was reported

```{r reopen}
original_report <- tw_report_read(root, "august-v1", values_only = TRUE)
corrected_report <- tw_report_read(root, "august-v2", values_only = TRUE)
original_report
corrected_report
stopifnot(identical(as.numeric(original_report$value), c(300, 50)),
  identical(as.numeric(corrected_report$value), c(300, 80)),
  sum(tw_collect(first)$amount) == 350,
  sum(tw_collect(second)$amount) == 380,
  difference$counts[["changed"]] == 1)
```

Reading reports does not recalculate them. The later reference correction
changes neither issued report.

## Add capabilities when the task calls for them

Use [complete-month replacement](getting-started.html) when each delivery covers
one whole month instead of the whole table. Use the
[layered data guide](layered-data-stack.html) for separate receipt evidence and
dbt SQL steps, or the [insurance guide](relational-insurance.html) for a larger
process. Optional engines stay behind `tw_execution_config(quality = "pointblank",
relationships = "dm")`; the ordinary lookup and predicate syntax stays the same.

Lake writes require one coordinated writer. Publication across several products
and dbt execution are not a single transaction. These operating choices are
separate from the everyday table workflow above.

```{r cleanup, include=FALSE}
unlink(root, recursive = TRUE)
```


[Back to Learn](learn.html) · [Repeat dependent steps](everyday-workflows.html) · [Troubleshooting](troubleshooting.html)
