---
title: "Workflow design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflow design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(lakefold)
```

## Define before executing

Like a tidymodels workflow, a lakefold specification describes work before it
is executed. The analogy is conceptual: there is no training/resampling system,
and `dl_plan()` is structural inspection, not SQL execution or a data dry run.

```{r specification}
config <- dl_config(backend = "duckdb")
source <- dl_source("orders.file", "orders.csv", utils::read.csv)
contract <- dl_contract(
  "orders.contract", "1.0.0", "Analytics", "Order amounts", "One order",
  c(order_id = "integer", amount = "numeric"), key = "order_id"
)
pipeline <- dl_pipeline("orders.import", config, code_version = "v1") |>
  dl_step_land(source) |>
  dl_step_extract() |>
  dl_step_transform(\(data) dplyr::mutate(data, amount = amount * 100),
    id = "amount_in_cents") |>
  dl_step_validate(contract) |>
  dl_step_publish("orders")
pipeline
dl_plan(pipeline)
```

These chunks do not open connections, read `orders.csv` or create folders.
Transforms are applied in order after extraction. They leave the landed original
and raw table intact, and return data frames or lazy tables. Use `collect()`
explicitly when an R-only algorithm needs in-memory data.

## Execute with an object-first API

| Specification | Unified entry point | Specific operation |
|---|---|---|
| Pipeline | `dl_execute(pipeline)` | `dl_run(pipeline)` |
| R product | `dl_execute(product, lake)` | `dl_build(lake, product)` |
| Metric | `dl_execute(metric, lake)` | `dl_measure(lake, metric)` |
| dbt project | `dl_execute(project)` | `dl_dbt_build(project)` |

A pipeline carries its connection-free config. Products and metrics need an
explicit connection or config. Execution closes connections it opens; supplied
connections remain caller-owned. dbt opens its own connections in a separate
process, so do not pass `lake` when executing a dbt project.

The result types reflect their operation: run result, metric tibble or dbt
result. A universal diagnostic accessor is not implemented yet.

## Inspect incomplete definitions

```{r incomplete}
incomplete <- dl_pipeline("orders.import", config, code_version = "v1") |>
  dl_step_land(source)
dl_plan(incomplete)
attr(dl_plan(incomplete), "complete")
```

`complete` only means the declared steps and layers form a valid sequence.
Availability of files, credentials, SQL translations and data quality are
checked at execution. Errors while adding steps explain invalid ordering early.

## Version definitions deliberately

Registered `(kind, id, version)` definitions are immutable. Reconstruct a changed
source or contract with a new version. Because a pipeline embeds these objects,
increase its version too. Change `code_version` when callbacks, captured values
or dependencies change; function-body hashes cannot capture all environments.

Specifications can be saved as RDS but are not a portable language-independent
format. Prefer version-controlled R definitions and explicit dependency locks.
Update/remove-step helpers and an external adapter extension contract are future
work. See the design review for the remaining gap to mature tidy frameworks.
